---
title: "R Notebook"
output: html_notebook
---

```{r warning=FALSE}
library(tidyverse)
library(janitor)
library(RColorBrewer)
library(ggeffects)
library(glmmTMB)
```

```{r read in data}
obs <- read.csv("data/KBAYKATMKEFJ_2006-2023_Summer_MBM_Observations.csv")
transects <- read.csv("data/KBAYKATMKEFJ_2006-2023_Summer_MBM_Transects.csv")
```

```{r make a single df}
#make year and site columns from the trip ID
obs <- obs %>%
  mutate(
    site = substr(TripID, 1, 4),
    year = substr(TripID, 8, 11)
  ) %>%
  filter()

species <- unique(obs$Detect_SpeciesCode)
sites <- unique(obs$site)
years <- unique(obs$year)

#make transects by site
transects <- transects  %>%
  mutate(
    site = substr(TripID, 1, 4),
    year = substr(TripID, 8, 11)
  )

#identify pelagic transects
pelagic_transects <- transects %>%
  filter(TransType == "PELAGIC") %>%
  pull(TXNumber) %>%
  unique()

#now remove pelagic transects from bird obs
obs <- obs %>% filter(!TXNumber %in% pelagic_transects)
```

link obs with transect effort

```{r}
#sum species total for each survey (year and site)
obs_summary <-obs %>%
  group_by(year, site, Detect_SpeciesCode) %>%
  summarise(count = sum(Count)) %>%
  rename(species = Detect_SpeciesCode)

#find all unique combinations of year and site then repeat the bird list for each of them
sampling_events <- obs %>%
  distinct(year, site)

species_df <- species %>% 
  data.frame() %>%
  rename(species = ".")

species_list <- list()

for(i in 1:nrow(sampling_events)) {
  event <- sampling_events[i,]
  species_df$year <- event$year
  species_df$site <- event$site
  
  species_list[[i]] <- species_df
}

#convert list to dataframe
sampling_events <- bind_rows(species_list)

#this adds in zeros for the species not observed on each survey but doesnt make up surveys that didn't happen (as opposed to expand_grid)
obs_summary <- left_join(sampling_events, obs_summary) %>%
  mutate(count = replace_na(count, 0))

#this adds up all the area and distance sampled in each year
transects_summary <- transects %>% 
  group_by(year, site) %>%
  summarise(length = sum(Length_km),
            area = sum(Area_Sampled_km2))

#add the effort to the counts and calculate density
bird_densities <- left_join(obs_summary, transects_summary, by = c("site", "year")) %>%
  mutate(density = count/area) %>%
  relocate(species, .after = site)
```

find the most common birds to look at

```{r}
birds <- obs %>%
  filter(!Detect_SpeciesCode %in% c("UNCO")) %>%
  group_by(Detect_SpeciesCode) %>%
  summarise(count = sum(Count)) %>%
  arrange(desc(count)) %>%
  head(20) %>% #change this number for more or less species
  pull(Detect_SpeciesCode)
```

```{r}
bird_densities %>%
  filter(species %in% birds) %>%
  #mutate(density = scale(density)) %>%
  ggplot() +
  geom_path(aes(year, density, colour = site, group = site)) +
  geom_point(aes(year, density, colour = site, group = site)) +
  facet_wrap(~species, scales = "free_y") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Bird Density",
       subtitle = "20 most abundant species")
```

```{r}
bird_densities %>%
  filter(species %in% birds) %>%
  group_by(species) %>%
  mutate(density = scale(density)) %>%
  ungroup() %>%
  ggplot() +
  geom_path(aes(year, density, colour = species, group = species)) +
  geom_point(aes(year, density, colour = species, group = species)) +
  facet_wrap(~site, scales = "free_y", nrow = 3) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "scaled bird abundances")
```

```{r}
bird_densities %>%
  filter(species == "BLOY") %>%
  #mutate(density = scale(density)) %>%
  ggplot() +
  geom_path(aes(year, density, colour = site, group = site)) +
  facet_wrap(~species, scales = "free_y") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
mussel_counts <- read.csv("data/KATMKEFJWPWS_2008-2025_Mytilus_20mm_Count.csv")
mussel_size <- read.csv("data/KATMKEFJWPWS_2008-2025_Mytilus_20mm_Size.csv")
seastar_counts <- read.csv("data/KATMKEFJWPWS_2006-2025_Sea_Star_Count.csv")
```

```{r}
mussel_counts %>%
  filter(!BlockName %in% c("EPWS", "NPWS")) %>%
  mutate(Count = as.numeric(Count)) %>%
  group_by(BlockName, SiteName, YearSample) %>%
  summarise(avg = mean(Count)) %>%
  ggplot() +
  geom_point(aes(YearSample, avg, group = SiteName, colour = SiteName)) +
  facet_wrap(~BlockName, nrow = 3) +
  labs(title = "Mussel Counts")
```

```{r}
mussel_size %>%
  filter(!BlockName %in% c("EPWS", "NPWS")) %>%
  mutate(Size = as.numeric(Size)) %>%
  group_by(BlockName, SiteName, YearSample) %>%
  summarise(avg = mean(Size)) %>%
  ggplot() +
  geom_point(aes(YearSample, avg, group = SiteName, colour = SiteName)) +
  facet_wrap(~BlockName, nrow = 3) +
  labs(title = "Mussel Size")
```

```{r}
mussel_counts %>%
  filter(!BlockName %in% c("EPWS", "NPWS")) %>%
  mutate(Count = as.numeric(Count)) %>%
  group_by(BlockName, SiteName, YearSample) %>%
  summarise(avg = mean(Count)) %>%
  ggplot() +
  geom_boxplot(aes(YearSample, avg, group = YearSample)) +
  facet_wrap(~BlockName, nrow = 3, scales = "free_y") +
  labs(title = "Mussel Counts",
       y = "Count")
```

```{r}
mussel_size %>%
  filter(!BlockName %in% c("EPWS", "NPWS")) %>%
  mutate(Size = as.numeric(Size)) %>%
  group_by(BlockName, SiteName, YearSample) %>%
  summarise(avg = mean(Size)) %>%
  ggplot() +
  geom_boxplot(aes(YearSample, avg, group = YearSample)) +
  facet_wrap(~BlockName, nrow = 3, scales = "free_y") +
  labs(title = "Mussel Size",
       y = "Size (mm)")
```

noww check out the seastars

```{r}
seastar_counts %>%
  mutate(site = substr(SiteID, 1, 3)) %>%
  group_by(Year, site, Species) %>%
  summarise(avg_density = mean(Density_individual200SqM)) %>%
  group_by(Species) %>%
  mutate(avg_density = as.numeric(scale(avg_density))) %>%
  ungroup() %>%
  ggplot() +
  geom_path(aes(Year, avg_density, color = Species)) +
  facet_wrap(~site, nrow = 3, scales = "free_y") +
  labs(title = "Seastar density (scaled)")
```

```{r}
seastar_counts %>%
  mutate(site = substr(SiteID, 1, 3)) %>%
  group_by(Year, site, Species) %>%
  summarise(avg_density = mean(Density_individual200SqM)) %>%
  ggplot() +
  geom_smooth(aes(Year, avg_density, group = site, color = site)) +
  facet_wrap(~Species, nrow = 3, scales = "free_y") +
  labs(title = "Seastar Density")
```

```{r}
bird_densities %>%
  group_by(year, site) %>%
  summarise(avg_density = mean(density)) %>%
  ggplot() +
  geom_path(aes(year, avg_density, color = site, group = site)) +
  labs(title = "all birds combined",
       subtitle = "is this stupid?")
```

combine everything into a single data frame for plotting

```{r}
seastar_wide <- seastar_counts %>%
  mutate(site = substr(SiteID, 1, 3),
         site = case_when(site == "AKP" ~ "KATM",
                          site == "PWS" ~ "WPWS",
                          site == "KEP" ~ "KEFJ")) %>%
  group_by(Year, site, Species) %>%
  summarise(avg_density = mean(Density_individual200SqM)) %>%
  pivot_wider(names_from = Species, values_from = avg_density) %>%
  rename(year = Year)

mussels_wide <- mussel_counts %>%
  rename(site = BlockName,
         year = YearSample) %>%
  filter(!site %in% c("EPWS", "NPWS")) %>%
  mutate(Count = as.numeric(Count)) %>%
  group_by(site, year) %>%
  summarise(avg_count = mean(Count, na.rm = T)) # check that this is ok and isnt getting rid of zeros

birds_wide <- bird_densities %>%
  select(!c(length, area, count)) %>%
  pivot_wider(names_from = species, values_from = density) %>%
  mutate(year = as.numeric(year))

all_taxa <- left_join(birds_wide, seastar_wide, by = c("year", "site")) %>%
  left_join(., mussels_wide, by = c("year", "site")) %>%
  rename(mussel_count = avg_count) %>%
  mutate(
    across(
      c(seq(3, ncol(.))),   
      ~ scale(.)[,1]
    )
  ) %>%
  rename_with(janitor::make_clean_names, .cols = 128:135)
```

```{r}
all_taxa %>%
  filter(site != "KBAY") %>%
  select(year, site, mussel_count, BLOY, HADU, pycnopodia_helianthoides) %>%
  pivot_longer(cols = !c(year, site), names_to = "species", values_to = "Abundance") %>%
  ggplot() +
  geom_path(aes(year, Abundance, color = species)) +
  geom_point(aes(year, Abundance, color = species)) +
  geom_hline(aes( yintercept = 0)) +
  facet_wrap(~site, nrow = 3)
```


```{r}
bloy_nest <- read.csv("data/KBAYKATMKEFJWPWS_2006-2025_blackOystercatcher_Transect_Summary.csv")

bloy_nest %>%
  group_by(Year, BlockName) %>%
  summarise(avg_density = mean(as.numeric(NestDensity))) %>%
  ggplot() +
  geom_path(aes(Year, avg_density, color = BlockName)) +
  labs(title = "BLOY Nest Density")
```

```{r}
isotopes <- read.csv("data/Isotope_data_all_regions_2014-2025.csv", fileEncoding='latin1')
isotopes <- isotopes %>% select(!c(X, X.1, X.2))

```

```{r}
isotopes %>%
  mutate(Year = as.factor(Year)) %>%
  ggplot() +
  geom_jitter(aes(d15N...., d13C...., colour = Year), alpha = 0.4)
  facet_wrap(~Year)
```

```{r}
isotopes %>%
  mutate(Year = as.factor(Year)) %>%
  ggplot() +
  geom_jitter(aes(d15N...., d13C...., colour = Species), alpha = 0.7) +
  facet_wrap(~Year)
```

```{r}
isotopes %>%
  filter(Species == "trossulus") %>%
  mutate(Year = as.factor(Year)) %>%
  ggplot() +
  geom_jitter(aes(d15N...., d13C...., colour = Year), alpha = 0.9, size = 3) +
    scale_x_continuous(limits = c(5, 10))
```

```{r}
trossulus <-isotopes %>%
  filter(Species == "trossulus") %>%
  filter(!is.na(d15N....),
         !is.na(d13C....)) %>%
#mutate(Year = as.factor(Year)) %>%
 rename_with(janitor::make_clean_names)
 
hulls <- trossulus %>%
  group_by(year) %>%
  slice(chull(d15n, d13c)) %>%
  ungroup()


ggplot(trossulus) +
  geom_point(aes(x = d15n, y = d13c),
              alpha = 0.1, size = 3) +
  geom_polygon(data = hulls,
               aes(x = d15n, y = d13c, group = year),
               alpha = 0.2, colour = "black") +
  #scale_x_continuous(limits = c(5, 10)) +
  facet_wrap(~year)
```

```{r}
trossulus %>%
  ggplot() +
  geom_jitter(aes(d15n, d13c, colour = region), alpha = 0.5, size = 3) +
  facet_wrap(~year) +
  labs(title = "mussel isotopes")
```

```{r}
isotopes %>%
  filter(Sample.ID == "POM") %>%
  ggplot() +
  geom_jitter(aes(d15N...., d13C...., colour = Region), alpha = 0.5, size = 3) +
  facet_wrap(~Year) +
  labs(title = "POM")
```

Limpet time!

```{r}
limpet_count <- read.csv("data/KATMKEFJWPWS_2010-2025_Limpet_Count.csv")
limpet_size <- read.csv("data/KATMKEFJWPWS_2006-2025_Limpet_Size.csv")
```

```{r}
limpet_count %>%
  mutate(site = substr(SiteID, 1, 3),
         site = case_when(site == "AKP" ~ "KATM",
                          site == "PWS" ~ "WPWS",
                          site == "KEP" ~ "KEFJ"),
         Density = as.numeric(Density)) %>%
  group_by(Year, site) %>%
  summarise(Density = mean(Density, na.rm = T)) %>%
  mutate(Density = scale(Density)[,1]) %>%
  ggplot() +
  geom_path(aes(Year, Density, color = site)) +
  geom_point(aes(Year, Density, color = site)) +
  labs(title = "limpet density")
```

The limpet abundances between KATM and KEFJ seem to trend oppositely

```{r}
limpet_size %>%
  mutate(site = substr(SiteID, 1, 3),
         site = case_when(site == "AKP" ~ "KATM",
                          site == "PWS" ~ "WPWS",
                          site == "KEP" ~ "KEFJ"),
         Limpet_Size_mm = as.numeric(Limpet_Size_mm)) %>%
  group_by(Year, site) %>%
  summarise(size_avg = mean(Limpet_Size_mm, na.rm = T)) %>%
  #mutate(size_avg = scale(size_avg)[,1]) %>%
  ggplot() +
  geom_path(aes(Year, size_avg, color = site)) +
  geom_point(aes(Year, size_avg, color = site)) +
  labs(title = "limpet size")
```

but the size fluctuations seem to trend similarly between sites

```{r}
limpet_size %>%
  mutate(site = substr(SiteID, 1, 3),
         site = case_when(site == "AKP" ~ "KATM",
                          site == "PWS" ~ "WPWS",
                          site == "KEP" ~ "KEFJ"),
         Limpet_Size_mm = as.numeric(Limpet_Size_mm)) %>%
  group_by(Year, site, SiteName) %>%
  summarise(size_avg = mean(Limpet_Size_mm, na.rm = T)) %>%
  #mutate(size_avg = scale(size_avg)[,1]) %>%
  ggplot() +
  geom_path(aes(Year, size_avg, color = SiteName)) +
  geom_point(aes(Year, size_avg, color = SiteName)) +
  facet_wrap(~site, ncol = 1) +
  labs(title = "limpet size by site") +
  theme(legend.position = "none")
```

snails:

```{r}
snails <- read.csv("data/KATMKEFJWPWS_2006-2025_Nucella_Katharina_Lirabuccinum_Count.csv")
```

```{r}
snails %>%
  mutate(site = substr(SiteID, 1, 3),
         site = case_when(site == "AKP" ~ "KATM",
                          site == "PWS" ~ "WPWS",
                          site == "KEP" ~ "KEFJ")) %>%
  group_by(Year, Elevation_Position, site, SpeciesName) %>%
  summarise(count = mean(Count)) %>%
  ggplot() +
  geom_path(aes(Year, count, color = SpeciesName)) +
  geom_point(aes(Year, count, color = SpeciesName)) +
  facet_grid(site~Elevation_Position) +
  labs(title = "snail abundance")
```

check out the temperature data

```{r message=FALSE}
temperature <- list.files("data/temperature",
                       pattern = "\\.csv$",
                       full.names = TRUE) %>%
  map_dfr(read_csv)

#make month and year columns
temperature$year <- year(temperature$date)
temperature$month <- month(temperature$date)
```

```{r}
temperature %>%
  filter(exposure == "water",
         tide_height == 0.5) %>%
  ggplot() +
  geom_path(aes(date, temperature, color = site)) 
```

```{r}
temperature %>%
  filter(exposure == "water",
         month == 7,
         tide_height == 0.5) %>%
  ggplot() +
  geom_boxplot(aes(year, temperature, group = year, color = block)) +
  facet_wrap(~block, ncol =1)
```

```{r}
temperature %>%
  filter(exposure == "water",
         month == 5,
         tide_height == 0.5) %>%
  group_by(year, block) %>%
  summarise(avg_temp = mean(temperature)) %>%
  ggplot() +
  geom_path(aes(year, avg_temp, color = block)) +
  geom_point(aes(year, avg_temp, color = block)) +
  labs(title = "May Temp")
```

```{r}
temperature %>%
  filter(exposure == "water",
         month == 6,
         tide_height == 0.5) %>%
  group_by(year, block) %>%
  summarise(avg_temp = mean(temperature)) %>%
  ggplot() +
  geom_path(aes(year, avg_temp, color = block)) +
  geom_point(aes(year, avg_temp, color = block)) +
  labs(title = "June Temp")
```

```{r}
temperature %>%
  filter(exposure == "water",
         month == 7,
         tide_height == 0.5) %>%
  group_by(year, block) %>%
  summarise(avg_temp = mean(temperature)) %>%
  ggplot() +
  geom_path(aes(year, avg_temp, color = block)) +
  geom_point(aes(year, avg_temp, color = block)) +
  labs(title = "July Temp")
```

```{r}

temperature %>%
  filter(exposure == "water",
         month %in% c(4,5,6,7,8,9),
         tide_height == 0.5) %>%
  mutate(month_name = factor(month.name[month],
                             levels = month.name[4:9])) %>%
  group_by(year, block, month_name) %>%
  summarise(avg_temp = mean(temperature), .groups = "drop") %>%
  ggplot() +
  geom_path(aes(year, avg_temp, color = block)) +
  geom_point(aes(year, avg_temp, color = block)) +
  facet_wrap(~month_name)
```

```{r}
temperature %>%
  filter(exposure == "water",
         month %in% c(10,11,12,1,2,3),
         tide_height == 0.5) %>%
  mutate(month_name = factor(month.name[month],
                             levels = c("October","November","December",
                                        "January","February","March"))) %>%
  group_by(year, block, month_name) %>%
  summarise(avg_temp = mean(temperature), .groups = "drop") %>%
  ggplot() +
  geom_path(aes(year, avg_temp, color = block)) +
  geom_point(aes(year, avg_temp, color = block)) +
  facet_wrap(~month_name)
```

```{r}
winter_df <- temperature %>%
  filter(exposure == "water",
         month %in% c(10,11,12,1,2,3),
         tide_height == 0.5) %>%
  mutate(month_name = factor(month.name[month],
                             levels = c("October","November","December",
                                        "January","February","March")))

# Step 2: Compute climatology (long-term mean for each month)
climatology <- winter_df %>%
  group_by(month_name, block) %>%
  summarise(clim_mean = mean(temperature), .groups="drop")

# Step 3: Compute avg_temp by group and join climatology
winter_anoms <- winter_df %>%
  group_by(year, block, month_name) %>%
  summarise(avg_temp = mean(temperature), .groups="drop") %>%
  left_join(climatology, by="month_name") %>%
  mutate(anomaly = avg_temp - clim_mean)

# Step 4: Plot anomalies
ggplot(winter_anoms) +
  geom_col(aes(year, anomaly, fill = block.y), position = position_dodge()) +
  facet_wrap(~month_name) +
  labs(y = "Temperature anomaly (°C)")
```

```{r}
obs %>%
  filter(Detect_SpeciesCode == "SEOT") %>%
  pull(Comment) %>%
  write.csv("C:/Users/bhiggins/OneDrive - DOI/Desktop/otters.csv")

otters <- obs %>%
  filter(Detect_SpeciesCode == "SEOT") %>%
  mutate(
    x_lower = str_to_lower(Comment),

    # Define what counts as a "pup" indicator, including your new variants
    has_pup = str_detect(
      x_lower,
      paste0(
        "(",
        "pup",           # 'pup' substring (covers 'pup' and 'pups')
        "|\\bpups\\b",   # explicit plural
        "|\\bp\\b",      # bare 'p' as a whole word (e.g., '1=p', '1 = p')
        "|large\\s*p\\b",# 'large p' (with optional spaces)
        "|large\\s*up\\b",# 'large up' (with optional spaces)
        "|largep",       # 'largep' without space
        "|largeup",      # 'largeup' without space
        ")"
      )
    ),

    any_num = str_detect(x_lower, "\\d+"),

    pups = case_when(
      has_pup & any_num ~ as.numeric(str_extract(x_lower, "\\d+")), # grab leading number
      has_pup & !any_num ~ 1,                                        # 'pup' without number => 1
      TRUE ~ NA_real_
    )
  ) %>%
  select(-x_lower, -has_pup, -any_num) %>%
  mutate(pups = replace_na(pups, 0),
         adults = Count - pups)

#check
which((otters$pups + otters$adults) != otters$Count) #good to go!

```

```{r}
otter_summary <- otters %>%
  group_by(year, site) %>%
  summarise(adults = sum(adults),
            pups = sum(pups),
            ratio = pups/adults) %>%
  mutate(year = as.numeric(year))
```

```{r}
otter_summary %>%
  ggplot() +
  geom_path(aes(year, ratio, color = site)) +
  geom_point(aes(year, ratio, color = site)) +
  labs(title = "SEOT pups/adult ratio")
```

```{r}
bird_densities %>%
  mutate(year = as.numeric(year)) %>%
  filter(species == "SEOT") %>%
  ggplot() +
  geom_path(aes(year, density, color = site)) +
  geom_point(aes(year, density, color = site)) +
  labs(title = "SEOT adult counts")
```

```{r}
temp_df <- temperature %>%
  filter(exposure == "water") %>%
  group_by(year, block) %>%
  summarise(avg_temp = mean(temperature), .groups = "drop")

left_join(otter_summary, temp_df, by = c("year", "site" = "block")) %>%
  filter(site != "KBAY") %>%
  ungroup() %>%
  mutate(
    ratio2 = as.numeric(scale(ratio)),
    temp   = as.numeric(scale(avg_temp))) %>%
  ggplot() +
  geom_path(aes(year, ratio2, color = site)) +
  geom_point(aes(year, ratio2, color = site)) +
  geom_path(aes(year, temp, color = "temp")) +
  facet_wrap(~site, ncol = 1) +
  labs(title = "SEOT pups/adult ratio")
  
```

```{r}
all_taxa %>%
  filter(site != "KBAY") %>%
  select(year, site, mussel_count, BLOY, SEOT, pycnopodia_helianthoides) %>%
  pivot_longer(cols = !c(year, site), names_to = "species", values_to = "Abundance") %>%
  ggplot() +
  geom_path(aes(year, Abundance, color = species)) +
  geom_point(aes(year, Abundance, color = species)) +
  geom_hline(aes( yintercept = 0)) +
  facet_wrap(~site, nrow = 3)
```

```{r}
birds_most_detected <- obs %>%
  filter(!Detect_SpeciesCode %in% c("UNCO")) %>%
  group_by(Detect_SpeciesCode) %>%
  summarise(count = sum(Count),
            detections = n()) %>%
  arrange(desc(detections)) %>%
  head(20) %>% #change this number for more or less species
  pull(Detect_SpeciesCode)

#find the birds we dont want
trash <- obs %>%
  filter(!Detect_SpeciesCode %in% birds_most_detected) %>%
  pull(Detect_SpeciesCode) %>%
  unique()

#add GAK1 temp
#https://research.cfos.uaf.edu/gak1/
gak1 <- read_table(
  "https://research.cfos.uaf.edu/gak1/data/GAK1_Compilation_1970-2025/GAK1_Compilation_1970-2025.dat"
)

gak1$`Date_Time,` <- ymd_hms(gak1$`Date_Time,`)
gak1$year <- year(gak1$`Date_Time,`)

gak_Summary <- gak1 %>%
  mutate(temp = str_replace(`Temperature_[C],`, ",", "")) %>%
  mutate(temp = as.numeric(temp),
         depth = as.numeric(`Pressure_[dbar],`)) %>%
  filter(depth < 10) %>%
  group_by(year) %>%
  summarise(temp = mean(as.numeric(temp), na.rm = T)) %>%
  mutate(temp_scaled = as.numeric(scale(temp)))
  


library(ggcorrplot)
cor_mat <- all_taxa %>% 
   filter(site == "KEFJ") %>%
  select(!site) %>%
  select(-any_of(trash)) %>%
  select(!LEAU) %>%
  select(!UNJA) %>%
  left_join(., gak_Summary, by = "year") %>%
  slice(2:16) %>%
  cor()

ggcorrplot(cor_mat, type = "lower")


hc <- hclust(dist(cor_mat))

order_vec <- hc$order

ggcorrplot(cor_mat[order_vec, order_vec], type = "lower") +
  labs(title ="KEFJ")

```

```{r}
cor_katm <- all_taxa %>% 
   filter(site == "KATM") %>%
  filter(year != 2007) %>%
  select(!site) %>%
  select(-any_of(trash)) %>%
  select(!LEAU) %>%
  select(!UNJA) %>%
  select(!`mediaster_aequalis`) %>%
  left_join(., gak_Summary, by = "year") %>%
  left_join(., otter_summary[otter_summary$site == "KATM",], by = "year") %>%
  select(!site) %>%
  cor(use = "complete.obs")

hc_katm<- hclust(dist(cor_katm))

order_vec_katm <- hc_katm$order

ggcorrplot(cor_katm[order_vec_katm, order_vec_katm], type = "lower") +
  labs(title ="KATM")

```

```{r}
all_taxa %>%
  filter(site != "KBAY") %>%
  select(year, site, RFCO, SEOT) %>%
    left_join(., gak_Summary, by = "year") %>%
  left_join(., otter_summary %>%
              mutate( adults = as.numeric(scale(adults)),
                      pups = as.numeric(scale(pups))), by = c("site", "year")) %>%
  select(!temp) %>%
  pivot_longer(cols = !c(year, site), names_to = "species", values_to = "Abundance") %>%
  filter(species != "adults", species != "pups") %>%
  ggplot() +
  geom_path(aes(year, Abundance, color = species)) +
  geom_point(aes(year, Abundance, color = species)) +
  geom_hline(aes( yintercept = 0)) +
  facet_wrap(~site, nrow = 3)

```

```{r}
all_taxa %>%
  filter(site != "KBAY") %>%
  select(year, site, RFCO, SEOT) %>%
    left_join(., gak_Summary, by = "year") %>%
  left_join(., otter_summary, by = c("site", "year"))
```

```{r}
all_taxa %>%
   filter(site != "KBAY") %>%
   select(year, site, RFCO, SEOT) %>%
     left_join(., gak_Summary, by = "year") %>%
   left_join(., otter_summary %>%
               mutate( adults = as.numeric(scale(adults)),
                       pups = as.numeric(scale(pups))), by = c("site", "year")) %>%
   select(!temp) %>%
   pivot_longer(cols = !c(year, site), names_to = "species", values_to = "Abundance") %>%
   filter(species != "adults", species != "pups") %>%
   ggplot() +
   geom_path(aes(year, Abundance, color = species)) +
   geom_point(aes(year, Abundance, color = species)) +
   geom_hline(aes( yintercept = 0)) +
   facet_wrap(~site, nrow = 3)
```

```{r}
p.mat.katm <- all_taxa %>% 
   filter(site == "KATM") %>%
  filter(year != 2007) %>%
  select(!site) %>%
  select(-any_of(trash)) %>%
  select(!LEAU) %>%
  select(!UNJA) %>%
  select(!`mediaster_aequalis`) %>%
  left_join(., gak_Summary, by = "year") %>%
  left_join(., otter_summary[otter_summary$site == "KATM",], by = "year") %>%
  select(!c(temp_scaled, site, adults, pups)) %>%
  cor_pmat()

cor_katm <- all_taxa %>% 
   filter(site == "KATM") %>%
  filter(year != 2007) %>%
  select(!site) %>%
  select(-any_of(trash)) %>%
  select(!LEAU) %>%
  select(!UNJA) %>%
  select(!`mediaster_aequalis`) %>%
  left_join(., gak_Summary, by = "year") %>%
  left_join(., otter_summary[otter_summary$site == "KATM",], by = "year") %>%
  select(!c(temp_scaled, site, adults, pups)) %>%
  cor(use = "complete.obs")

hc_katm<- hclust(dist(cor_katm))

order_vec_katm <- hc_katm$order

ggcorrplot(cor_katm, p.mat = p.mat.katm, hc.order = F, insig = "stars") +
  labs(title ="KATM")
  ggsave("C:/Users/bhiggins/OneDrive - DOI/Desktop/KATM_data_dredging.png", height = 10, width = 10)

```

```{r}
p.mat.kefj <- all_taxa %>% 
   filter(site == "KEFJ") %>%
  filter(year != 2007) %>%
  select(!site) %>%
  select(-any_of(trash)) %>%
  select(!LEAU) %>%
  select(!UNJA) %>%
  select(!`mediaster_aequalis`) %>%
  left_join(., gak_Summary, by = "year") %>%
  left_join(., otter_summary[otter_summary$site == "KEFJ",], by = "year") %>%
  mutate(seot_pup_ratio = ratio) %>%
  select(!c(temp_scaled, site, adults, pups, ratio)) %>%
  cor_pmat()

cor_mat <- all_taxa %>% 
   filter(site == "KEFJ") %>%
  filter(year != 2007) %>%
  select(!site) %>%
  select(-any_of(trash)) %>%
  select(!LEAU) %>%
  select(!UNJA) %>%
  select(!`mediaster_aequalis`) %>%
  left_join(., gak_Summary, by = "year") %>%
  left_join(., otter_summary[otter_summary$site == "KEFJ",], by = "year") %>%
    mutate(seot_pup_ratio = ratio) %>%
  select(!c(temp_scaled, site, adults, pups, ratio)) %>%
  cor(use = "complete.obs")

#hc <- hclust(dist(cor_mat))

#order_vec <- hc$order

ggcorrplot(cor_mat, p.mat = p.mat.kefj,  insig = "stars", hc.order = F) +
  labs(title ="KEFJ")
 ggsave("C:/Users/bhiggins/OneDrive - DOI/Desktop/KEFJ_data_dredging.png", height = 10, width = 10)
```

keep making it more of a menace. add the snails!

```{r}
snails_upper <- snails %>%
  mutate(site = substr(SiteID, 1, 3),
         site = case_when(site == "AKP" ~ "KATM",
                          site == "PWS" ~ "WPWS",
                          site == "KEP" ~ "KEFJ")) %>%
  group_by(Year, Elevation_Position, site, SpeciesName) %>%
  summarise(count = mean(Count)) %>%
  pivot_wider(names_from = SpeciesName, values_from = count) %>%
  ungroup() %>%
    filter(Elevation_Position == "Upper (1.5 m MLLW)") %>%
  select(-Elevation_Position)

all_taxa <- left_join(all_taxa, snails_upper, by = c("year" = "Year", "site"))
```

thinking out loud. all the heatwave analyses use the heatwave as a factor (pre, during, post), but looking at the temperatures, there sort of isnt a post, or at least its less straight forward. try using temp as a continious variable?

```{r}
gak_Summary %>%
  filter(year > 2006) %>%
  mutate(lta = mean(temp_scaled)) %>%
  ggplot() +
  geom_col(aes(year, temp_scaled))
```

```{r}
nppsd <- read.csv("C:/Users/bhiggins/OneDrive - DOI/MBM_Data_Proofing/SpeciesTaxonomyNPPSDv4.csv")

obs %>%
  filter(site == "KATM") %>%
  filter(!Detect_SpeciesCode %in% c("UNCO")) %>%
  group_by(Detect_SpeciesCode) %>%
  summarise(count = n()) %>%
  arrange(desc(count)) %>%
  left_join(., nppsd, by = c("Detect_SpeciesCode" = "species_code")) %>%
  select(Detect_SpeciesCode, common_name, count) %>%
  rename(number_of_detections= count) %>%
  write.csv(., "C:/Users/bhiggins/OneDrive - DOI/Desktop/KATM_bird_species.csv")
```

thinking more: I think we should be able to bootstrap density estimates for each year, this would give us confidence intervals so we would know more about our ability to detect change

first step is to make a data frame with the transects

```{r}
obs_boot <- obs %>%
  group_by(year, site, TXNumber, Detect_SpeciesCode) %>%
  summarise(count = sum(Count)) %>%
  rename(species = Detect_SpeciesCode)

#find all unique combinations of year, site, and transect, then repeat the bird list for each of them
sampling_events_boot <- obs %>%
  distinct(year, site, TXNumber)

species_df <- species %>% 
  data.frame() %>%
  rename(species = ".")

species_list <- list()

for(i in 1:nrow(sampling_events_boot)) {
  event <- sampling_events_boot[i,]
  species_df$year <- event$year
  species_df$site <- event$site
  species_df$transect <- event$TXNumber
  
  species_list[[i]] <- species_df
}

#convert list to dataframe
sampling_events_boot <- bind_rows(species_list)

#this adds in zeros for the species not observed on each survey but doesnt make up surveys that didn't happen (as opposed to expand_grid)
obs_boot <- left_join(sampling_events_boot, obs_boot, by = c("species", "year", "site", "transect" = "TXNumber")) %>%
  mutate(count = replace_na(count, 0))

#add the effort to the counts and calculate density
bird_densities_by_transect <- left_join(obs_boot, transects, by = c("site", "year", "transect" = "TXNumber")) %>%
  select(year, site, species, transect, Length_km, Area_Sampled_km2, count) %>%
  mutate(count = as.numeric(count),
         area = as.numeric(Area_Sampled_km2),
         density = count/area,
         year = as.numeric(year))
```

```{r}
library(boot)
set.seed(47)
#create the mean function for bootstrapping
mean_stat <- function(d, i) mean(d[i], na.rm = T)

#identify each survey
distinct_samples <- sampling_events_boot %>%
 distinct(year, site) %>%
 mutate(year = as.numeric(year))

#create an object to store results in
bootstrap_estimates <- list()

#loop through each survey and bootstrap density estimates
for(i in 1:37) {
  year = as.numeric(distinct_samples$year[i])
  site = distinct_samples$site[i]
  
  df = bird_densities_by_transect %>%
          filter(site == .env$site,
               year == .env$year,
               species == "HADU")

  boot_out <- boot(data = df$density, statistic = mean_stat, R = 2000)

  ci <- boot.ci(boot_out, type = c("perc", "bca"))
  
  estimates <- data.frame(
    year = year,
    site = site,
    est  = ci$t0,
    lci  = ci$bca[4],
    uci  = ci$bca[5]
 )
  
  bootstrap_estimates[[i]] <- estimates
}

bootstrap_estimates <- bind_rows(bootstrap_estimates)
```

the graph below is funky becasue it includes estimates generated using both parks. just filtering to loop through for a single park breaks it becasue not all parks exist in all years. fix tomorow

```{r}
bootstrap_estimates %>%
  ggplot() +
  geom_ribbon(aes(x = year, ymin = lci, ymax = uci), fill = "blue", alpha = 0.2) +
  geom_path(aes(x = year, y = est)) +
  geom_point(aes(x = year, y = est)) +
  facet_wrap(~ site, ncol = 1, scales = "free_y") +
  labs(title = "HADU Bootstrap Density Estimates")
```

Ok so these aren't really any good for detecting trends. there isnt any good reason that the transects should have low variance because they aren't the same thing. you could do it by habitat category but that also might miss the point

```{r}
bootstrap_estimates <- list()

#loop through each survey and bootstrap density estimates
for(i in 1:37) {
  year = as.numeric(distinct_samples$year[i])
  site = distinct_samples$site[i]
  
  df = bird_densities_by_transect %>%
          filter(site == .env$site,
               year == .env$year,
               species == "SEOT")

  boot_out <- boot(data = df$density, statistic = mean_stat, R = 2000)

  ci <- boot.ci(boot_out, type = c("perc", "bca"))
  
  estimates <- data.frame(
    year = year,
    site = site,
    est  = ci$t0,
    lci  = ci$bca[4],
    uci  = ci$bca[5]
 )
  
  bootstrap_estimates[[i]] <- estimates
}

bootstrap_estimates <- bind_rows(bootstrap_estimates)

bootstrap_estimates %>%
  ggplot() +
  geom_ribbon(aes(x = year, ymin = lci, ymax = uci), fill = "blue", alpha = 0.2) +
  geom_path(aes(x = year, y = est)) +
  geom_point(aes(x = year, y = est)) +
  facet_wrap(~ site, ncol = 1, scales = "free_y") +
  labs(title = "SEOT Bootstrap Density Estimates")
```

```{r}
bootstrap_estimates <- list()

common_species = obs %>%
  filter(!Detect_SpeciesCode %in% c("UNCO", "BRMU")) %>%
  group_by(Detect_SpeciesCode) %>%
  summarise(count = n()) %>%
  arrange(desc(count)) %>%
  head(10) %>% #change this number for more or less species
  pull(Detect_SpeciesCode)



for(j in 1:10) {
  #pick the species
  species = common_species[j]
  
#loop through each survey and bootstrap density estimates
    for(i in 1:37) {
      species_j <- list()
      
      year = as.numeric(distinct_samples$year[i])
      site = distinct_samples$site[i]
  
      df = bird_densities_by_transect %>%
          filter(site == .env$site,
               year == .env$year,
               species == .env$species)

      boot_out <- boot(data = df$density, statistic = mean_stat, R = 2000)

      ci <- boot.ci(boot_out, type = c("perc"))
  
      estimates <- data.frame(
        year = year,
        site = site,
        species = species,
        est  = ci$t0,
        lci  = ci$percent[4],
        uci  = ci$percent[5]
    )
  
      species_j[[j]] <- estimates
    }
  bootstrap_estimates[[j]] <- bind_rows(species_j)
}

bootstrap_estimates <- bind_rows(bootstrap_estimates)
```

```{r}
library(dplyr)
library(purrr)
library(boot)

# Statistic for boot(): must accept (data, indices) and return a single number
mean_stat <- function(data, indices) mean(data[indices], na.rm = TRUE)

bootstrap_estimates <- vector("list", length = 10)

common_species <- obs %>%
  filter(!Detect_SpeciesCode %in% c("UNCO", "BRMU")) %>%
  group_by(Detect_SpeciesCode) %>%
  summarise(count = n(), .groups = "drop") %>%
  arrange(desc(count)) %>%
  head(10) %>%
  pull(Detect_SpeciesCode)

# Drive inner loop by the actual size of distinct_samples
n_groups <- nrow(distinct_samples)

for (j in seq_along(common_species)) {
  species <- common_species[j]

  # Create the list ONCE per species
  species_rows <- vector("list", length = n_groups)

  for (i in seq_len(n_groups)) {
    year <- distinct_samples$year[i]
    site <- distinct_samples$site[i]

    # Ensure types are aligned; use .env to refer to external variables
    df <- bird_densities_by_transect %>%
      filter(site == .env$site,
             year == .env$year,
             species == .env$species)

    if (nrow(df) == 0 || all(is.na(df$density))) {
      # Nothing to bootstrap for this (species, year, site)
      species_rows[[i]] <- data.frame(
        year    = year,
        site    = site,
        species = species,
        est     = NA_real_,
        lci     = NA_real_,
        uci     = NA_real_
      )
      next
    }

    # Bootstrap
    boot_out <- boot(data = df$density, statistic = mean_stat, R = 2000)

    # Percentile CI; guard extraction
    ci <- try(boot.ci(boot_out, type = "perc"), silent = TRUE)

    lci <- if (is.list(ci) && !is.null(ci$percent)) ci$percent[4] else NA_real_
    uci <- if (is.list(ci) && !is.null(ci$percent)) ci$percent[5] else NA_real_

    species_rows[[i]] <- data.frame(
      year    = year,
      site    = site,
      species = species,
      est     = boot_out$t0,  # use boot's point estimate
      lci     = lci,
      uci     = uci
    )
  }

  # Bind all (year, site) rows for this species
  bootstrap_estimates[[j]] <- dplyr::bind_rows(species_rows)
}

# Final table: all species × (year, site)
bootstrap_estimates <- dplyr::bind_rows(bootstrap_estimates)
bootstrap_estimates
```

```{r}
bootstrap_estimates %>%
  filter(site != "KBAY") %>%
  ggplot() +
  geom_errorbar(aes(x = year, ymin = lci, ymax = uci), fill = "blue", alpha = 0.2) +
  geom_path(aes(x = year, y = est)) +
  geom_point(aes(x = year, y = est)) +
  facet_grid(species~ site, scales = "free_y")
```

```{r}
bootstrap_estimates %>%
  filter(site != "KBAY") %>%
ggplot( aes(year, est)) +
  geom_smooth(aes(ymin = lci, ymax = uci, color = site, group = site, fill = site),
              stat = "identity",
              alpha = 0.2) +
  geom_point() +
  facet_wrap(~ species, scales = "free_y", ncol = 2) +
  labs(title = "Bootstrapped average densities")

```

instead of using transects we could use habitat type, this should have less variance than transect and maybe make some better confidence intervals...?

import shorezone

```{r}
library(sf)
# List feature classes inside the .gdb
st_layers("C:/PRO_PROJECTS/PRO_SHARING/ShoreZone_pro_2026_0530_26ad97/commondata/shorezone.gdb")

# Read one layer
HabitatClass <- st_read("C:/PRO_PROJECTS/PRO_SHARING/ShoreZone_pro_2026_0530_26ad97/commondata/shorezone.gdb", layer = "HabitatClass")
CoastalClass <- st_read("C:/PRO_PROJECTS/PRO_SHARING/ShoreZone_pro_2026_0530_26ad97/commondata/shorezone.gdb", layer = "CoastalClass")
BiologicalWaveExposure <- st_read("C:/PRO_PROJECTS/PRO_SHARING/ShoreZone_pro_2026_0530_26ad97/commondata/shorezone.gdb", layer = "BiologicalWaveExposure")
ShoreType <- st_read("C:/PRO_PROJECTS/PRO_SHARING/ShoreZone_pro_2026_0530_26ad97/commondata/shorezone.gdb", layer = "ShoreType")

```

```{r}
obs_sf <- obs %>%
  st_as_sf(., coords = c("Long", "Lat"), crs = 4326)

HabitatClass <- st_transform(HabitatClass, 4326)
CoastalClass <- st_transform(CoastalClass, 4326)
BiologicalWaveExposure <- st_transform(BiologicalWaveExposure, 4326)
ShoreType <- st_transform(ShoreType, 4326)
```

```{r}
library(leaflet)

bbox <- st_bbox(obs_sf)

bbox_poly <- st_as_sfc(bbox)

HabitatClass <- st_filter(HabitatClass, bbox_poly)
CoastalClass <- st_filter(CoastalClass, bbox_poly)
BiologicalWaveExposure <- st_filter(BiologicalWaveExposure, bbox_poly)
ShoreType <- st_filter(BiologicalWaveExposure, bbox_poly)

#join habitat parameters with bird obs
obs_sf <- st_join(obs_sf, HabitatClass, join = st_nearest_feature)
obs_sf <- st_join(obs_sf, CoastalClass, join = st_nearest_feature)
obs_sf <- st_join(obs_sf, BiologicalWaveExposure, join = st_nearest_feature)
obs_sf <- st_join(obs_sf, ShoreType, join = st_nearest_feature)
```

now we need to see how many habitats we have

```{r}
obs_sf %>%
  st_drop_geometry() %>%
  group_by(HAB_CLASS_Name) %>%
  summarise(counts = n()) %>%
  arrange(desc(counts)) %>%
  ggplot() +
  geom_col(aes(HAB_CLASS_Name, counts)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
library(vegan)

bc.table <- bird_densities %>% 
  filter(!species %in% c("SOSA", "SEOT", "BRBE", "HASE", "STSL", "BLBE", "STJA", "MOOS", "GOEA", "HAPO", "DAPO", "GRWO", "RLHA", "GRWH", "KIWH", "NOHA", "FOSP", "TRES", "RUHU", "MIWH", "PIGR", "BOGU", "UNPE", "BOAT", "YELL", "PORC", "HUWH", "MACR", "UNWH", "REFO", "MOGO")) %>%
  select(-c(count, length, area)) %>%
  pivot_wider(names_from = species, values_from = density) %>%
  filter(!year %in% c(2006, 2011)) %>%
  filter(site != "KBAY")

#bray-curtis distance matrix
bc.matrix <- as.matrix(bc.table[,3:ncol(bc.table)], labels = T) #why did I subset this?
row.names(bc.matrix) <- paste(bc.table$site, bc.table$year, sep = "_")
bc.matrix <- vegdist(bc.matrix, method = "bray")
```

```{r}
by_survey_mds <- metaMDS(bc.matrix,
          distance = "bray",
          k = 3,
          maxit = 999, 
          trymax = 500,
          wascores = TRUE)
```

```{r}
data.scores = as.data.frame(scores(by_survey_mds))
data.scores <- data.scores %>%
  mutate(site = substr(row.names(.), start = 1, 4),
         year = substr(rownames(.), 6, 10))

data.scores <- data.scores %>%
  mutate(
    year = as.numeric(year),
    blob = case_when(
      year < 2014 ~ "pre",
      year >= 2014 ~ "post"
    ),
    salmon = case_when(
      year %% 2 == 0 ~ "even",
      year %% 2 == 1 ~ "odd"
    )
  )
```

```{r}
hulls <- data.scores %>%
  group_by(salmon) %>%
  slice(chull(NMDS1, NMDS2))

okabe_ito <- c(
  "#E69F00", "#56B4E9", "#009E73",
  "#0072B2", "#D55E00", "#CC79A7"
)

ggplot(data.scores, aes(NMDS1, NMDS2)) +
  geom_polygon(
    data = hulls,
    aes(fill = salmon, group = salmon),
    color = NA, alpha = 0.35
  ) +
  geom_point(
    aes(color = site, shape = site),
    size = 3, stroke = 0.4
  ) +
  scale_fill_viridis_d(option = "C") +       # options: "A","B","C","D","E"
  scale_color_manual(values = okabe_ito) +
  theme_minimal(base_size = 12)

```
try a lunatic exercise in plotting nmds scores against temp. this is basically an attempt to look at the blob using continuous variables
```{r}
temp_df %>%
  filter(block == "KEFJ" | block == "KATM") %>%
  left_join(data.scores, ., by = c("year", "site" = "block")) %>%
  mutate(avg_temp = as.numeric(scale(avg_temp))) %>%
  ggplot() +
  geom_path(aes(year, NMDS1, color = site, group = site)) +
  geom_path(aes(year, avg_temp, group = site))
```  
```{r}
temp_df %>%
  filter(block == "KEFJ" | block == "KATM") %>%
  mutate(year = as.numeric(year)) %>%
  left_join(data.scores, ., by = c("year", "site" = "block")) %>%
  mutate(avg_temp = as.numeric(scale(avg_temp))) %>%
  ggplot() +
  geom_point(aes(avg_temp, NMDS1))
```
now try grouping birds based on family
```{r}
bc.table <- bird_densities %>% 
  filter(!species %in% c("SOSA", "SEOT", "BRBE", "HASE", "STSL", "BLBE", "STJA", "MOOS", "GOEA", "HAPO", "DAPO", "GRWO", "RLHA", "GRWH", "KIWH", "NOHA", "FOSP", "TRES", "RUHU", "MIWH", "PIGR", "BOGU", "UNPE", "BOAT", "YELL", "PORC", "HUWH", "MACR", "UNWH", "REFO", "MOGO")) %>%
  select(-c(count, length, area)) %>%
  pivot_wider(names_from = species, values_from = density) %>%
  filter(!year %in% c(2006, 2011)) %>%
  filter(site != "KBAY")

bird_densities$species[bird_densities$species == "NOCR"] <- "AMCR"
bird_densities$species[bird_densities$species == "NOSH"] <- "NSHO"
bird_densities$species[bird_densities$species == "HEGU"] <- "HERG"

family.table <- bird_densities %>%
  filter(!species %in% c("SOSA", "SEOT", "BRBE", "HASE", "STSL", "BLBE", "STJA", "MOOS", "GOEA", "HAPO", "DAPO", "GRWO", "RLHA", "GRWH", "KIWH", "NOHA", "FOSP", "TRES", "RUHU", "MIWH", "PIGR", "BOGU", "UNPE", "BOAT", "YELL", "PORC", "HUWH", "MACR", "UNWH", "REFO", "MOGO", "PEFA", "RIOT", "HBWH", "GYRF", "UNJA")) %>%
  left_join(., nppsd, by = c("species" = "species_code")) %>%
  group_by(year, site, family) %>%
  summarise(count = sum(count),
            density = sum(log(density + 1))) %>%
  mutate(family = ifelse(family == "Laridae\xa0", "Laridae", family)) %>%
  select(-c(count)) %>%
  pivot_wider(names_from = family, values_from = density) %>%
  filter(!year %in% c(2006, 2011)) %>%
  filter(site != "KBAY") %>%
  ungroup()
```
```{r}
family.matrix <- as.matrix(family.table[,3:ncol(family.table)], labels = T) #why did I subset this?
row.names(family.matrix) <- paste(family.table$site, family.table$year, sep = "_")
family.matrix <- vegdist(family.matrix, method = "bray")
```

```{r}
family_mds <- metaMDS(family.matrix,
          distance = "bray",
          k = 3,
          maxit = 999, 
          trymax = 500,
          wascores = TRUE)
```

```{r}
data.scores = as.data.frame(scores(family_mds))
data.scores <- data.scores %>%
  mutate(site = substr(row.names(.), start = 1, 4),
         year = substr(rownames(.), 6, 10))

data.scores <- data.scores %>%
  mutate(
    year = as.numeric(year),
    blob = case_when(
      year < 2014 ~ "pre",
      year >= 2014 ~ "post"
    ),
    salmon = case_when(
      year %% 2 == 0 ~ "even",
      year %% 2 == 1 ~ "odd"
    )
  )
```

```{r}
hulls <- data.scores %>%
  group_by(blob, site) %>%
  slice(chull(NMDS1, NMDS2))

ggplot(data.scores, aes(NMDS1, NMDS2)) +
  geom_polygon(
    data = hulls,
    aes(fill = blob, group = blob),
    color = NA, alpha = 0.35
  ) +
  geom_point(
    aes(color = blob),
    size = 3, stroke = 0.4
  ) +       # options: "A","B","C","D","E"
  theme_minimal(base_size = 12) +
  facet_wrap(~site, scales = "free")
```

```{r}
#order families taxonomically
families <- bird_densities %>%
  filter(!species %in% c("SOSA", "SEOT", "BRBE", "HASE", "STSL", "BLBE", "STJA", "MOOS", "GOEA", "HAPO", "DAPO", "GRWO", "RLHA", "GRWH", "KIWH", "NOHA", "FOSP", "TRES", "RUHU", "MIWH", "PIGR", "BOGU", "UNPE", "BOAT", "YELL", "PORC", "HUWH", "MACR", "UNWH", "REFO", "MOGO", "PEFA", "RIOT", "HBWH", "GYRF", "UNJA")) %>%
  left_join(., nppsd, by = c("species" = "species_code")) %>%
  group_by(year, site, family) %>%
  summarise(count = sum(count),
            density = sum(log(density + 1))) %>%
  pull(family) %>%
  unique()

family_order <- data.frame(families) %>%
  mutate(keep = 1)
family_order <- left_join(nppsd, family_order, by = c("family" = "families")) %>%
  mutate(family = ifelse(family == "Laridae\xa0", "Laridae", family)) %>%
  filter(keep == 1) %>%
  pull(family) %>%
  unique() %>%
  as.vector()


#palette <- colorRampPalette(brewer.pal(9, "Spectral"))(17)
#palette <- sample(palette)   # randomize order

palette <- c("#F2FAAB", "#FDC775", "#88CFA4", "#FEEFA4", "#E6F598", "#ABDDA4", "#4CA4B1", "#C8E99E", "#FFFFBF", "#FEE08B", "#D53E4F", "#F88D51", "#66C2A5", "#F46D43", "#3288BD", "#E45549", "#FDAE61")

bird_densities %>%
  filter(!species %in% c("SOSA", "SEOT", "BRBE", "HASE", "STSL", "BLBE", "STJA", "MOOS", "GOEA", "HAPO", "DAPO", "GRWO", "RLHA", "GRWH", "KIWH", "NOHA", "FOSP", "TRES", "RUHU", "MIWH", "PIGR", "BOGU", "UNPE", "BOAT", "YELL", "PORC", "HUWH", "MACR", "UNWH", "REFO", "MOGO", "PEFA", "RIOT", "HBWH", "GYRF", "UNJA")) %>%
  left_join(., nppsd, by = c("species" = "species_code")) %>%
  group_by(year, site, family) %>%
  summarise(count = sum(count),
            density = sum(log(density + 1))) %>%
  mutate(family = ifelse(family == "Laridae\xa0", "Laridae", family),
         family = factor(family, levels = family_order)) %>%
  select(-c(count)) %>%
  ggplot(aes(x = year, y = density, fill = family)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    y = "Percent of total density",
    x = "Year",
    fill = "Family"
  ) +
  theme_minimal() +
  facet_wrap(~site, ncol = 1) +
  scale_fill_manual(values = palette)

```
```{r}
bird_densities %>%
  filter(!species %in% c("SOSA", "SEOT", "BRBE", "HASE", "STSL", "BLBE", "STJA", "MOOS", "GOEA", "HAPO", "DAPO", "GRWO", "RLHA", "GRWH", "KIWH", "NOHA", "FOSP", "TRES", "RUHU", "MIWH", "PIGR", "BOGU", "UNPE", "BOAT", "YELL", "PORC", "HUWH", "MACR", "UNWH", "REFO", "MOGO", "PEFA", "RIOT", "HBWH", "GYRF", "UNJA")) %>%
  left_join(., nppsd, by = c("species" = "species_code")) %>%
  group_by(year, site, species) %>%
  summarise(count = sum(count),
            density = sum(log(density + 1))) %>%
  select(-c(count)) %>%
  ggplot(aes(x = year, y = density, fill = species)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    y = "Percent of total density",
    x = "Year",
    fill = "Family"
  ) +
  theme_minimal() +
  facet_wrap(~site, ncol = 1)
```
ok the plan for today is to do this by diet and by functional group
```{r}
#bird_densities %>%
#  filter(!species %in% c("SOSA", "SEOT", "BRBE", "HASE", "STSL", "BLBE", "STJA", "MOOS", "GOEA", "HAPO", "DAPO", "GRWO", "RLHA", "GRWH", "KIWH", "NOHA", "FOSP", "TRES", "RUHU", #"MIWH", "PIGR", "BOGU", "UNPE", "BOAT", "YELL", "PORC", "HUWH", "MACR", "UNWH", "REFO", "MOGO", "PEFA", "RIOT", "HBWH", "GYRF", "UNJA")) %>%
#  pull(species) %>%
#  unique() %>%
#  data.frame() %>%
#  rename(species_code = ".") %>%
#  left_join(., nppsd, by = c("species_code" = "species_code")) %>%
#  write.csv(., row.names = FALSE, "data/functional_groups_template.csv")
  
functional_groups <- read.csv("data/functional_groups.csv")
diet_palette <- sample(colorRampPalette(brewer.pal(9, "Spectral"))(5))
diet_palette <- c("#FFFFBF", "#3288BD",  "#D53E4F", "#ABDDA4",  "#FDAE61")
bird_densities %>%
  filter(!species %in% c("SOSA", "SEOT", "BRBE", "HASE", "STSL", "BLBE", "STJA", "MOOS", "GOEA", "HAPO", "DAPO", "GRWO", "RLHA", "GRWH", "KIWH", "NOHA", "FOSP", "TRES", "RUHU", "MIWH", "PIGR", "BOGU", "UNPE", "BOAT", "YELL", "PORC", "HUWH", "MACR", "UNWH", "REFO", "MOGO", "PEFA", "RIOT", "HBWH", "GYRF", "UNJA")) %>%
  left_join(., nppsd, by = c("species" = "species_code")) %>%
  left_join(., functional_groups, by = c("species" = "species_code")) %>%
  group_by(year, site, diet) %>%
  summarise(count = sum(count),
            density = sum(log(density + 1))) %>%
  select(-c(count)) %>%
  ggplot(aes(x = year, y = density, fill = diet)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    y = "Percent of total density",
    x = "Year",
    fill = "Diet"
  ) +
  theme_minimal() +
  facet_wrap(~site, ncol = 1) +
  scale_fill_manual(values = diet_palette)
```

```{r}
bird_densities %>%
  filter(!species %in% c("SOSA", "SEOT", "BRBE", "HASE", "STSL", "BLBE", "STJA", "MOOS", "GOEA", "HAPO", "DAPO", "GRWO", "RLHA", "GRWH", "KIWH", "NOHA", "FOSP", "TRES", "RUHU", "MIWH", "PIGR", "BOGU", "UNPE", "BOAT", "YELL", "PORC", "HUWH", "MACR", "UNWH", "REFO", "MOGO", "PEFA", "RIOT", "HBWH", "GYRF", "UNJA")) %>%
  left_join(., nppsd, by = c("species" = "species_code")) %>%
  left_join(., functional_groups, by = c("species" = "species_code")) %>%
  group_by(year, site, functional_group) %>%
  summarise(count = sum(count),
            density = sum(log(density + 1))) %>%
  select(-c(count)) %>%
  ggplot(aes(x = year, y = density, fill = functional_group)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    y = "Percent of total density",
    x = "Year",
    fill = "Functional Group"
  ) +
  theme_minimal() +
  facet_wrap(~site, ncol = 1) +
  scale_fill_manual(values = colorRampPalette(brewer.pal(9, "Spectral"))(3))
```
make ecological groups by comining diet and foraging strategy
```{r}
#niche_palette <- sample(colorRampPalette(brewer.pal(9, "Spectral"))(11))

niche_palette <- c("#3288BD", "#C2E69F", "#EBF79F", "#5BB6A9", "#FFFFBF", "#FDC271", "#8FD2A4", "#FEE695", "#F99455", "#D53E4F", "#ED6345")


bird_densities %>%
  filter(!species %in% c("SOSA", "SEOT", "BRBE", "HASE", "STSL", "BLBE", "STJA", "MOOS", "GOEA", "HAPO", "DAPO", "GRWO", "RLHA", "GRWH", "KIWH", "NOHA", "FOSP", "TRES", "RUHU", "MIWH", "PIGR", "BOGU", "UNPE", "BOAT", "YELL", "PORC", "HUWH", "MACR", "UNWH", "REFO", "MOGO", "PEFA", "RIOT", "HBWH", "GYRF", "UNJA")) %>%
  left_join(., nppsd, by = c("species" = "species_code")) %>%
  left_join(., functional_groups, by = c("species" = "species_code")) %>%
  mutate(niche = paste(functional_group, diet, sep = " ")) %>%
  group_by(year, site, niche) %>%
  summarise(count = sum(count),
            density = sum(log(density + 1))) %>%
  select(-c(count)) %>%
  ggplot(aes(x = year, y = density, fill = niche)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    y = "Percent of total density",
    x = "Year",
    fill = "Niche"
  ) +
  theme_minimal() +
  facet_wrap(~site, ncol = 1) +
  scale_fill_manual(values = niche_palette)
```

```{r}
bird_densities %>%
  filter(!species %in% c("SOSA", "SEOT", "BRBE", "HASE", "STSL", "BLBE", "STJA", "MOOS", "GOEA", "HAPO", "DAPO", "GRWO", "RLHA", "GRWH", "KIWH", "NOHA", "FOSP", "TRES", "RUHU", "MIWH", "PIGR", "BOGU", "UNPE", "BOAT", "YELL", "PORC", "HUWH", "MACR", "UNWH", "REFO", "MOGO", "PEFA", "RIOT", "HBWH", "GYRF", "UNJA")) %>%
  left_join(., nppsd, by = c("species" = "species_code")) %>%
  left_join(., functional_groups, by = c("species" = "species_code")) %>%
  mutate(niche = paste(functional_group, diet, sep = " ")) %>%
  group_by(year, site, niche) %>%
  summarise(count = sum(count),
            density = sum(log(density + 1))) %>%
  mutate(density_scaled = as.numeric(scale(density))) %>%
  ggplot(aes(year, density_scaled, color = niche, group = niche)) +
  geom_path() +
  geom_point() +
  facet_wrap(~site, ncol = 1) +
  theme_minimal()
```

ok the question is how redundant these groups are. figure out how to plot that
```{r}
library(ggridges)

bird_densities %>%
  filter(!species %in% c("SOSA", "SEOT", "BRBE", "HASE", "STSL", "BLBE", "STJA", "MOOS", "GOEA", "HAPO", "DAPO", "GRWO", "RLHA", "GRWH", "KIWH", "NOHA", "FOSP", "TRES", "RUHU", "MIWH", "PIGR", "BOGU", "UNPE", "BOAT", "YELL", "PORC", "HUWH", "MACR", "UNWH", "REFO", "MOGO", "PEFA", "RIOT", "HBWH", "GYRF", "UNJA")) %>%
  left_join(., nppsd, by = c("species" = "species_code")) %>%
  left_join(., functional_groups, by = c("species" = "species_code")) %>%
  mutate(niche = paste(functional_group, diet, sep = " ")) %>%
  filter(density != 0) %>%
  group_by(year, site, niche) %>%
  summarise(count = sum(count),
            density = sum(log(density + 1)),
            n_species = n_distinct(species)) %>%
  ungroup() %>%
  ggplot() +
  geom_col(aes(year, n_species, fill = niche), position = "dodge") +
  labs(
    y = "Number of Species per Niche",
    x = "Year",
    fill = "Niche"
  ) +
  theme_minimal() +
  facet_wrap(~site, ncol = 1) +
  labs(title = "Ecosystem redundancy",
       subtitle = "Number of species per niche across year")
```
```{r}
bird_densities %>%
  filter(!species %in% c("SOSA", "SEOT", "BRBE", "HASE", "STSL", "BLBE", "STJA", "MOOS", "GOEA", "HAPO", "DAPO", "GRWO", "RLHA", "GRWH", "KIWH", "NOHA", "FOSP", "TRES", "RUHU", "MIWH", "PIGR", "BOGU", "UNPE", "BOAT", "YELL", "PORC", "HUWH", "MACR", "UNWH", "REFO", "MOGO", "PEFA", "RIOT", "HBWH", "GYRF", "UNJA")) %>%
  left_join(., nppsd, by = c("species" = "species_code")) %>%
  left_join(., functional_groups, by = c("species" = "species_code")) %>%
  mutate(niche = paste(functional_group, diet, sep = " ")) %>%
  filter(density != 0) %>%
  group_by(year, site, diet) %>%
  summarise(count = sum(count),
            density = sum(log(density + 1)),
            n_species = n_distinct(species)) %>%
  ungroup() %>%
  ggplot() +
  geom_col(aes(year, n_species, fill = diet), position = "dodge") +
  labs(
    y = "Number of Species per Niche",
    x = "Year",
    fill = "Niche"
  ) +
  theme_minimal() +
  facet_wrap(~site, ncol = 1) +
  labs(title = "Ecosystem redundancy",
       subtitle = "number of species per diet category")
```

```{r}
bird_densities %>%
  filter(!species %in% c("SOSA", "SEOT", "BRBE", "HASE", "STSL", "BLBE", "STJA", "MOOS", "GOEA",
                         "HAPO", "DAPO", "GRWO", "RLHA", "GRWH", "KIWH", "NOHA", "FOSP", "TRES",
                         "RUHU", "MIWH", "PIGR", "BOGU", "UNPE", "BOAT", "YELL", "PORC", "HUWH",
                         "MACR", "UNWH", "REFO", "MOGO", "PEFA", "RIOT", "HBWH", "GYRF", "UNJA")) %>%
  left_join(nppsd, by = c("species" = "species_code")) %>%
  left_join(functional_groups, by = c("species" = "species_code")) %>%
  mutate(niche = paste(functional_group, diet, sep = " ")) %>%
  filter(density != 0) %>%
  filter(!niche %in% c("land duck food", "land fish", "surface inverts")) %>%
  group_by(year, site, niche) %>%
  summarise(
    count = sum(count),
    density = sum(log(density + 1)),
    n_species = n_distinct(species),
    .groups = "drop"
  ) %>%
  group_by(site, niche) %>%          # compute anomaly baseline per niche
  mutate(mean_species = mean(n_species),
         anomaly = n_species - mean_species) %>%
  ungroup() %>%
  ggplot(aes(x = year, y = anomaly, fill = site, group = site)) +
  geom_col(position = "dodge") +
  labs(
    y = "Species Anomaly (Deviation From Mean)",
    x = "Year",
    fill = "Niche"
  ) +
  theme_minimal() +
  facet_wrap(~niche, ncol = 1) +
  geom_hline(yintercept = 0)
```
this should probably be modeled...
```{r}
niche_df <- bird_densities %>%
  filter(!species %in% c("SOSA", "SEOT", "BRBE", "HASE", "STSL", "BLBE", "STJA", "MOOS", "GOEA",
                         "HAPO", "DAPO", "GRWO", "RLHA", "GRWH", "KIWH", "NOHA", "FOSP", "TRES",
                         "RUHU", "MIWH", "PIGR", "BOGU", "UNPE", "BOAT", "YELL", "PORC", "HUWH",
                         "MACR", "UNWH", "REFO", "MOGO", "PEFA", "RIOT", "HBWH", "GYRF", "UNJA")) %>%
  left_join(nppsd, by = c("species" = "species_code")) %>%
  left_join(functional_groups, by = c("species" = "species_code")) %>%
  mutate(niche = paste(functional_group, diet, sep = " ")) %>%
  filter(density != 0) %>%
  filter(!niche %in% c("land duck food", "land fish", "surface inverts")) %>%
  group_by(year, site, niche) %>%
  summarise(
    count = sum(count),
    density = sum(log(density + 1)),
    n_species = n_distinct(species),
    .groups = "drop"
  ) %>%
  group_by(site, niche) %>%          # compute anomaly baseline per niche
  mutate(mean_species = mean(n_species),
         anomaly = n_species - mean_species) %>%
  ungroup() %>%
  mutate(year = as.numeric(year)) %>%
  left_join(., gak_Summary)

niche_df$blob[niche_df$year < 2014] <- "pre"
niche_df$blob[niche_df$year >= 2014] <- "post"

glm(density ~ temp*year + niche + site + year + blob, data = niche_df) %>% summary()
fit <- glm(density ~ temp*year + niche + site + year + blob, data = niche_df)

effect_data <- ggpredict(fit, terms = c("niche", "site", "year"))

plot(effect_data)
```
```{r}
library(bcp)
gak_Summary %>%
  ggplot(aes(year, temp)) +
  geom_path() +
  geom_point()

set.seed(101)
bcp_gak1 = bcp(gak_Summary$temp, p0 = 0.008, burnin = 1000, 
             mcmc = 10000)

plot(bcp_gak1, main = "GAK1 Temp changepoints")
```
```{r}
diver_bcp <- niche_df %>%
  filter(niche == "diving fish")
bcp_gak1 = bcp(diver_bcp$anomaly, p0 = 0.008, burnin = 1000, 
             mcmc = 10000)

plot(bcp_gak1, main = "GAK1 Temp changepoints")
```

```{r}
library(DirichletReg)

niche_df <- bird_densities %>%
  filter(!species %in% c("SOSA", "SEOT", "BRBE", "HASE", "STSL", "BLBE", "STJA", "MOOS", "GOEA",
                         "HAPO", "DAPO", "GRWO", "RLHA", "GRWH", "KIWH", "NOHA", "FOSP", "TRES",
                         "RUHU", "MIWH", "PIGR", "BOGU", "UNPE", "BOAT", "YELL", "PORC", "HUWH",
                         "MACR", "UNWH", "REFO", "MOGO", "PEFA", "RIOT", "HBWH", "GYRF", "UNJA")) %>%
  left_join(nppsd, by = c("species" = "species_code")) %>%
  left_join(functional_groups, by = c("species" = "species_code")) %>%
  mutate(niche = paste(functional_group, diet, sep = " ")) %>%
  filter(density != 0) %>%
  filter(!niche %in% c("land duck food", "land fish", "surface inverts")) %>%
  group_by(year, site, niche) %>%
  summarise(prop = density / sum(density))

prop_df <- bird_densities %>%
filter(!species %in% c("SOSA", "SEOT", "BRBE", "HASE", "STSL", "BLBE", "STJA", "MOOS", "GOEA",
                         "HAPO", "DAPO", "GRWO", "RLHA", "GRWH", "KIWH", "NOHA", "FOSP", "TRES",
                         "RUHU", "MIWH", "PIGR", "BOGU", "UNPE", "BOAT", "YELL", "PORC", "HUWH",
                         "MACR", "UNWH", "REFO", "MOGO", "PEFA", "RIOT", "HBWH", "GYRF", "UNJA")) %>%
  left_join(nppsd,              by = c("species" = "species_code")) %>%
  left_join(functional_groups,  by = c("species" = "species_code")) %>%
  mutate(niche = paste(functional_group, diet, sep = " ")) %>%
  filter(density != 0,
         !niche %in% c("land duck food", "land fish", "surface inverts")) %>%
  # 1) Aggregate to niche within site-year
  group_by(year, site, niche) %>%
  summarise(density_niche = sum(log(density + 1)), .groups = "drop") %>%
  # 2) Compute proportions across niches within each site-year
  group_by(year, site) %>%
  mutate(prop = density_niche / sum(density_niche)) %>%
  ungroup()


prop_wide <- prop_df %>% 
  select(-density_niche) %>%
  pivot_wider(names_from = niche, values_from = prop) %>%
  mutate(across(3:10, ~ replace_na(., 0))) %>%
  select(-`diving plankton`) %>%
  mutate(year = as.numeric(year)) %>%
  left_join(gak_Summary, by = "year")
```

```{r}
prop_df %>%
  filter(site != "KBAY") %>%
  ggplot(aes(year, prop, color = niche, group = niche)) +
  geom_path() +
  facet_wrap(~site, ncol = 1) +
  labs(title = "Proportion of total abundance by niche")
```

```{r}
library(streamgraph)

prop_df %>%
  filter(site == "KEFJ") %>%
  streamgraph(key = "niche", value = "density_niche", date = "year", offset = "silhouette") %>%
  sg_axis_x(1, "year", "%Y") %>%
  sg_legend(show=TRUE, label="Niches: ") %>%
  sg_title(title = "KEFJ")

```

```{r}
prop_wide %>%
  ggplot(aes(`diving fish`, `diving inverts`)) +
  geom_point(aes(color = site)) +
  geom_smooth(method = "lm") +
  labs(title = "Diving invert ~ diving fish") +
  facet_wrap(~site, ncol = 1, scales = "free")
```


```{r}
library(statmod)
#try some models
prop_wide <- janitor::clean_names(prop_wide)
fit.null <- glm(diving_fish ~ 1, data = prop_wide, family = Gamma(link = "log"))
fit.1 <- glm(diving_fish ~ year, data = prop_wide, family = Gamma(link = "log"))
fit.2 <- glm(diving_fish ~ year + site, data = prop_wide, family = Gamma(link = "log"))
fit.3 <- glm(diving_fish ~ year + site + diving_inverts, data = prop_wide, family = Gamma(link = "log"))
fit.4 <- glm(diving_fish ~ year + site + temp, data = prop_wide, family = Gamma(link = "log"))
fit.5 <- glm(diving_fish ~ temp, data = prop_wide, family = Gamma(link = "log"))
fit.6 <- glm(diving_fish ~ diving_inverts, data = prop_wide, family = Gamma(link = "log"))

AIC(fit.null, fit.1, fit.2, fit.3, fit.4, fit.5, fit.6) %>% arrange(AIC)
```

```{r}
summary(fit.5)
```
see if there is a way to make these into good models. 
why is the above graph such a good relationship but the fish eaters vs benthivores relationship wasnt obvious from the streamplot
```{r}
library(ggeffects)

predict.5 <- predict_response(fit.5, terms = "temp")

ggplot(predict.5, aes(x, predicted)) +
  geom_line() +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.1)
```
```{r}
ggcorrplot(corr = cor(prop_wide[,-2]), lab = T)
```
lord help us, we're adding atmospheric indices to the model....
```{r read in climate indices}
#climate indices from: https://psl.noaa.gov/enso/dashboard.html
enso <- read_table("data/meiv2.data", col_names = FALSE, 
     skip = 1) %>%
  slice(1:(n() - 5))

colnames(enso) <- c("year", "DJ", "JF", "FM", "MA", "AM",	"MJ", "JJ", "JA",	"AS",	"SO",	"ON",	"ND")

enso <- enso %>%
  pivot_longer(cols = -1) %>%
  mutate(year = as.numeric(year),
         value = as.numeric(value)) %>%
  group_by(year) %>%
  summarise(mean_enso = mean(value))

pdo <- read.csv("data/pdo.timeseries.sstens.csv") %>%
  rename(date = Date,
         value = 2) %>%
  mutate(date = mdy(date)) %>%
  na.omit() %>%
  slice(1:(n()-6))
```

```{r}

enso %>%
  mutate(sign = if_else(mean_enso >= 0, "Positive", "Negative")) %>%
  ggplot(aes(x = year, y = mean_enso)) +
  geom_col(aes(fill = sign), width = 0.9) +
  scale_fill_manual(
    values = c(Positive = "#d62728", Negative = "#3182bd"),
    guide = "none"
  ) +
  geom_hline(yintercept = 0, color = "grey40") +
  labs(x = "Year", y = "Mean ENSO", title = "ENSO: Positive (red) vs Negative (blue)") +
  theme_minimal()

```
```{r}
pdo %>%
  filter(year(date) > 2006) %>%
  mutate(sign = if_else(value >= 0, "Positive", "Negative")) %>%
  ggplot() +
  geom_col(aes(date, value, fill = sign)) +
  labs(title = "PDO index") +
  scale_fill_manual(
    values = c(Positive = "#d62728", Negative = "#3182bd"),
    guide = "none"
  )
```
now annualize the pdo and add to the modeling df
```{r}
pdo_annualized <- pdo %>%
  mutate(year = year(date)) %>%
  group_by(year) %>%
  summarise(pdo = mean(value))

prop_wide <- left_join(prop_wide, pdo_annualized, by = "year") %>%
  left_join(., enso, by = "year")
```

```{r}
prop_wide <- prop_wide %>%
  filter(site != "KBAY")

fit.null <- glm(diving_fish ~ 1, data = prop_wide, family = Gamma(link = "log"))
fit.1 <- glm(diving_fish ~ year, data = prop_wide, family = Gamma(link = "log"))
fit.2 <- glm(diving_fish ~ year + site, data = prop_wide, family = Gamma(link = "log"))
fit.3 <- glm(diving_fish ~ year + site + diving_inverts, data = prop_wide, family = Gamma(link = "log"))
fit.4 <- glm(diving_fish ~ year + site + temp, data = prop_wide, family = Gamma(link = "log"))
fit.5 <- glm(diving_fish ~ temp, data = prop_wide, family = Gamma(link = "log"))
fit.6 <- glm(diving_fish ~ diving_inverts, data = prop_wide, family = Gamma(link = "log"))

fit.7 <- glm(diving_fish ~ pdo, data = prop_wide, family = Gamma(link = "log"))
fit.8 <- glm(diving_fish ~ mean_enso, data = prop_wide, family = Gamma(link = "log"))
fit.9 <- glm(diving_fish ~ diving_inverts + mean_enso, data = prop_wide, family = Gamma(link = "log"))
fit.10 <- glm(diving_fish ~ diving_inverts + temp + mean_enso + pdo, data = prop_wide, family = Gamma(link = "log"))

fit.11 <- glm(diving_fish ~ diving_inverts + mean_enso + pdo + site + year, data = prop_wide, family = Gamma(link = "log"))
fit.12 <- glm(diving_fish ~ diving_inverts + pdo + site + year, data = prop_wide, family = Gamma(link = "log"))
fit.13 <- glm(diving_fish ~ diving_inverts + mean_enso + site + year, data = prop_wide, family = Gamma(link = "log"))

fit.14 <- glm(diving_fish ~ mean_enso + site + year, data = prop_wide, family = Gamma(link = "log"))
fit.15 <- glm(diving_fish ~ pdo + site + year, data = prop_wide, family = Gamma(link = "log"))

AIC(fit.null, fit.1, fit.2, fit.3, fit.4, fit.5, fit.6, fit.7, fit.8, fit.9, fit.10, fit.11, fit.12, fit.13, fit.14, fit.15) %>% arrange(AIC)

```


```{r}
library(dplyr)

# Ensure response is strictly positive for Gamma(log)
prop_wide2 <- prop_wide %>%
  mutate(
    diving_inverts_resp = pmax(diving_inverts, 1e-6),   # response
    year = as.numeric(year)                              # good practice
  )

# Fit the "same-structure" models with swapped response/predictors
fitI.null <- glm(diving_inverts_resp ~ 1, data = prop_wide2, family = Gamma(link = "log"))
fitI.1    <- glm(diving_inverts_resp ~ year, data = prop_wide2, family = Gamma(link = "log"))
fitI.2    <- glm(diving_inverts_resp ~ year + site, data = prop_wide2, family = Gamma(link = "log"))

# NOTE: where the original had + diving_inverts (predictor), we now use + diving_fish
fitI.3    <- glm(diving_inverts_resp ~ year + site + diving_fish, data = prop_wide2, family = Gamma(link = "log"))

fitI.4    <- glm(diving_inverts_resp ~ year + site + temp, data = prop_wide2, family = Gamma(link = "log"))
fitI.5    <- glm(diving_inverts_resp ~ temp, data = prop_wide2, family = Gamma(link = "log"))
fitI.6    <- glm(diving_inverts_resp ~ diving_fish, data = prop_wide2, family = Gamma(link = "log"))

fitI.7    <- glm(diving_inverts_resp ~ pdo, data = prop_wide2, family = Gamma(link = "log"))
fitI.8    <- glm(diving_inverts_resp ~ mean_enso, data = prop_wide2, family = Gamma(link = "log"))
fitI.9    <- glm(diving_inverts_resp ~ diving_fish + mean_enso, data = prop_wide2, family = Gamma(link = "log"))
fitI.10   <- glm(diving_inverts_resp ~ diving_fish + temp + mean_enso + pdo, data = prop_wide2, family = Gamma(link = "log"))

fitI.11   <- glm(diving_inverts_resp ~ diving_fish + mean_enso + pdo + site + year, data = prop_wide2, family = Gamma(link = "log"))
fitI.12   <- glm(diving_inverts_resp ~ diving_fish + pdo + site + year, data = prop_wide2, family = Gamma(link = "log"))
fitI.13   <- glm(diving_inverts_resp ~ diving_fish + mean_enso + site + year, data = prop_wide2, family = Gamma(link = "log"))

fitI.14   <- glm(diving_inverts_resp ~ mean_enso + site + year, data = prop_wide2, family = Gamma(link = "log"))
fitI.15   <- glm(diving_inverts_resp ~ pdo + site + year, data = prop_wide2, family = Gamma(link = "log"))

# AIC table (descending like your original; flip to ascending if you prefer best at top)
aic_tbl_inverts <- AIC(
  fitI.null, fitI.1, fitI.2, fitI.3, fitI.4, fitI.5, fitI.6,
  fitI.7, fitI.8, fitI.9, fitI.10, fitI.11, fitI.12, fitI.13, fitI.14, fitI.15
) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("model") %>%
  arrange(desc(AIC))

aic_tbl_inverts

```


```{r}
plot(ggpredict(beta.6, terms = "diving_fish"))
```

```{r}
library(patchwork)
pb11_mean <- ggpredict(beta.11, terms = "mean_enso")
pb11_pdo  <- ggpredict(beta.11, terms = "pdo")
pb11_fish <- ggpredict(beta.11, terms = "diving_fish")
pb11_enso <- ggpredict(beta.11, terms = "mean_enso")
pb11_year <- ggpredict(beta.11, terms = "year", type = "random")

plot(pb11_mean, colors = "blue") +
plot(pb11_pdo) /
plot(pb11_fish) +
plot(pb11_enso)/
plot(pb11_year)
```

```{r}
plot(predict_response(beta.11, terms = "year", margin = "empirical"))
```
```{r}
plot(predict_response(beta.11, terms = "diving_fish", margin = "empirical", condition = c("site" = "KATM")))
```

there's a few ways to parameraterize these depending on what we want to get out of it. 
1. is the question if birds change abundance (or proportion), or is it what effects bird niche proportions
2. thinking of the blob as binary is lame, using temp as a continuous variable makes it interesting
    a. find a good way to identify the warm years?
3. hav'nt touched the winter bird data yet. maybe that should be part of this    

steps
1. decide which temperature to use (hobo vs gak)
2. use temp as explanatory variable 
    model proportions and counts
    
1. look at all the temps data

```{r plot gak1 temp}
gak1 %>%
  filter(year > 2005) %>%
  mutate(temp = str_replace(`Temperature_[C],`, ",", "")) %>%
  mutate(temp = as.numeric(temp),
         depth = as.numeric(`Pressure_[dbar],`)) %>%
  filter(depth < 5) %>%
  group_by(year) %>%
  summarise(temp = mean(as.numeric(temp), na.rm = T)) %>%
  mutate(temp_scaled = as.numeric(scale(temp))) %>%
    mutate(avg = mean(temp),
         anomaly = temp - avg) %>%
  mutate(sign = if_else(anomaly >= 0, "Positive", "Negative"),
         year = factor(year)) %>%
  ggplot(aes(year, anomaly, fill = sign)) +
  geom_col() +
  scale_fill_manual(
    values = c(Positive = "#d62728", Negative = "#3182bd"),
    guide = "none"
  ) +
  geom_hline(yintercept = 0, color = "grey40") +
  theme_minimal()

```
this is the temp anomaly from gak1. do intertidal temps show the same pattern

```{r plot hobo temps}
temp_anom <- temperature %>%
  filter(exposure == "water",
         tide_height == 0.5,
         year > 2005) %>%
  group_by(year, block) %>%
  summarise(temp = mean(temperature, na.rm = TRUE), .groups = "drop") %>%
  group_by(block) %>%
  mutate(
    avg     = mean(temp, na.rm = TRUE),
    anomaly = temp - avg,
    sign    = if_else(anomaly >= 0, "Positive", "Negative"),
    st_dev  = sd(temp, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  mutate(year = factor(year))

ggplot(temp_anom, aes(year, anomaly, fill = sign)) +
  geom_col() +
  scale_fill_manual(values = c(Positive = "#d62728", Negative = "#3182bd"),
                    guide  = "none") +
  geom_hline(
    aes(yintercept = 0 - st_dev),
    color = "grey60",
    linetype = "dashed"
  ) +
  geom_hline(
    aes(yintercept = st_dev),
    color = "grey60",
    linetype = "dashed"
  ) +
  geom_hline(yintercept = 0, color = "grey40") +
  facet_wrap(~ block, ncol = 1) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
there are some differences in between the gak and the hobo data but they largely tell the same story. We'll use the hobo data for the analysis

```{r summarize hobo temps}
hobo_temps <- temperature %>%
  filter(exposure == "water", tide_height == 0.5, year > 2005) %>%
  group_by(year, block) %>%
  summarise(temp = mean(temperature, na.rm = TRUE), .groups = "drop") %>%
  group_by(block) %>%                              # <-- compute block-level mean
  mutate(
   # avg     = mean(temp, na.rm = TRUE),
   #anomaly = temp - avg,
    temp_scaled = as.numeric(scale(temp))
  ) %>%
  ungroup() %>%
  rename(site = block)
  
```

now combine the temperatures with the bird data
```{r summarise birds by niche}
prop_df <- bird_densities %>%
filter(!species %in% c("SOSA", "SEOT", "BRBE", "HASE", "STSL", "BLBE", "STJA", "MOOS", "GOEA",
                         "HAPO", "DAPO", "GRWO", "RLHA", "GRWH", "KIWH", "NOHA", "FOSP", "TRES",
                         "RUHU", "MIWH", "PIGR", "BOGU", "UNPE", "BOAT", "YELL", "PORC", "HUWH",
                         "MACR", "UNWH", "REFO", "MOGO", "PEFA", "RIOT", "HBWH", "GYRF", "UNJA")) %>%
  left_join(nppsd,              by = c("species" = "species_code")) %>%
  left_join(functional_groups,  by = c("species" = "species_code")) %>%
  mutate(niche = paste(functional_group, diet, sep = " ")) %>%
  filter(density != 0,
         !niche %in% c("land duck food", "land fish", "surface inverts")) %>%
  # 1) Aggregate to niche within site-year
  group_by(year, site, niche) %>%
  summarise(density_log = sum(log(density + 1)),
         count = sum(count),
         .groups = "drop") %>%
  # 2) Compute proportions across niches within each site-year
  group_by(year, site) %>%
  mutate(proportion = density_log / sum(density_log)) %>%
  ungroup()
```

```{r create full modeling dataframe}
#summarize proportions
prop_wide <- prop_df %>%
  select(-density_log) %>%
  # keep long structure so niche stays as a column
  group_by(year, site, niche) %>%
  summarise(
    proportion   = sum(proportion, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = niche, values_from = proportion) %>%
  mutate(across(3:10, ~ replace_na(., 0))) %>%
  rename_with(~ paste0("prop_", .x), 3:10) %>%
  mutate(year = as.numeric(year)) %>%
  left_join(hobo_temps, by = c("year", "site"))

#summarize counts
count_wide <- prop_df %>%
  select(-density_log) %>%
  # keep long structure so niche stays as a column
  group_by(year, site, niche) %>%
  summarise(
    count   = sum(count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = niche, values_from = count) %>%
  mutate(across(3:10, ~ replace_na(., 0))) %>%
  rename_with(~ paste0("count_", .x), 3:10) %>%
  mutate(year = as.numeric(year))

#combine counts and proportions
model_df <- left_join(prop_wide, count_wide, by = c("year", "site"))

#add mussels
model_df <- all_taxa %>%
  select(mussel_count, year, site) %>%
  left_join(model_df, ., by = c("year", "site"))

#add PDO and ENSO
pdo_annualized <- pdo %>%
  mutate(year = year(date)) %>%
  group_by(year) %>%
  summarise(pdo = mean(value))

model_df <- left_join(model_df, pdo_annualized, by = "year") %>%
  left_join(., enso, by = "year")

model_df <- model_df %>%
  mutate(blob = case_when(year < 2014 ~ "pre",
                          year >= 2014 ~ "post"),
         blob = factor(blob, , levels = c("pre", "post")),
         year_f = factor(year)) #for fitting random effects
#remove spaces in column names
names(model_df) <- janitor::make_clean_names(names(model_df))
model_df <- na.omit(model_df)
```

plot the proportions to look at trends
```{r}
model_df %>%
  select(year, site, prop_diving_fish, prop_diving_inverts) %>%
  pivot_longer(!c(year, site), names_to = "Niche", values_to = "proportion") %>%
  group_by(Niche, site) %>%
  mutate(proportion_scaled = as.numeric(scale(proportion))) %>%
  ggplot(aes(x = factor(year))) +
  geom_line(aes(y = proportion_scaled, group = Niche, color = Niche)) +
  facet_wrap(~site, ncol = 1)
```
```{r}
model_df %>%
  select(year, site, prop_diving_fish, prop_diving_inverts) %>%
  pivot_longer(!c(year, site), names_to = "Niche", values_to = "proportion") %>%
  group_by(Niche, site) %>%
  mutate(proportion_scaled = as.numeric(scale(proportion))) %>%
  ggplot(aes(x = factor(year))) +
  geom_line(aes(y = proportion, group = Niche, color = Niche)) +
  facet_wrap(~site, ncol = 1)
```

fit beta distributed models for the proportion of invert eating diving birds
using the proportions as response variables is good but explaining them with other proportions doesn't make sense because they are inherently linked
you can argue that getting rid of year makes sense because everything is annualized
need to work on nesting and offsets to make sure it is specified correctly

```{r fit beta models for proportions}
inverts.beta.1 <- glmmTMB(prop_diving_inverts ~ 1, data = model_df, family = beta_family(), REML = TRUE)
inverts.beta.2 <- glmmTMB(prop_diving_inverts ~ (1 | year_f), data = model_df, family = beta_family(), REML = TRUE)
inverts.beta.3 <- glmmTMB(prop_diving_inverts ~ (1 | site), data = model_df, family = beta_family(), REML = TRUE)
inverts.beta.4 <- glmmTMB(prop_diving_inverts ~ (1 | year_f) + (1 | site), data = model_df, family = beta_family(), REML = TRUE)

inverts.beta.5 <- glmmTMB(prop_diving_inverts ~ temp_scaled + (1 | year_f) + (1 | site), data = model_df, family = beta_family(), REML = TRUE)
inverts.beta.6 <- glmmTMB(prop_diving_inverts ~ mussel_count + (1 | year_f) + (1 | site), data = model_df, family = beta_family(), REML = TRUE)
inverts.beta.7 <- glmmTMB(prop_diving_inverts ~ temp_scaled + mussel_count + (1 | year_f) + (1 | site), data = model_df, family = beta_family(), REML = TRUE)

inverts.beta.8 <- glmmTMB(prop_diving_inverts ~ temp_scaled, data = model_df, family = beta_family(), REML = TRUE)
inverts.beta.9 <- glmmTMB(prop_diving_inverts ~ blob + (1 | site) + (1 | year_f), data = model_df, family = beta_family(), REML = TRUE)

inverts.beta.10 <- glmmTMB(prop_diving_inverts ~ blob + temp_scaled, data = model_df, family = beta_family(), REML = TRUE)
inverts.beta.11 <- glmmTMB(prop_diving_inverts ~ temp_scaled, data = model_df, family = beta_family(), REML = TRUE)
inverts.beta.12 <- glmmTMB(prop_diving_inverts ~ temp_scaled + blob, data = model_df, family = beta_family(), REML = TRUE)
inverts.beta.13 <- glmmTMB(prop_diving_inverts ~ blob*temp_scaled, data = model_df, family = beta_family(), REML = TRUE)
```

```{r beta model comparisson table}
list(
  inverts.beta.1,
  inverts.beta.2,
  inverts.beta.3,
  inverts.beta.4,
  inverts.beta.5,
  inverts.beta.6,
  inverts.beta.7,
  inverts.beta.8,
  inverts.beta.9,
  inverts.beta.10,
  inverts.beta.11,
  inverts.beta.12,
  inverts.beta.13
) %>%
  AICcmodavg::aictab()
```
```{r}
summary(inverts.beta.9)
```
Interpretation:
```{r}
intercept = fixef(inverts.beta.9)$cond["(Intercept)"]
blob = fixef(inverts.beta.9)$cond["blobpost"]

data.frame(blob = c("pre", "post"),
           estimate = c(plogis(intercept), 
                        plogis(intercept + blob)))
```
those are reported on the proportion scale so there is a 5% decrease in the proportion of birds that are diving invertebrate foragers

```{r}
plot(predict_response(inverts.beta.9, terms = "blob"))
```

Now model the proportion of diving piscivores
```{r fit beta models for proportions}
fish.beta.1 <- glmmTMB(prop_diving_fish ~ 1, data = model_df, family = beta_family(), REML = TRUE)
fish.beta.2 <- glmmTMB(prop_diving_fish ~ (1 | year_f), data = model_df, family = beta_family(), REML = TRUE)
fish.beta.3 <- glmmTMB(prop_diving_fish ~ (1 | site), data = model_df, family = beta_family(), REML = TRUE)
fish.beta.4 <- glmmTMB(prop_diving_fish ~ (1 | year_f) + (1 | site), data = model_df, family = beta_family(), REML = TRUE)

fish.beta.5 <- glmmTMB(prop_diving_fish ~ temp_scaled + (1 | year_f) + (1 | site), data = model_df, family = beta_family(), REML = TRUE)
fish.beta.6 <- glmmTMB(prop_diving_fish ~ mussel_count + (1 | year_f) + (1 | site), data = model_df, family = beta_family(), REML = TRUE)
fish.beta.7 <- glmmTMB(prop_diving_fish ~ temp_scaled + mussel_count + (1 | year_f) + (1 | site), data = model_df, family = beta_family(), REML = TRUE)

fish.beta.8 <- glmmTMB(prop_diving_fish ~ temp_scaled, data = model_df, family = beta_family(), REML = TRUE)
fish.beta.9 <- glmmTMB(prop_diving_fish ~ blob + (1 | site) + (1 | year_f), data = model_df, family = beta_family(), REML = TRUE)

fish.beta.10 <- glmmTMB(prop_diving_fish ~ blob + temp_scaled, data = model_df, family = beta_family(), REML = TRUE)
fish.beta.11 <- glmmTMB(prop_diving_fish ~ temp_scaled, data = model_df, family = beta_family(), REML = TRUE)
fish.beta.12 <- glmmTMB(prop_diving_fish ~ temp_scaled + blob, data = model_df, family = beta_family(), REML = TRUE)
fish.beta.13 <- glmmTMB(prop_diving_fish ~ blob*temp_scaled, data = model_df, family = beta_family(), REML = TRUE)

#remove random effects...
fish.beta.1 <- glmmTMB(prop_diving_fish ~ 1, data = model_df, family = beta_family(), REML = TRUE)
fish.beta.2 <- glmmTMB(prop_diving_fish ~ year, data = model_df, family = beta_family(), REML = TRUE)
fish.beta.3 <- glmmTMB(prop_diving_fish ~ site, data = model_df, family = beta_family(), REML = TRUE)
fish.beta.4 <- glmmTMB(prop_diving_fish ~ year + site, data = model_df, family = beta_family(), REML = TRUE)

fish.beta.5 <- glmmTMB(prop_diving_fish ~ temp_scaled + year + site, data = model_df, family = beta_family(), REML = TRUE)
fish.beta.6 <- glmmTMB(prop_diving_fish ~ mussel_count + year + site, data = model_df, family = beta_family(), REML = TRUE)
fish.beta.7 <- glmmTMB(prop_diving_fish ~ temp_scaled + mussel_count + year + site, data = model_df, family = beta_family(), REML = TRUE)

fish.beta.8 <- glmmTMB(prop_diving_fish ~ temp_scaled, data = model_df, family = beta_family(), REML = TRUE)
fish.beta.9 <- glmmTMB(prop_diving_fish ~ blob + site + year, data = model_df, family = beta_family(), REML = TRUE)

fish.beta.10 <- glmmTMB(prop_diving_fish ~ blob + temp_scaled, data = model_df, family = beta_family(), REML = TRUE)
fish.beta.11 <- glmmTMB(prop_diving_fish ~ temp_scaled, data = model_df, family = beta_family(), REML = TRUE)
fish.beta.12 <- glmmTMB(prop_diving_fish ~ temp_scaled + blob, data = model_df, family = beta_family(), REML = TRUE)
fish.beta.13 <- glmmTMB(prop_diving_fish ~ temp_scaled*year, data = model_df, family = beta_family(), REML = TRUE)
```

```{r beta model comparisson table}
list(
  fish.beta.1,
  fish.beta.2,
  fish.beta.3,
  fish.beta.4,
  fish.beta.5, 
  fish.beta.6,
  fish.beta.7, 
  fish.beta.8,
  fish.beta.9,
  fish.beta.10,
  fish.beta.11,
  fish.beta.12,
  fish.beta.13
) %>%
  AICcmodavg::aictab()
```

fit poisson distributed models for the counts of invert eating diving birds
offset of area surveyed
this accounts for differing effort but still allows us to use count data offset(log(area))
```{r fit poisson models for proportions}
fish.poisson.1 <- glmmTMB(count_diving_fish ~ 1, data = model_df, family = poisson(), REML = TRUE)
fish.poisson.2 <- glmmTMB(count_diving_fish ~ (1 | year), data = model_df, family = poisson(), REML = TRUE)
fish.poisson.3 <- glmmTMB(count_diving_fish ~ (1 | site), data = model_df, family = poisson(), REML = TRUE)
fish.poisson.4 <- glmmTMB(count_diving_fish ~ (1 | site) + (1 | year), data = model_df, family = poisson(), REML = TRUE)

fish.poisson.5 <- glmmTMB(count_diving_fish ~ mussel_count, data = model_df, family = poisson(), REML = TRUE)
fish.poisson.6 <- glmmTMB(count_diving_fish ~ temp_scaled, data = model_df, family = poisson(), REML = TRUE)
fish.poisson.7 <- glmmTMB(count_diving_fish ~ blob, data = model_df, family = poisson(), REML = TRUE)
fish.poisson.8 <- glmmTMB(count_diving_fish ~ mussel_count + (1 | year) + (1 | site), data = model_df, family = poisson(), REML = TRUE)
fish.poisson.9 <- glmmTMB(count_diving_fish ~ temp_scaled + (1 | year) + (1 | site), data = model_df, family = poisson(), REML = TRUE)
fish.poisson.10 <- glmmTMB(count_diving_fish ~ blob + (1 | year) + (1 | site), data = model_df, family = poisson(), REML = TRUE)

fish.poisson.11 <- glmmTMB(count_diving_fish ~ blob + temp_scaled + (1 | site) + (1 | year), data = model_df, family = poisson(), REML = TRUE)
fish.poisson.12 <- glmmTMB(count_diving_fish ~ blob + temp_scaled + mussel_count + (1 | site) + (1 | year), data = model_df, family = poisson(), REML = TRUE)
fish.poisson.13 <- glmmTMB(count_diving_fish ~ temp_scaled*year, data = model_df, family = poisson(), REML = TRUE)
```

```{r}

fish_poisson_models <- list(
  fish.poisson.1,
  fish.poisson.2,
  fish.poisson.3,
  fish.poisson.4,
  fish.poisson.5,
  fish.poisson.6,
  fish.poisson.7,
  fish.poisson.8,
  fish.poisson.9,
  fish.poisson.10,
  fish.poisson.11,
  fish.poisson.12,
  fish.poisson.13
)

AICcmodavg::aictab(fish_poisson_models)

```
```{r}
summary(fish.poisson.12)
```

```{r}
plot(ggpredict(fish.poisson.12, terms = "mussel_count")) /
  plot(ggpredict(fish.poisson.12, terms = "blob")) /
  plot(ggpredict(fish.poisson.12, terms = "temp_scaled"))
```
```{r}
plot(residuals(inverts_poisson.12, type = "pearson"))
```


```{r}
inverts_poisson.1  <- glmmTMB(count_diving_inverts ~ 1, data = model_df, family = poisson(), REML = TRUE)
inverts_poisson.2  <- glmmTMB(count_diving_inverts ~ (1 | year), data = model_df, family = poisson(), REML = TRUE)
inverts_poisson.3  <- glmmTMB(count_diving_inverts ~ (1 | site), data = model_df, family = poisson(), REML = TRUE)
inverts_poisson.4  <- glmmTMB(count_diving_inverts ~ (1 | site) + (1 | year), data = model_df, family = poisson(), REML = TRUE)
inverts_poisson.5  <- glmmTMB(count_diving_inverts ~ mussel_count, data = model_df, family = poisson(), REML = TRUE)
inverts_poisson.6  <- glmmTMB(count_diving_inverts ~ temp_scaled, data = model_df, family = poisson(), REML = TRUE)
inverts_poisson.7  <- glmmTMB(count_diving_inverts ~ blob, data = model_df, family = poisson(), REML = TRUE)
inverts_poisson.8  <- glmmTMB(count_diving_inverts ~ mussel_count + (1 | year) + (1 | site), data = model_df, family = poisson(), REML = TRUE)
inverts_poisson.9  <- glmmTMB(count_diving_inverts ~ temp_scaled + (1 | year) + (1 | site), data = model_df, family = poisson(), REML = TRUE)
inverts_poisson.10 <- glmmTMB(count_diving_inverts ~ blob + (1 | year) + (1 | site), data = model_df, family = poisson(), REML = TRUE)
inverts_poisson.11 <- glmmTMB(count_diving_inverts ~ blob + temp_scaled + (1 | site) + (1 | year), data = model_df, family = poisson(), REML = TRUE)
inverts_poisson.12 <- glmmTMB(count_diving_inverts ~ blob + temp_scaled + mussel_count + (1 | site) + (1 | year), data = model_df, family = poisson(), REML = TRUE)
inverts_poisson.13 <- glmmTMB(count_diving_inverts ~ temp_scaled * year, data = model_df, family = poisson(), REML = TRUE)
```

```{r}
list(
  inverts_poisson.1,
  inverts_poisson.2,
  inverts_poisson.3,
  inverts_poisson.4,
  inverts_poisson.5,
  inverts_poisson.6,
  inverts_poisson.7,
  inverts_poisson.8,
  inverts_poisson.9,
  inverts_poisson.10,
  inverts_poisson.11,
  inverts_poisson.12,
  inverts_poisson.13
) %>%
  AICcmodavg::aictab()

```

```{r}
summary(inverts_poisson.12)
```

```{r}
plot(ggpredict(inverts_poisson.12, terms = "mussel_count")) /
  plot(ggpredict(inverts_poisson.12, terms = "blob")) /
  plot(ggpredict(inverts_poisson.12, terms = "temp_scaled"))
```

next steps:
-figure out the random effects structures or lack theirof
-probably write down specific questions so that fitting the models actually makes sense
-make sure there aren't autocorrelation issues
-is there a better family than poison, need to check for overdisperson and maybe go quasibonomail
-try modeling temp as a smooth in case there is a stabilizing or non linear effeect!
-should year be a smooth? I don't think there is a obvious trend for most of these


is there a heatwave effect on benthivores
is there a heatwave effect on fishivores
