---
title: "R Notebook"
output: html_notebook
---

```{r warning=FALSE}
library(tidyverse)
library(janitor)
library(RColorBrewer)
library(ggeffects)
```

```{r read in data}
obs <- read.csv("data/KBAYKATMKEFJ_2006-2023_Summer_MBM_Observations.csv")
transects <- read.csv("data/KBAYKATMKEFJ_2006-2023_Summer_MBM_Transects.csv")
```

```{r make a single df}
#make year and site columns from the trip ID
obs <- obs %>%
  mutate(
    site = substr(TripID, 1, 4),
    year = substr(TripID, 8, 11)
  ) %>%
  filter()

species <- unique(obs$Detect_SpeciesCode)
sites <- unique(obs$site)
years <- unique(obs$year)

#make transects by site
transects <- transects  %>%
  mutate(
    site = substr(TripID, 1, 4),
    year = substr(TripID, 8, 11)
  )

#identify pelagic transects
pelagic_transects <- transects %>%
  filter(TransType == "PELAGIC") %>%
  pull(TXNumber) %>%
  unique()

#now remove pelagic transects from bird obs
obs <- obs %>% filter(!TXNumber %in% pelagic_transects)
```

link obs with transect effort

```{r}
#sum species total for each survey (year and site)
obs_summary <-obs %>%
  group_by(year, site, Detect_SpeciesCode) %>%
  summarise(count = sum(Count)) %>%
  rename(species = Detect_SpeciesCode)

#find all unique combinations of year and site then repeat the bird list for each of them
sampling_events <- obs %>%
  distinct(year, site)

species_df <- species %>% 
  data.frame() %>%
  rename(species = ".")

species_list <- list()

for(i in 1:nrow(sampling_events)) {
  event <- sampling_events[i,]
  species_df$year <- event$year
  species_df$site <- event$site
  
  species_list[[i]] <- species_df
}

#convert list to dataframe
sampling_events <- bind_rows(species_list)

#this adds in zeros for the species not observed on each survey but doesnt make up surveys that didn't happen (as opposed to expand_grid)
obs_summary <- left_join(sampling_events, obs_summary) %>%
  mutate(count = replace_na(count, 0))

#this adds up all the area and distance sampled in each year
transects_summary <- transects %>% 
  group_by(year, site) %>%
  summarise(length = sum(Length_km),
            area = sum(Area_Sampled_km2))

#add the effort to the counts and calculate density
bird_densities <- left_join(obs_summary, transects_summary, by = c("site", "year")) %>%
  mutate(density = count/area) %>%
  relocate(species, .after = site)
```

find the most common birds to look at

```{r}
birds <- obs %>%
  filter(!Detect_SpeciesCode %in% c("UNCO")) %>%
  group_by(Detect_SpeciesCode) %>%
  summarise(count = sum(Count)) %>%
  arrange(desc(count)) %>%
  head(20) %>% #change this number for more or less species
  pull(Detect_SpeciesCode)
```

```{r}
bird_densities %>%
  filter(species %in% birds) %>%
  #mutate(density = scale(density)) %>%
  ggplot() +
  geom_path(aes(year, density, colour = site, group = site)) +
  geom_point(aes(year, density, colour = site, group = site)) +
  facet_wrap(~species, scales = "free_y") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Bird Density",
       subtitle = "20 most abundant species")
```

```{r}
bird_densities %>%
  filter(species %in% birds) %>%
  group_by(species) %>%
  mutate(density = scale(density)) %>%
  ungroup() %>%
  ggplot() +
  geom_path(aes(year, density, colour = species, group = species)) +
  geom_point(aes(year, density, colour = species, group = species)) +
  facet_wrap(~site, scales = "free_y", nrow = 3) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "scaled bird abundances")
```

```{r}
bird_densities %>%
  filter(species == "BLOY") %>%
  #mutate(density = scale(density)) %>%
  ggplot() +
  geom_path(aes(year, density, colour = site, group = site)) +
  facet_wrap(~species, scales = "free_y") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
mussel_counts <- read.csv("data/KATMKEFJWPWS_2008-2025_Mytilus_20mm_Count.csv")
mussel_size <- read.csv("data/KATMKEFJWPWS_2008-2025_Mytilus_20mm_Size.csv")
seastar_counts <- read.csv("data/KATMKEFJWPWS_2006-2025_Sea_Star_Count.csv")
```

```{r}
mussel_counts %>%
  filter(!BlockName %in% c("EPWS", "NPWS")) %>%
  mutate(Count = as.numeric(Count)) %>%
  group_by(BlockName, SiteName, YearSample) %>%
  summarise(avg = mean(Count)) %>%
  ggplot() +
  geom_point(aes(YearSample, avg, group = SiteName, colour = SiteName)) +
  facet_wrap(~BlockName, nrow = 3) +
  labs(title = "Mussel Counts")
```

```{r}
mussel_size %>%
  filter(!BlockName %in% c("EPWS", "NPWS")) %>%
  mutate(Size = as.numeric(Size)) %>%
  group_by(BlockName, SiteName, YearSample) %>%
  summarise(avg = mean(Size)) %>%
  ggplot() +
  geom_point(aes(YearSample, avg, group = SiteName, colour = SiteName)) +
  facet_wrap(~BlockName, nrow = 3) +
  labs(title = "Mussel Size")
```

```{r}
mussel_counts %>%
  filter(!BlockName %in% c("EPWS", "NPWS")) %>%
  mutate(Count = as.numeric(Count)) %>%
  group_by(BlockName, SiteName, YearSample) %>%
  summarise(avg = mean(Count)) %>%
  ggplot() +
  geom_boxplot(aes(YearSample, avg, group = YearSample)) +
  facet_wrap(~BlockName, nrow = 3, scales = "free_y") +
  labs(title = "Mussel Counts",
       y = "Count")
```

```{r}
mussel_size %>%
  filter(!BlockName %in% c("EPWS", "NPWS")) %>%
  mutate(Size = as.numeric(Size)) %>%
  group_by(BlockName, SiteName, YearSample) %>%
  summarise(avg = mean(Size)) %>%
  ggplot() +
  geom_boxplot(aes(YearSample, avg, group = YearSample)) +
  facet_wrap(~BlockName, nrow = 3, scales = "free_y") +
  labs(title = "Mussel Size",
       y = "Size (mm)")
```

noww check out the seastars

```{r}
seastar_counts %>%
  mutate(site = substr(SiteID, 1, 3)) %>%
  group_by(Year, site, Species) %>%
  summarise(avg_density = mean(Density_individual200SqM)) %>%
  group_by(Species) %>%
  mutate(avg_density = as.numeric(scale(avg_density))) %>%
  ungroup() %>%
  ggplot() +
  geom_path(aes(Year, avg_density, color = Species)) +
  facet_wrap(~site, nrow = 3, scales = "free_y") +
  labs(title = "Seastar density (scaled)")
```

```{r}
seastar_counts %>%
  mutate(site = substr(SiteID, 1, 3)) %>%
  group_by(Year, site, Species) %>%
  summarise(avg_density = mean(Density_individual200SqM)) %>%
  ggplot() +
  geom_smooth(aes(Year, avg_density, group = site, color = site)) +
  facet_wrap(~Species, nrow = 3, scales = "free_y") +
  labs(title = "Seastar Density")
```

```{r}
bird_densities %>%
  group_by(year, site) %>%
  summarise(avg_density = mean(density)) %>%
  ggplot() +
  geom_path(aes(year, avg_density, color = site, group = site)) +
  labs(title = "all birds combined",
       subtitle = "is this stupid?")
```

combine everything into a single data frame for plotting

```{r}
seastar_wide <- seastar_counts %>%
  mutate(site = substr(SiteID, 1, 3),
         site = case_when(site == "AKP" ~ "KATM",
                          site == "PWS" ~ "WPWS",
                          site == "KEP" ~ "KEFJ")) %>%
  group_by(Year, site, Species) %>%
  summarise(avg_density = mean(Density_individual200SqM)) %>%
  pivot_wider(names_from = Species, values_from = avg_density) %>%
  rename(year = Year)

mussels_wide <- mussel_counts %>%
  rename(site = BlockName,
         year = YearSample) %>%
  filter(!site %in% c("EPWS", "NPWS")) %>%
  mutate(Count = as.numeric(Count)) %>%
  group_by(site, year) %>%
  summarise(avg_count = mean(Count, na.rm = T)) # check that this is ok and isnt getting rid of zeros

birds_wide <- bird_densities %>%
  select(!c(length, area, count)) %>%
  pivot_wider(names_from = species, values_from = density) %>%
  mutate(year = as.numeric(year))

all_taxa <- left_join(birds_wide, seastar_wide, by = c("year", "site")) %>%
  left_join(., mussels_wide, by = c("year", "site")) %>%
  rename(mussel_count = avg_count) %>%
  mutate(
    across(
      c(seq(3, ncol(.))),   
      ~ scale(.)[,1]
    )
  ) %>%
  rename_with(janitor::make_clean_names, .cols = 128:135)
```

```{r}
all_taxa %>%
  filter(site != "KBAY") %>%
  select(year, site, mussel_count, BLOY, HADU, pycnopodia_helianthoides) %>%
  pivot_longer(cols = !c(year, site), names_to = "species", values_to = "Abundance") %>%
  ggplot() +
  geom_path(aes(year, Abundance, color = species)) +
  geom_point(aes(year, Abundance, color = species)) +
  geom_hline(aes( yintercept = 0)) +
  facet_wrap(~site, nrow = 3)
```


```{r}
bloy_nest <- read.csv("data/KBAYKATMKEFJWPWS_2006-2025_blackOystercatcher_Transect_Summary.csv")

bloy_nest %>%
  group_by(Year, BlockName) %>%
  summarise(avg_density = mean(as.numeric(NestDensity))) %>%
  ggplot() +
  geom_path(aes(Year, avg_density, color = BlockName)) +
  labs(title = "BLOY Nest Density")
```

```{r}
isotopes <- read.csv("data/Isotope_data_all_regions_2014-2025.csv", fileEncoding='latin1')
isotopes <- isotopes %>% select(!c(X, X.1, X.2))

```

```{r}
isotopes %>%
  mutate(Year = as.factor(Year)) %>%
  ggplot() +
  geom_jitter(aes(d15N...., d13C...., colour = Year), alpha = 0.4)
  facet_wrap(~Year)
```

```{r}
isotopes %>%
  mutate(Year = as.factor(Year)) %>%
  ggplot() +
  geom_jitter(aes(d15N...., d13C...., colour = Species), alpha = 0.7) +
  facet_wrap(~Year)
```

```{r}
isotopes %>%
  filter(Species == "trossulus") %>%
  mutate(Year = as.factor(Year)) %>%
  ggplot() +
  geom_jitter(aes(d15N...., d13C...., colour = Year), alpha = 0.9, size = 3) +
    scale_x_continuous(limits = c(5, 10))
```

```{r}
trossulus <-isotopes %>%
  filter(Species == "trossulus") %>%
  filter(!is.na(d15N....),
         !is.na(d13C....)) %>%
#mutate(Year = as.factor(Year)) %>%
 rename_with(janitor::make_clean_names)
 
hulls <- trossulus %>%
  group_by(year) %>%
  slice(chull(d15n, d13c)) %>%
  ungroup()


ggplot(trossulus) +
  geom_point(aes(x = d15n, y = d13c),
              alpha = 0.1, size = 3) +
  geom_polygon(data = hulls,
               aes(x = d15n, y = d13c, group = year),
               alpha = 0.2, colour = "black") +
  #scale_x_continuous(limits = c(5, 10)) +
  facet_wrap(~year)
```

```{r}
trossulus %>%
  ggplot() +
  geom_jitter(aes(d15n, d13c, colour = region), alpha = 0.5, size = 3) +
  facet_wrap(~year) +
  labs(title = "mussel isotopes")
```

```{r}
isotopes %>%
  filter(Sample.ID == "POM") %>%
  ggplot() +
  geom_jitter(aes(d15N...., d13C...., colour = Region), alpha = 0.5, size = 3) +
  facet_wrap(~Year) +
  labs(title = "POM")
```

Limpet time!

```{r}
limpet_count <- read.csv("data/KATMKEFJWPWS_2010-2025_Limpet_Count.csv")
limpet_size <- read.csv("data/KATMKEFJWPWS_2006-2025_Limpet_Size.csv")
```

```{r}
limpet_count %>%
  mutate(site = substr(SiteID, 1, 3),
         site = case_when(site == "AKP" ~ "KATM",
                          site == "PWS" ~ "WPWS",
                          site == "KEP" ~ "KEFJ"),
         Density = as.numeric(Density)) %>%
  group_by(Year, site) %>%
  summarise(Density = mean(Density, na.rm = T)) %>%
  mutate(Density = scale(Density)[,1]) %>%
  ggplot() +
  geom_path(aes(Year, Density, color = site)) +
  geom_point(aes(Year, Density, color = site)) +
  labs(title = "limpet density")
```

The limpet abundances between KATM and KEFJ seem to trend oppositely

```{r}
limpet_size %>%
  mutate(site = substr(SiteID, 1, 3),
         site = case_when(site == "AKP" ~ "KATM",
                          site == "PWS" ~ "WPWS",
                          site == "KEP" ~ "KEFJ"),
         Limpet_Size_mm = as.numeric(Limpet_Size_mm)) %>%
  group_by(Year, site) %>%
  summarise(size_avg = mean(Limpet_Size_mm, na.rm = T)) %>%
  #mutate(size_avg = scale(size_avg)[,1]) %>%
  ggplot() +
  geom_path(aes(Year, size_avg, color = site)) +
  geom_point(aes(Year, size_avg, color = site)) +
  labs(title = "limpet size")
```

but the size fluctuations seem to trend similarly between sites

```{r}
limpet_size %>%
  mutate(site = substr(SiteID, 1, 3),
         site = case_when(site == "AKP" ~ "KATM",
                          site == "PWS" ~ "WPWS",
                          site == "KEP" ~ "KEFJ"),
         Limpet_Size_mm = as.numeric(Limpet_Size_mm)) %>%
  group_by(Year, site, SiteName) %>%
  summarise(size_avg = mean(Limpet_Size_mm, na.rm = T)) %>%
  #mutate(size_avg = scale(size_avg)[,1]) %>%
  ggplot() +
  geom_path(aes(Year, size_avg, color = SiteName)) +
  geom_point(aes(Year, size_avg, color = SiteName)) +
  facet_wrap(~site, ncol = 1) +
  labs(title = "limpet size by site") +
  theme(legend.position = "none")
```

snails:

```{r}
snails <- read.csv("data/KATMKEFJWPWS_2006-2025_Nucella_Katharina_Lirabuccinum_Count.csv")
```

```{r}
snails %>%
  mutate(site = substr(SiteID, 1, 3),
         site = case_when(site == "AKP" ~ "KATM",
                          site == "PWS" ~ "WPWS",
                          site == "KEP" ~ "KEFJ")) %>%
  group_by(Year, Elevation_Position, site, SpeciesName) %>%
  summarise(count = mean(Count)) %>%
  ggplot() +
  geom_path(aes(Year, count, color = SpeciesName)) +
  geom_point(aes(Year, count, color = SpeciesName)) +
  facet_grid(site~Elevation_Position) +
  labs(title = "snail abundance")
```

check out the temperature data

```{r message=FALSE}
temperature <- list.files("data/temperature",
                       pattern = "\\.csv$",
                       full.names = TRUE) %>%
  map_dfr(read_csv)

#make month and year columns
temperature$year <- year(temperature$date)
temperature$month <- month(temperature$date)
```

```{r}
temperature %>%
  filter(exposure == "water",
         tide_height == 0.5) %>%
  ggplot() +
  geom_path(aes(date, temperature, color = site)) 
```

```{r}
temperature %>%
  filter(exposure == "water",
         month == 7,
         tide_height == 0.5) %>%
  ggplot() +
  geom_boxplot(aes(year, temperature, group = year, color = block)) +
  facet_wrap(~block, ncol =1)
```

```{r}
temperature %>%
  filter(exposure == "water",
         month == 5,
         tide_height == 0.5) %>%
  group_by(year, block) %>%
  summarise(avg_temp = mean(temperature)) %>%
  ggplot() +
  geom_path(aes(year, avg_temp, color = block)) +
  geom_point(aes(year, avg_temp, color = block)) +
  labs(title = "May Temp")
```

```{r}
temperature %>%
  filter(exposure == "water",
         month == 6,
         tide_height == 0.5) %>%
  group_by(year, block) %>%
  summarise(avg_temp = mean(temperature)) %>%
  ggplot() +
  geom_path(aes(year, avg_temp, color = block)) +
  geom_point(aes(year, avg_temp, color = block)) +
  labs(title = "June Temp")
```

```{r}
temperature %>%
  filter(exposure == "water",
         month == 7,
         tide_height == 0.5) %>%
  group_by(year, block) %>%
  summarise(avg_temp = mean(temperature)) %>%
  ggplot() +
  geom_path(aes(year, avg_temp, color = block)) +
  geom_point(aes(year, avg_temp, color = block)) +
  labs(title = "July Temp")
```

```{r}

temperature %>%
  filter(exposure == "water",
         month %in% c(4,5,6,7,8,9),
         tide_height == 0.5) %>%
  mutate(month_name = factor(month.name[month],
                             levels = month.name[4:9])) %>%
  group_by(year, block, month_name) %>%
  summarise(avg_temp = mean(temperature), .groups = "drop") %>%
  ggplot() +
  geom_path(aes(year, avg_temp, color = block)) +
  geom_point(aes(year, avg_temp, color = block)) +
  facet_wrap(~month_name)
```

```{r}
temperature %>%
  filter(exposure == "water",
         month %in% c(10,11,12,1,2,3),
         tide_height == 0.5) %>%
  mutate(month_name = factor(month.name[month],
                             levels = c("October","November","December",
                                        "January","February","March"))) %>%
  group_by(year, block, month_name) %>%
  summarise(avg_temp = mean(temperature), .groups = "drop") %>%
  ggplot() +
  geom_path(aes(year, avg_temp, color = block)) +
  geom_point(aes(year, avg_temp, color = block)) +
  facet_wrap(~month_name)
```

```{r}
winter_df <- temperature %>%
  filter(exposure == "water",
         month %in% c(10,11,12,1,2,3),
         tide_height == 0.5) %>%
  mutate(month_name = factor(month.name[month],
                             levels = c("October","November","December",
                                        "January","February","March")))

# Step 2: Compute climatology (long-term mean for each month)
climatology <- winter_df %>%
  group_by(month_name, block) %>%
  summarise(clim_mean = mean(temperature), .groups="drop")

# Step 3: Compute avg_temp by group and join climatology
winter_anoms <- winter_df %>%
  group_by(year, block, month_name) %>%
  summarise(avg_temp = mean(temperature), .groups="drop") %>%
  left_join(climatology, by="month_name") %>%
  mutate(anomaly = avg_temp - clim_mean)

# Step 4: Plot anomalies
ggplot(winter_anoms) +
  geom_col(aes(year, anomaly, fill = block.y), position = position_dodge()) +
  facet_wrap(~month_name) +
  labs(y = "Temperature anomaly (°C)")
```

```{r}
obs %>%
  filter(Detect_SpeciesCode == "SEOT") %>%
  pull(Comment) %>%
  write.csv("C:/Users/bhiggins/OneDrive - DOI/Desktop/otters.csv")

otters <- obs %>%
  filter(Detect_SpeciesCode == "SEOT") %>%
  mutate(
    x_lower = str_to_lower(Comment),

    # Define what counts as a "pup" indicator, including your new variants
    has_pup = str_detect(
      x_lower,
      paste0(
        "(",
        "pup",           # 'pup' substring (covers 'pup' and 'pups')
        "|\\bpups\\b",   # explicit plural
        "|\\bp\\b",      # bare 'p' as a whole word (e.g., '1=p', '1 = p')
        "|large\\s*p\\b",# 'large p' (with optional spaces)
        "|large\\s*up\\b",# 'large up' (with optional spaces)
        "|largep",       # 'largep' without space
        "|largeup",      # 'largeup' without space
        ")"
      )
    ),

    any_num = str_detect(x_lower, "\\d+"),

    pups = case_when(
      has_pup & any_num ~ as.numeric(str_extract(x_lower, "\\d+")), # grab leading number
      has_pup & !any_num ~ 1,                                        # 'pup' without number => 1
      TRUE ~ NA_real_
    )
  ) %>%
  select(-x_lower, -has_pup, -any_num) %>%
  mutate(pups = replace_na(pups, 0),
         adults = Count - pups)

#check
which((otters$pups + otters$adults) != otters$Count) #good to go!

```

```{r}
otter_summary <- otters %>%
  group_by(year, site) %>%
  summarise(adults = sum(adults),
            pups = sum(pups),
            ratio = pups/adults) %>%
  mutate(year = as.numeric(year))
```

```{r}
otter_summary %>%
  ggplot() +
  geom_path(aes(year, ratio, color = site)) +
  geom_point(aes(year, ratio, color = site)) +
  labs(title = "SEOT pups/adult ratio")
```

```{r}
bird_densities %>%
  mutate(year = as.numeric(year)) %>%
  filter(species == "SEOT") %>%
  ggplot() +
  geom_path(aes(year, density, color = site)) +
  geom_point(aes(year, density, color = site)) +
  labs(title = "SEOT adult counts")
```

```{r}
temp_df <- temperature %>%
  filter(exposure == "water") %>%
  group_by(year, block) %>%
  summarise(avg_temp = mean(temperature), .groups = "drop")

left_join(otter_summary, temp_df, by = c("year", "site" = "block")) %>%
  filter(site != "KBAY") %>%
  ungroup() %>%
  mutate(
    ratio2 = as.numeric(scale(ratio)),
    temp   = as.numeric(scale(avg_temp))) %>%
  ggplot() +
  geom_path(aes(year, ratio2, color = site)) +
  geom_point(aes(year, ratio2, color = site)) +
  geom_path(aes(year, temp, color = "temp")) +
  facet_wrap(~site, ncol = 1) +
  labs(title = "SEOT pups/adult ratio")
  
```

```{r}
all_taxa %>%
  filter(site != "KBAY") %>%
  select(year, site, mussel_count, BLOY, SEOT, pycnopodia_helianthoides) %>%
  pivot_longer(cols = !c(year, site), names_to = "species", values_to = "Abundance") %>%
  ggplot() +
  geom_path(aes(year, Abundance, color = species)) +
  geom_point(aes(year, Abundance, color = species)) +
  geom_hline(aes( yintercept = 0)) +
  facet_wrap(~site, nrow = 3)
```

```{r}
birds_most_detected <- obs %>%
  filter(!Detect_SpeciesCode %in% c("UNCO")) %>%
  group_by(Detect_SpeciesCode) %>%
  summarise(count = sum(Count),
            detections = n()) %>%
  arrange(desc(detections)) %>%
  head(20) %>% #change this number for more or less species
  pull(Detect_SpeciesCode)

#find the birds we dont want
trash <- obs %>%
  filter(!Detect_SpeciesCode %in% birds_most_detected) %>%
  pull(Detect_SpeciesCode) %>%
  unique()

#add GAK1 temp
#https://research.cfos.uaf.edu/gak1/
gak1 <- read_table(
  "https://research.cfos.uaf.edu/gak1/data/GAK1_Compilation_1970-2025/GAK1_Compilation_1970-2025.dat"
)

gak1$`Date_Time,` <- ymd_hms(gak1$`Date_Time,`)
gak1$year <- year(gak1$`Date_Time,`)

gak_Summary <- gak1 %>%
  mutate(temp = str_replace(`Temperature_[C],`, ",", "")) %>%
  mutate(temp = as.numeric(temp),
         depth = as.numeric(`Pressure_[dbar],`)) %>%
  filter(depth < 10) %>%
  group_by(year) %>%
  summarise(temp = mean(as.numeric(temp), na.rm = T)) %>%
  mutate(temp_scaled = as.numeric(scale(temp)))
  


library(ggcorrplot)
cor_mat <- all_taxa %>% 
   filter(site == "KEFJ") %>%
  select(!site) %>%
  select(-any_of(trash)) %>%
  select(!LEAU) %>%
  select(!UNJA) %>%
  left_join(., gak_Summary, by = "year") %>%
  slice(2:16) %>%
  cor()

ggcorrplot(cor_mat, type = "lower")


hc <- hclust(dist(cor_mat))

order_vec <- hc$order

ggcorrplot(cor_mat[order_vec, order_vec], type = "lower") +
  labs(title ="KEFJ")

```

```{r}
cor_katm <- all_taxa %>% 
   filter(site == "KATM") %>%
  filter(year != 2007) %>%
  select(!site) %>%
  select(-any_of(trash)) %>%
  select(!LEAU) %>%
  select(!UNJA) %>%
  select(!`mediaster_aequalis`) %>%
  left_join(., gak_Summary, by = "year") %>%
  left_join(., otter_summary[otter_summary$site == "KATM",], by = "year") %>%
  select(!site) %>%
  cor(use = "complete.obs")

hc_katm<- hclust(dist(cor_katm))

order_vec_katm <- hc_katm$order

ggcorrplot(cor_katm[order_vec_katm, order_vec_katm], type = "lower") +
  labs(title ="KATM")

```

```{r}
all_taxa %>%
  filter(site != "KBAY") %>%
  select(year, site, RFCO, SEOT) %>%
    left_join(., gak_Summary, by = "year") %>%
  left_join(., otter_summary %>%
              mutate( adults = as.numeric(scale(adults)),
                      pups = as.numeric(scale(pups))), by = c("site", "year")) %>%
  select(!temp) %>%
  pivot_longer(cols = !c(year, site), names_to = "species", values_to = "Abundance") %>%
  filter(species != "adults", species != "pups") %>%
  ggplot() +
  geom_path(aes(year, Abundance, color = species)) +
  geom_point(aes(year, Abundance, color = species)) +
  geom_hline(aes( yintercept = 0)) +
  facet_wrap(~site, nrow = 3)

```

```{r}
all_taxa %>%
  filter(site != "KBAY") %>%
  select(year, site, RFCO, SEOT) %>%
    left_join(., gak_Summary, by = "year") %>%
  left_join(., otter_summary, by = c("site", "year"))
```

```{r}
all_taxa %>%
   filter(site != "KBAY") %>%
   select(year, site, RFCO, SEOT) %>%
     left_join(., gak_Summary, by = "year") %>%
   left_join(., otter_summary %>%
               mutate( adults = as.numeric(scale(adults)),
                       pups = as.numeric(scale(pups))), by = c("site", "year")) %>%
   select(!temp) %>%
   pivot_longer(cols = !c(year, site), names_to = "species", values_to = "Abundance") %>%
   filter(species != "adults", species != "pups") %>%
   ggplot() +
   geom_path(aes(year, Abundance, color = species)) +
   geom_point(aes(year, Abundance, color = species)) +
   geom_hline(aes( yintercept = 0)) +
   facet_wrap(~site, nrow = 3)
```

```{r}
p.mat.katm <- all_taxa %>% 
   filter(site == "KATM") %>%
  filter(year != 2007) %>%
  select(!site) %>%
  select(-any_of(trash)) %>%
  select(!LEAU) %>%
  select(!UNJA) %>%
  select(!`mediaster_aequalis`) %>%
  left_join(., gak_Summary, by = "year") %>%
  left_join(., otter_summary[otter_summary$site == "KATM",], by = "year") %>%
  select(!c(temp_scaled, site, adults, pups)) %>%
  cor_pmat()

cor_katm <- all_taxa %>% 
   filter(site == "KATM") %>%
  filter(year != 2007) %>%
  select(!site) %>%
  select(-any_of(trash)) %>%
  select(!LEAU) %>%
  select(!UNJA) %>%
  select(!`mediaster_aequalis`) %>%
  left_join(., gak_Summary, by = "year") %>%
  left_join(., otter_summary[otter_summary$site == "KATM",], by = "year") %>%
  select(!c(temp_scaled, site, adults, pups)) %>%
  cor(use = "complete.obs")

hc_katm<- hclust(dist(cor_katm))

order_vec_katm <- hc_katm$order

ggcorrplot(cor_katm, p.mat = p.mat.katm, hc.order = F, insig = "stars") +
  labs(title ="KATM")
  ggsave("C:/Users/bhiggins/OneDrive - DOI/Desktop/KATM_data_dredging.png", height = 10, width = 10)

```

```{r}
p.mat.kefj <- all_taxa %>% 
   filter(site == "KEFJ") %>%
  filter(year != 2007) %>%
  select(!site) %>%
  select(-any_of(trash)) %>%
  select(!LEAU) %>%
  select(!UNJA) %>%
  select(!`mediaster_aequalis`) %>%
  left_join(., gak_Summary, by = "year") %>%
  left_join(., otter_summary[otter_summary$site == "KEFJ",], by = "year") %>%
  mutate(seot_pup_ratio = ratio) %>%
  select(!c(temp_scaled, site, adults, pups, ratio)) %>%
  cor_pmat()

cor_mat <- all_taxa %>% 
   filter(site == "KEFJ") %>%
  filter(year != 2007) %>%
  select(!site) %>%
  select(-any_of(trash)) %>%
  select(!LEAU) %>%
  select(!UNJA) %>%
  select(!`mediaster_aequalis`) %>%
  left_join(., gak_Summary, by = "year") %>%
  left_join(., otter_summary[otter_summary$site == "KEFJ",], by = "year") %>%
    mutate(seot_pup_ratio = ratio) %>%
  select(!c(temp_scaled, site, adults, pups, ratio)) %>%
  cor(use = "complete.obs")

#hc <- hclust(dist(cor_mat))

#order_vec <- hc$order

ggcorrplot(cor_mat, p.mat = p.mat.kefj,  insig = "stars", hc.order = F) +
  labs(title ="KEFJ")
 ggsave("C:/Users/bhiggins/OneDrive - DOI/Desktop/KEFJ_data_dredging.png", height = 10, width = 10)
```

keep making it more of a menace. add the snails!

```{r}
snails_upper <- snails %>%
  mutate(site = substr(SiteID, 1, 3),
         site = case_when(site == "AKP" ~ "KATM",
                          site == "PWS" ~ "WPWS",
                          site == "KEP" ~ "KEFJ")) %>%
  group_by(Year, Elevation_Position, site, SpeciesName) %>%
  summarise(count = mean(Count)) %>%
  pivot_wider(names_from = SpeciesName, values_from = count) %>%
  ungroup() %>%
    filter(Elevation_Position == "Upper (1.5 m MLLW)") %>%
  select(-Elevation_Position)

all_taxa <- left_join(all_taxa, snails_upper, by = c("year" = "Year", "site"))
```

thinking out loud. all the heatwave analyses use the heatwave as a factor (pre, during, post), but looking at the temperatures, there sort of isnt a post, or at least its less straight forward. try using temp as a continious variable?

```{r}
gak_Summary %>%
  filter(year > 2006) %>%
  mutate(lta = mean(temp_scaled)) %>%
  ggplot() +
  geom_col(aes(year, temp_scaled))
```

```{r}
nppsd <- read.csv("C:/Users/bhiggins/OneDrive - DOI/MBM_Data_Proofing/SpeciesTaxonomyNPPSDv4.csv")

obs %>%
  filter(site == "KATM") %>%
  filter(!Detect_SpeciesCode %in% c("UNCO")) %>%
  group_by(Detect_SpeciesCode) %>%
  summarise(count = n()) %>%
  arrange(desc(count)) %>%
  left_join(., nppsd, by = c("Detect_SpeciesCode" = "species_code")) %>%
  select(Detect_SpeciesCode, common_name, count) %>%
  rename(number_of_detections= count) %>%
  write.csv(., "C:/Users/bhiggins/OneDrive - DOI/Desktop/KATM_bird_species.csv")
```

thinking more: I think we should be able to bootstrap density estimates for each year, this would give us confidence intervals so we would know more about our ability to detect change

first step is to make a data frame with the transects

```{r}
obs_boot <- obs %>%
  group_by(year, site, TXNumber, Detect_SpeciesCode) %>%
  summarise(count = sum(Count)) %>%
  rename(species = Detect_SpeciesCode)

#find all unique combinations of year, site, and transect, then repeat the bird list for each of them
sampling_events_boot <- obs %>%
  distinct(year, site, TXNumber)

species_df <- species %>% 
  data.frame() %>%
  rename(species = ".")

species_list <- list()

for(i in 1:nrow(sampling_events_boot)) {
  event <- sampling_events_boot[i,]
  species_df$year <- event$year
  species_df$site <- event$site
  species_df$transect <- event$TXNumber
  
  species_list[[i]] <- species_df
}

#convert list to dataframe
sampling_events_boot <- bind_rows(species_list)

#this adds in zeros for the species not observed on each survey but doesnt make up surveys that didn't happen (as opposed to expand_grid)
obs_boot <- left_join(sampling_events_boot, obs_boot, by = c("species", "year", "site", "transect" = "TXNumber")) %>%
  mutate(count = replace_na(count, 0))

#add the effort to the counts and calculate density
bird_densities_by_transect <- left_join(obs_boot, transects, by = c("site", "year", "transect" = "TXNumber")) %>%
  select(year, site, species, transect, Length_km, Area_Sampled_km2, count) %>%
  mutate(count = as.numeric(count),
         area = as.numeric(Area_Sampled_km2),
         density = count/area,
         year = as.numeric(year))
```

```{r}
library(boot)
set.seed(47)
#create the mean function for bootstrapping
mean_stat <- function(d, i) mean(d[i], na.rm = T)

#identify each survey
distinct_samples <- sampling_events_boot %>%
 distinct(year, site) %>%
 mutate(year = as.numeric(year))

#create an object to store results in
bootstrap_estimates <- list()

#loop through each survey and bootstrap density estimates
for(i in 1:37) {
  year = as.numeric(distinct_samples$year[i])
  site = distinct_samples$site[i]
  
  df = bird_densities_by_transect %>%
          filter(site == .env$site,
               year == .env$year,
               species == "HADU")

  boot_out <- boot(data = df$density, statistic = mean_stat, R = 2000)

  ci <- boot.ci(boot_out, type = c("perc", "bca"))
  
  estimates <- data.frame(
    year = year,
    site = site,
    est  = ci$t0,
    lci  = ci$bca[4],
    uci  = ci$bca[5]
 )
  
  bootstrap_estimates[[i]] <- estimates
}

bootstrap_estimates <- bind_rows(bootstrap_estimates)
```

the graph below is funky becasue it includes estimates generated using both parks. just filtering to loop through for a single park breaks it becasue not all parks exist in all years. fix tomorow

```{r}
bootstrap_estimates %>%
  ggplot() +
  geom_ribbon(aes(x = year, ymin = lci, ymax = uci), fill = "blue", alpha = 0.2) +
  geom_path(aes(x = year, y = est)) +
  geom_point(aes(x = year, y = est)) +
  facet_wrap(~ site, ncol = 1, scales = "free_y")
```

Ok so these aren't really any good for detecting trends. there isnt any good reason that the transects should have low variance because they aren't the same thing. you could do it by habitat category but that also might miss the point

```{r}
bootstrap_estimates <- list()

#loop through each survey and bootstrap density estimates
for(i in 1:37) {
  year = as.numeric(distinct_samples$year[i])
  site = distinct_samples$site[i]
  
  df = bird_densities_by_transect %>%
          filter(site == .env$site,
               year == .env$year,
               species == "SEOT")

  boot_out <- boot(data = df$density, statistic = mean_stat, R = 2000)

  ci <- boot.ci(boot_out, type = c("perc", "bca"))
  
  estimates <- data.frame(
    year = year,
    site = site,
    est  = ci$t0,
    lci  = ci$bca[4],
    uci  = ci$bca[5]
 )
  
  bootstrap_estimates[[i]] <- estimates
}

bootstrap_estimates <- bind_rows(bootstrap_estimates)

bootstrap_estimates %>%
  ggplot() +
  geom_ribbon(aes(x = year, ymin = lci, ymax = uci), fill = "blue", alpha = 0.2) +
  geom_path(aes(x = year, y = est)) +
  geom_point(aes(x = year, y = est)) +
  facet_wrap(~ site, ncol = 1, scales = "free_y")
```

```{r}
bootstrap_estimates <- list()

common_species = obs %>%
  filter(!Detect_SpeciesCode %in% c("UNCO", "BRMU")) %>%
  group_by(Detect_SpeciesCode) %>%
  summarise(count = n()) %>%
  arrange(desc(count)) %>%
  head(10) %>% #change this number for more or less species
  pull(Detect_SpeciesCode)



for(j in 1:10) {
  #pick the species
  species = common_species[j]
  
#loop through each survey and bootstrap density estimates
    for(i in 1:37) {
      species_j <- list()
      
      year = as.numeric(distinct_samples$year[i])
      site = distinct_samples$site[i]
  
      df = bird_densities_by_transect %>%
          filter(site == .env$site,
               year == .env$year,
               species == .env$species)

      boot_out <- boot(data = df$density, statistic = mean_stat, R = 2000)

      ci <- boot.ci(boot_out, type = c("perc"))
  
      estimates <- data.frame(
        year = year,
        site = site,
        species = species,
        est  = ci$t0,
        lci  = ci$percent[4],
        uci  = ci$percent[5]
    )
  
      species_j[[j]] <- estimates
    }
  bootstrap_estimates[[j]] <- bind_rows(species_j)
}

bootstrap_estimates <- bind_rows(bootstrap_estimates)
```

```{r}
library(dplyr)
library(purrr)
library(boot)

# Statistic for boot(): must accept (data, indices) and return a single number
mean_stat <- function(data, indices) mean(data[indices], na.rm = TRUE)

bootstrap_estimates <- vector("list", length = 10)

common_species <- obs %>%
  filter(!Detect_SpeciesCode %in% c("UNCO", "BRMU")) %>%
  group_by(Detect_SpeciesCode) %>%
  summarise(count = n(), .groups = "drop") %>%
  arrange(desc(count)) %>%
  head(10) %>%
  pull(Detect_SpeciesCode)

# Drive inner loop by the actual size of distinct_samples
n_groups <- nrow(distinct_samples)

for (j in seq_along(common_species)) {
  species <- common_species[j]

  # Create the list ONCE per species
  species_rows <- vector("list", length = n_groups)

  for (i in seq_len(n_groups)) {
    year <- distinct_samples$year[i]
    site <- distinct_samples$site[i]

    # Ensure types are aligned; use .env to refer to external variables
    df <- bird_densities_by_transect %>%
      filter(site == .env$site,
             year == .env$year,
             species == .env$species)

    if (nrow(df) == 0 || all(is.na(df$density))) {
      # Nothing to bootstrap for this (species, year, site)
      species_rows[[i]] <- data.frame(
        year    = year,
        site    = site,
        species = species,
        est     = NA_real_,
        lci     = NA_real_,
        uci     = NA_real_
      )
      next
    }

    # Bootstrap
    boot_out <- boot(data = df$density, statistic = mean_stat, R = 2000)

    # Percentile CI; guard extraction
    ci <- try(boot.ci(boot_out, type = "perc"), silent = TRUE)

    lci <- if (is.list(ci) && !is.null(ci$percent)) ci$percent[4] else NA_real_
    uci <- if (is.list(ci) && !is.null(ci$percent)) ci$percent[5] else NA_real_

    species_rows[[i]] <- data.frame(
      year    = year,
      site    = site,
      species = species,
      est     = boot_out$t0,  # use boot's point estimate
      lci     = lci,
      uci     = uci
    )
  }

  # Bind all (year, site) rows for this species
  bootstrap_estimates[[j]] <- dplyr::bind_rows(species_rows)
}

# Final table: all species × (year, site)
bootstrap_estimates <- dplyr::bind_rows(bootstrap_estimates)
bootstrap_estimates
```

```{r}
bootstrap_estimates %>%
  filter(site != "KBAY") %>%
  ggplot() +
  geom_errorbar(aes(x = year, ymin = lci, ymax = uci), fill = "blue", alpha = 0.2) +
  geom_path(aes(x = year, y = est)) +
  geom_point(aes(x = year, y = est)) +
  facet_grid(species~ site, scales = "free_y")
```

```{r}
bootstrap_estimates %>%
  filter(site != "KBAY") %>%
ggplot( aes(year, est)) +
  geom_smooth(aes(ymin = lci, ymax = uci, color = site, group = site, fill = site),
              stat = "identity",
              alpha = 0.2) +
  geom_point() +
  facet_wrap(~ species, scales = "free_y", ncol = 2) +
  labs(title = "Bootstrapped average densities")

```

instead of using transects we could use habitat type, this should have less variance than transect and maybe make some better confidence intervals...?

import shorezone

```{r}
library(sf)
# List feature classes inside the .gdb
st_layers("C:/PRO_PROJECTS/PRO_SHARING/ShoreZone_pro_2026_0530_26ad97/commondata/shorezone.gdb")

# Read one layer
HabitatClass <- st_read("C:/PRO_PROJECTS/PRO_SHARING/ShoreZone_pro_2026_0530_26ad97/commondata/shorezone.gdb", layer = "HabitatClass")
CoastalClass <- st_read("C:/PRO_PROJECTS/PRO_SHARING/ShoreZone_pro_2026_0530_26ad97/commondata/shorezone.gdb", layer = "CoastalClass")
BiologicalWaveExposure <- st_read("C:/PRO_PROJECTS/PRO_SHARING/ShoreZone_pro_2026_0530_26ad97/commondata/shorezone.gdb", layer = "BiologicalWaveExposure")
ShoreType <- st_read("C:/PRO_PROJECTS/PRO_SHARING/ShoreZone_pro_2026_0530_26ad97/commondata/shorezone.gdb", layer = "ShoreType")

```

```{r}
obs_sf <- obs %>%
  st_as_sf(., coords = c("Long", "Lat"), crs = 4326)

HabitatClass <- st_transform(HabitatClass, 4326)
CoastalClass <- st_transform(CoastalClass, 4326)
BiologicalWaveExposure <- st_transform(BiologicalWaveExposure, 4326)
ShoreType <- st_transform(ShoreType, 4326)
```

```{r}
library(leaflet)

bbox <- st_bbox(obs_sf)

bbox_poly <- st_as_sfc(bbox)

HabitatClass <- st_filter(HabitatClass, bbox_poly)
CoastalClass <- st_filter(CoastalClass, bbox_poly)
BiologicalWaveExposure <- st_filter(BiologicalWaveExposure, bbox_poly)
ShoreType <- st_filter(BiologicalWaveExposure, bbox_poly)

#join habitat parameters with bird obs
obs_sf <- st_join(obs_sf, HabitatClass, join = st_nearest_feature)
obs_sf <- st_join(obs_sf, CoastalClass, join = st_nearest_feature)
obs_sf <- st_join(obs_sf, BiologicalWaveExposure, join = st_nearest_feature)
obs_sf <- st_join(obs_sf, ShoreType, join = st_nearest_feature)
```

now we need to see how many habitats we have

```{r}
obs_sf %>%
  st_drop_geometry() %>%
  group_by(HAB_CLASS_Name) %>%
  summarise(counts = n()) %>%
  arrange(desc(counts)) %>%
  ggplot() +
  geom_col(aes(HAB_CLASS_Name, counts)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
library(vegan)

bc.table <- bird_densities %>% 
  filter(!species %in% c("SOSA", "SEOT", "BRBE", "HASE", "STSL", "BLBE", "STJA", "MOOS", "GOEA", "HAPO", "DAPO", "GRWO", "RLHA", "GRWH", "KIWH", "NOHA", "FOSP", "TRES", "RUHU", "MIWH", "PIGR", "BOGU", "UNPE", "BOAT", "YELL", "PORC", "HUWH", "MACR", "UNWH", "REFO", "MOGO")) %>%
  select(-c(count, length, area)) %>%
  pivot_wider(names_from = species, values_from = density) %>%
  filter(!year %in% c(2006, 2011)) %>%
  filter(site != "KBAY")

#bray-curtis distance matrix
bc.matrix <- as.matrix(bc.table[,3:ncol(bc.table)], labels = T) #why did I subset this?
row.names(bc.matrix) <- paste(bc.table$site, bc.table$year, sep = "_")
bc.matrix <- vegdist(bc.matrix, method = "bray")
```

```{r}
by_survey_mds <- metaMDS(bc.matrix,
          distance = "bray",
          k = 3,
          maxit = 999, 
          trymax = 500,
          wascores = TRUE)
```

```{r}
data.scores = as.data.frame(scores(by_survey_mds))
data.scores <- data.scores %>%
  mutate(site = substr(row.names(.), start = 1, 4),
         year = substr(rownames(.), 6, 10))

data.scores <- data.scores %>%
  mutate(
    year = as.numeric(year),
    blob = case_when(
      year < 2014 ~ "pre",
      year >= 2014 ~ "post"
    ),
    salmon = case_when(
      year %% 2 == 0 ~ "even",
      year %% 2 == 1 ~ "odd"
    )
  )
```

```{r}
hulls <- data.scores %>%
  group_by(salmon) %>%
  slice(chull(NMDS1, NMDS2))

okabe_ito <- c(
  "#E69F00", "#56B4E9", "#009E73",
  "#0072B2", "#D55E00", "#CC79A7"
)

ggplot(data.scores, aes(NMDS1, NMDS2)) +
  geom_polygon(
    data = hulls,
    aes(fill = salmon, group = salmon),
    color = NA, alpha = 0.35
  ) +
  geom_point(
    aes(color = site, shape = site),
    size = 3, stroke = 0.4
  ) +
  scale_fill_viridis_d(option = "C") +       # options: "A","B","C","D","E"
  scale_color_manual(values = okabe_ito) +
  theme_minimal(base_size = 12)

```
try a lunatic exercise in plotting nmds scores against temp. this is basically an attempt to look at the blob using continuous variables
```{r}
temp_df %>%
  filter(block == "KEFJ" | block == "KATM") %>%
  left_join(data.scores, ., by = c("year", "site" = "block")) %>%
  mutate(avg_temp = as.numeric(scale(avg_temp))) %>%
  ggplot() +
  geom_path(aes(year, NMDS1, color = site, group = site)) +
  geom_path(aes(year, avg_temp, group = site))
```  
```{r}
temp_df %>%
  filter(block == "KEFJ" | block == "KATM") %>%
  mutate(year = as.numeric(year)) %>%
  left_join(data.scores, ., by = c("year", "site" = "block")) %>%
  mutate(avg_temp = as.numeric(scale(avg_temp))) %>%
  ggplot() +
  geom_point(aes(avg_temp, NMDS1))
```
now try grouping birds based on family
```{r}
bc.table <- bird_densities %>% 
  filter(!species %in% c("SOSA", "SEOT", "BRBE", "HASE", "STSL", "BLBE", "STJA", "MOOS", "GOEA", "HAPO", "DAPO", "GRWO", "RLHA", "GRWH", "KIWH", "NOHA", "FOSP", "TRES", "RUHU", "MIWH", "PIGR", "BOGU", "UNPE", "BOAT", "YELL", "PORC", "HUWH", "MACR", "UNWH", "REFO", "MOGO")) %>%
  select(-c(count, length, area)) %>%
  pivot_wider(names_from = species, values_from = density) %>%
  filter(!year %in% c(2006, 2011)) %>%
  filter(site != "KBAY")

bird_densities$species[bird_densities$species == "NOCR"] <- "AMCR"
bird_densities$species[bird_densities$species == "NOSH"] <- "NSHO"
bird_densities$species[bird_densities$species == "HEGU"] <- "HERG"

family.table <- bird_densities %>%
  filter(!species %in% c("SOSA", "SEOT", "BRBE", "HASE", "STSL", "BLBE", "STJA", "MOOS", "GOEA", "HAPO", "DAPO", "GRWO", "RLHA", "GRWH", "KIWH", "NOHA", "FOSP", "TRES", "RUHU", "MIWH", "PIGR", "BOGU", "UNPE", "BOAT", "YELL", "PORC", "HUWH", "MACR", "UNWH", "REFO", "MOGO", "PEFA", "RIOT", "HBWH", "GYRF", "UNJA")) %>%
  left_join(., nppsd, by = c("species" = "species_code")) %>%
  group_by(year, site, family) %>%
  summarise(count = sum(count),
            density = sum(log(density + 1))) %>%
  mutate(family = ifelse(family == "Laridae\xa0", "Laridae", family)) %>%
  select(-c(count)) %>%
  pivot_wider(names_from = family, values_from = density) %>%
  filter(!year %in% c(2006, 2011)) %>%
  filter(site != "KBAY") %>%
  ungroup()
```
```{r}
family.matrix <- as.matrix(family.table[,3:ncol(family.table)], labels = T) #why did I subset this?
row.names(family.matrix) <- paste(family.table$site, family.table$year, sep = "_")
family.matrix <- vegdist(family.matrix, method = "bray")
```

```{r}
family_mds <- metaMDS(family.matrix,
          distance = "bray",
          k = 3,
          maxit = 999, 
          trymax = 500,
          wascores = TRUE)
```

```{r}
data.scores = as.data.frame(scores(family_mds))
data.scores <- data.scores %>%
  mutate(site = substr(row.names(.), start = 1, 4),
         year = substr(rownames(.), 6, 10))

data.scores <- data.scores %>%
  mutate(
    year = as.numeric(year),
    blob = case_when(
      year < 2014 ~ "pre",
      year >= 2014 ~ "post"
    ),
    salmon = case_when(
      year %% 2 == 0 ~ "even",
      year %% 2 == 1 ~ "odd"
    )
  )
```

```{r}
hulls <- data.scores %>%
  group_by(blob, site) %>%
  slice(chull(NMDS1, NMDS2))

ggplot(data.scores, aes(NMDS1, NMDS2)) +
  geom_polygon(
    data = hulls,
    aes(fill = blob, group = blob),
    color = NA, alpha = 0.35
  ) +
  geom_point(
    aes(color = blob),
    size = 3, stroke = 0.4
  ) +       # options: "A","B","C","D","E"
  theme_minimal(base_size = 12) +
  facet_wrap(~site, scales = "free")
```

```{r}
#order families taxonomically
families <- bird_densities %>%
  filter(!species %in% c("SOSA", "SEOT", "BRBE", "HASE", "STSL", "BLBE", "STJA", "MOOS", "GOEA", "HAPO", "DAPO", "GRWO", "RLHA", "GRWH", "KIWH", "NOHA", "FOSP", "TRES", "RUHU", "MIWH", "PIGR", "BOGU", "UNPE", "BOAT", "YELL", "PORC", "HUWH", "MACR", "UNWH", "REFO", "MOGO", "PEFA", "RIOT", "HBWH", "GYRF", "UNJA")) %>%
  left_join(., nppsd, by = c("species" = "species_code")) %>%
  group_by(year, site, family) %>%
  summarise(count = sum(count),
            density = sum(log(density + 1))) %>%
  pull(family) %>%
  unique()

family_order <- data.frame(families) %>%
  mutate(keep = 1)
family_order <- left_join(nppsd, family_order, by = c("family" = "families")) %>%
  mutate(family = ifelse(family == "Laridae\xa0", "Laridae", family)) %>%
  filter(keep == 1) %>%
  pull(family) %>%
  unique() %>%
  as.vector()


#palette <- colorRampPalette(brewer.pal(9, "Spectral"))(17)
#palette <- sample(palette)   # randomize order

palette <- c("#F2FAAB", "#FDC775", "#88CFA4", "#FEEFA4", "#E6F598", "#ABDDA4", "#4CA4B1", "#C8E99E", "#FFFFBF", "#FEE08B", "#D53E4F", "#F88D51", "#66C2A5", "#F46D43", "#3288BD", "#E45549", "#FDAE61")

bird_densities %>%
  filter(!species %in% c("SOSA", "SEOT", "BRBE", "HASE", "STSL", "BLBE", "STJA", "MOOS", "GOEA", "HAPO", "DAPO", "GRWO", "RLHA", "GRWH", "KIWH", "NOHA", "FOSP", "TRES", "RUHU", "MIWH", "PIGR", "BOGU", "UNPE", "BOAT", "YELL", "PORC", "HUWH", "MACR", "UNWH", "REFO", "MOGO", "PEFA", "RIOT", "HBWH", "GYRF", "UNJA")) %>%
  left_join(., nppsd, by = c("species" = "species_code")) %>%
  group_by(year, site, family) %>%
  summarise(count = sum(count),
            density = sum(log(density + 1))) %>%
  mutate(family = ifelse(family == "Laridae\xa0", "Laridae", family),
         family = factor(family, levels = family_order)) %>%
  select(-c(count)) %>%
  ggplot(aes(x = year, y = density, fill = family)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    y = "Percent of total density",
    x = "Year",
    fill = "Family"
  ) +
  theme_minimal() +
  facet_wrap(~site, ncol = 1) +
  scale_fill_manual(values = palette)

```
```{r}
bird_densities %>%
  filter(!species %in% c("SOSA", "SEOT", "BRBE", "HASE", "STSL", "BLBE", "STJA", "MOOS", "GOEA", "HAPO", "DAPO", "GRWO", "RLHA", "GRWH", "KIWH", "NOHA", "FOSP", "TRES", "RUHU", "MIWH", "PIGR", "BOGU", "UNPE", "BOAT", "YELL", "PORC", "HUWH", "MACR", "UNWH", "REFO", "MOGO", "PEFA", "RIOT", "HBWH", "GYRF", "UNJA")) %>%
  left_join(., nppsd, by = c("species" = "species_code")) %>%
  group_by(year, site, species) %>%
  summarise(count = sum(count),
            density = sum(log(density + 1))) %>%
  select(-c(count)) %>%
  ggplot(aes(x = year, y = density, fill = species)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    y = "Percent of total density",
    x = "Year",
    fill = "Family"
  ) +
  theme_minimal() +
  facet_wrap(~site, ncol = 1)
```
ok the plan for today is to do this by diet and by functional group
```{r}
#bird_densities %>%
#  filter(!species %in% c("SOSA", "SEOT", "BRBE", "HASE", "STSL", "BLBE", "STJA", "MOOS", "GOEA", "HAPO", "DAPO", "GRWO", "RLHA", "GRWH", "KIWH", "NOHA", "FOSP", "TRES", "RUHU", #"MIWH", "PIGR", "BOGU", "UNPE", "BOAT", "YELL", "PORC", "HUWH", "MACR", "UNWH", "REFO", "MOGO", "PEFA", "RIOT", "HBWH", "GYRF", "UNJA")) %>%
#  pull(species) %>%
#  unique() %>%
#  data.frame() %>%
#  rename(species_code = ".") %>%
#  left_join(., nppsd, by = c("species_code" = "species_code")) %>%
#  write.csv(., row.names = FALSE, "data/functional_groups_template.csv")
  
functional_groups <- read.csv("data/functional_groups.csv")
diet_palette <- sample(colorRampPalette(brewer.pal(9, "Spectral"))(5))
diet_palette <- c("#FFFFBF", "#3288BD",  "#D53E4F", "#ABDDA4",  "#FDAE61")
bird_densities %>%
  filter(!species %in% c("SOSA", "SEOT", "BRBE", "HASE", "STSL", "BLBE", "STJA", "MOOS", "GOEA", "HAPO", "DAPO", "GRWO", "RLHA", "GRWH", "KIWH", "NOHA", "FOSP", "TRES", "RUHU", "MIWH", "PIGR", "BOGU", "UNPE", "BOAT", "YELL", "PORC", "HUWH", "MACR", "UNWH", "REFO", "MOGO", "PEFA", "RIOT", "HBWH", "GYRF", "UNJA")) %>%
  left_join(., nppsd, by = c("species" = "species_code")) %>%
  left_join(., functional_groups, by = c("species" = "species_code")) %>%
  group_by(year, site, diet) %>%
  summarise(count = sum(count),
            density = sum(log(density + 1))) %>%
  select(-c(count)) %>%
  ggplot(aes(x = year, y = density, fill = diet)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    y = "Percent of total density",
    x = "Year",
    fill = "Diet"
  ) +
  theme_minimal() +
  facet_wrap(~site, ncol = 1) +
  scale_fill_manual(values = diet_palette)
```

```{r}
bird_densities %>%
  filter(!species %in% c("SOSA", "SEOT", "BRBE", "HASE", "STSL", "BLBE", "STJA", "MOOS", "GOEA", "HAPO", "DAPO", "GRWO", "RLHA", "GRWH", "KIWH", "NOHA", "FOSP", "TRES", "RUHU", "MIWH", "PIGR", "BOGU", "UNPE", "BOAT", "YELL", "PORC", "HUWH", "MACR", "UNWH", "REFO", "MOGO", "PEFA", "RIOT", "HBWH", "GYRF", "UNJA")) %>%
  left_join(., nppsd, by = c("species" = "species_code")) %>%
  left_join(., functional_groups, by = c("species" = "species_code")) %>%
  group_by(year, site, functional_group) %>%
  summarise(count = sum(count),
            density = sum(log(density + 1))) %>%
  select(-c(count)) %>%
  ggplot(aes(x = year, y = density, fill = functional_group)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    y = "Percent of total density",
    x = "Year",
    fill = "Functional Group"
  ) +
  theme_minimal() +
  facet_wrap(~site, ncol = 1) +
  scale_fill_manual(values = colorRampPalette(brewer.pal(9, "Spectral"))(3))
```
make ecological groups by comining diet and foraging strategy
```{r}
#niche_palette <- sample(colorRampPalette(brewer.pal(9, "Spectral"))(11))

niche_palette <- c("#3288BD", "#C2E69F", "#EBF79F", "#5BB6A9", "#FFFFBF", "#FDC271", "#8FD2A4", "#FEE695", "#F99455", "#D53E4F", "#ED6345")


bird_densities %>%
  filter(!species %in% c("SOSA", "SEOT", "BRBE", "HASE", "STSL", "BLBE", "STJA", "MOOS", "GOEA", "HAPO", "DAPO", "GRWO", "RLHA", "GRWH", "KIWH", "NOHA", "FOSP", "TRES", "RUHU", "MIWH", "PIGR", "BOGU", "UNPE", "BOAT", "YELL", "PORC", "HUWH", "MACR", "UNWH", "REFO", "MOGO", "PEFA", "RIOT", "HBWH", "GYRF", "UNJA")) %>%
  left_join(., nppsd, by = c("species" = "species_code")) %>%
  left_join(., functional_groups, by = c("species" = "species_code")) %>%
  mutate(niche = paste(functional_group, diet, sep = " ")) %>%
  group_by(year, site, niche) %>%
  summarise(count = sum(count),
            density = sum(log(density + 1))) %>%
  select(-c(count)) %>%
  ggplot(aes(x = year, y = density, fill = niche)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    y = "Percent of total density",
    x = "Year",
    fill = "Niche"
  ) +
  theme_minimal() +
  facet_wrap(~site, ncol = 1) +
  scale_fill_manual(values = niche_palette)
```

```{r}
bird_densities %>%
  filter(!species %in% c("SOSA", "SEOT", "BRBE", "HASE", "STSL", "BLBE", "STJA", "MOOS", "GOEA", "HAPO", "DAPO", "GRWO", "RLHA", "GRWH", "KIWH", "NOHA", "FOSP", "TRES", "RUHU", "MIWH", "PIGR", "BOGU", "UNPE", "BOAT", "YELL", "PORC", "HUWH", "MACR", "UNWH", "REFO", "MOGO", "PEFA", "RIOT", "HBWH", "GYRF", "UNJA")) %>%
  left_join(., nppsd, by = c("species" = "species_code")) %>%
  left_join(., functional_groups, by = c("species" = "species_code")) %>%
  mutate(niche = paste(functional_group, diet, sep = " ")) %>%
  group_by(year, site, niche) %>%
  summarise(count = sum(count),
            density = sum(log(density + 1))) %>%
  mutate(density_scaled = as.numeric(scale(density))) %>%
  ggplot(aes(year, density_scaled, color = niche, group = niche)) +
  geom_path() +
  geom_point() +
  facet_wrap(~site, ncol = 1) +
  theme_minimal()
```

ok the question is how redundant these groups are. figure out how to plot that
```{r}
library(ggridges)

bird_densities %>%
  filter(!species %in% c("SOSA", "SEOT", "BRBE", "HASE", "STSL", "BLBE", "STJA", "MOOS", "GOEA", "HAPO", "DAPO", "GRWO", "RLHA", "GRWH", "KIWH", "NOHA", "FOSP", "TRES", "RUHU", "MIWH", "PIGR", "BOGU", "UNPE", "BOAT", "YELL", "PORC", "HUWH", "MACR", "UNWH", "REFO", "MOGO", "PEFA", "RIOT", "HBWH", "GYRF", "UNJA")) %>%
  left_join(., nppsd, by = c("species" = "species_code")) %>%
  left_join(., functional_groups, by = c("species" = "species_code")) %>%
  mutate(niche = paste(functional_group, diet, sep = " ")) %>%
  filter(density != 0) %>%
  group_by(year, site, niche) %>%
  summarise(count = sum(count),
            density = sum(log(density + 1)),
            n_species = n_distinct(species)) %>%
  ungroup() %>%
  ggplot() +
  geom_col(aes(year, n_species, fill = niche), position = "dodge") +
  labs(
    y = "Number of Species per Niche",
    x = "Year",
    fill = "Niche"
  ) +
  theme_minimal() +
  facet_wrap(~site, ncol = 1) +
  labs(title = "Ecosystem redundancy")
```
```{r}
bird_densities %>%
  filter(!species %in% c("SOSA", "SEOT", "BRBE", "HASE", "STSL", "BLBE", "STJA", "MOOS", "GOEA", "HAPO", "DAPO", "GRWO", "RLHA", "GRWH", "KIWH", "NOHA", "FOSP", "TRES", "RUHU", "MIWH", "PIGR", "BOGU", "UNPE", "BOAT", "YELL", "PORC", "HUWH", "MACR", "UNWH", "REFO", "MOGO", "PEFA", "RIOT", "HBWH", "GYRF", "UNJA")) %>%
  left_join(., nppsd, by = c("species" = "species_code")) %>%
  left_join(., functional_groups, by = c("species" = "species_code")) %>%
  mutate(niche = paste(functional_group, diet, sep = " ")) %>%
  filter(density != 0) %>%
  group_by(year, site, diet) %>%
  summarise(count = sum(count),
            density = sum(log(density + 1)),
            n_species = n_distinct(species)) %>%
  ungroup() %>%
  ggplot() +
  geom_col(aes(year, n_species, fill = diet), position = "dodge") +
  labs(
    y = "Number of Species per Niche",
    x = "Year",
    fill = "Niche"
  ) +
  theme_minimal() +
  facet_wrap(~site, ncol = 1) +
  labs(title = "Ecosystem redundancy")
```

```{r}
bird_densities %>%
  filter(!species %in% c("SOSA", "SEOT", "BRBE", "HASE", "STSL", "BLBE", "STJA", "MOOS", "GOEA",
                         "HAPO", "DAPO", "GRWO", "RLHA", "GRWH", "KIWH", "NOHA", "FOSP", "TRES",
                         "RUHU", "MIWH", "PIGR", "BOGU", "UNPE", "BOAT", "YELL", "PORC", "HUWH",
                         "MACR", "UNWH", "REFO", "MOGO", "PEFA", "RIOT", "HBWH", "GYRF", "UNJA")) %>%
  left_join(nppsd, by = c("species" = "species_code")) %>%
  left_join(functional_groups, by = c("species" = "species_code")) %>%
  mutate(niche = paste(functional_group, diet, sep = " ")) %>%
  filter(density != 0) %>%
  filter(!niche %in% c("land duck food", "land fish", "surface inverts")) %>%
  group_by(year, site, niche) %>%
  summarise(
    count = sum(count),
    density = sum(log(density + 1)),
    n_species = n_distinct(species),
    .groups = "drop"
  ) %>%
  group_by(site, niche) %>%          # compute anomaly baseline per niche
  mutate(mean_species = mean(n_species),
         anomaly = n_species - mean_species) %>%
  ungroup() %>%
  ggplot(aes(x = year, y = anomaly, fill = site, group = site)) +
  geom_col(position = "dodge") +
  labs(
    y = "Species Anomaly (Deviation From Mean)",
    x = "Year",
    fill = "Niche"
  ) +
  theme_minimal() +
  facet_wrap(~niche, ncol = 1) +
  geom_hline(yintercept = 0)
```
this should probably be modeled...
```{r}
niche_df <- bird_densities %>%
  filter(!species %in% c("SOSA", "SEOT", "BRBE", "HASE", "STSL", "BLBE", "STJA", "MOOS", "GOEA",
                         "HAPO", "DAPO", "GRWO", "RLHA", "GRWH", "KIWH", "NOHA", "FOSP", "TRES",
                         "RUHU", "MIWH", "PIGR", "BOGU", "UNPE", "BOAT", "YELL", "PORC", "HUWH",
                         "MACR", "UNWH", "REFO", "MOGO", "PEFA", "RIOT", "HBWH", "GYRF", "UNJA")) %>%
  left_join(nppsd, by = c("species" = "species_code")) %>%
  left_join(functional_groups, by = c("species" = "species_code")) %>%
  mutate(niche = paste(functional_group, diet, sep = " ")) %>%
  filter(density != 0) %>%
  filter(!niche %in% c("land duck food", "land fish", "surface inverts")) %>%
  group_by(year, site, niche) %>%
  summarise(
    count = sum(count),
    density = sum(log(density + 1)),
    n_species = n_distinct(species),
    .groups = "drop"
  ) %>%
  group_by(site, niche) %>%          # compute anomaly baseline per niche
  mutate(mean_species = mean(n_species),
         anomaly = n_species - mean_species) %>%
  ungroup() %>%
  mutate(year = as.numeric(year)) %>%
  left_join(., gak_Summary)

niche_df$blob[niche_df$year < 2014] <- "pre"
niche_df$blob[niche_df$year >= 2014] <- "post"

glm(density ~ temp*year + niche + site + year + blob, data = niche_df) %>% summary()
fit <- glm(density ~ temp*year + niche + site + year + blob, data = niche_df)

effect_data <- ggpredict(fit, terms = c("niche", "site", "year"))

plot(effect_data)
```
```{r}
library(bcp)
gak_Summary %>%
  ggplot(aes(year, temp)) +
  geom_path() +
  geom_point()

set.seed(101)
bcp_gak1 = bcp(gak_Summary$temp, p0 = 0.008, burnin = 1000, 
             mcmc = 10000)

plot(bcp_gak1, main = "GAK1 Temp changepoints")
```
```{r}
diver_bcp <- niche_df %>%
  filter(niche == "diving fish")
bcp_gak1 = bcp(diver_bcp$anomaly, p0 = 0.008, burnin = 1000, 
             mcmc = 10000)

plot(bcp_gak1, main = "GAK1 Temp changepoints")
```

```{r}
library(DirichletReg)

niche_df <- bird_densities %>%
  filter(!species %in% c("SOSA", "SEOT", "BRBE", "HASE", "STSL", "BLBE", "STJA", "MOOS", "GOEA",
                         "HAPO", "DAPO", "GRWO", "RLHA", "GRWH", "KIWH", "NOHA", "FOSP", "TRES",
                         "RUHU", "MIWH", "PIGR", "BOGU", "UNPE", "BOAT", "YELL", "PORC", "HUWH",
                         "MACR", "UNWH", "REFO", "MOGO", "PEFA", "RIOT", "HBWH", "GYRF", "UNJA")) %>%
  left_join(nppsd, by = c("species" = "species_code")) %>%
  left_join(functional_groups, by = c("species" = "species_code")) %>%
  mutate(niche = paste(functional_group, diet, sep = " ")) %>%
  filter(density != 0) %>%
  filter(!niche %in% c("land duck food", "land fish", "surface inverts")) %>%
  group_by(year, site, niche) %>%
  summarise(prop = density / sum(density))

prop_df <- bird_densities %>%
filter(!species %in% c("SOSA", "SEOT", "BRBE", "HASE", "STSL", "BLBE", "STJA", "MOOS", "GOEA",
                         "HAPO", "DAPO", "GRWO", "RLHA", "GRWH", "KIWH", "NOHA", "FOSP", "TRES",
                         "RUHU", "MIWH", "PIGR", "BOGU", "UNPE", "BOAT", "YELL", "PORC", "HUWH",
                         "MACR", "UNWH", "REFO", "MOGO", "PEFA", "RIOT", "HBWH", "GYRF", "UNJA")) %>%
  left_join(nppsd,              by = c("species" = "species_code")) %>%
  left_join(functional_groups,  by = c("species" = "species_code")) %>%
  mutate(niche = paste(functional_group, diet, sep = " ")) %>%
  filter(density != 0,
         !niche %in% c("land duck food", "land fish", "surface inverts")) %>%
  # 1) Aggregate to niche within site-year
  group_by(year, site, niche) %>%
  summarise(density_niche = sum(log(density + 1)), .groups = "drop") %>%
  # 2) Compute proportions across niches within each site-year
  group_by(year, site) %>%
  mutate(prop = density_niche / sum(density_niche)) %>%
  ungroup()


prop_wide <- prop_df %>% 
  select(-density_niche) %>%
  pivot_wider(names_from = niche, values_from = prop) %>%
  mutate(across(3:10, ~ replace_na(., 0))) %>%
  select(-`diving plankton`) %>%
  mutate(year = as.numeric(year)) %>%
  left_join(gak_Summary, by = "year")
```

```{r}
prop_df %>%
  filter(site != "KBAY") %>%
  ggplot(aes(year, prop, color = niche, group = niche)) +
  geom_path() +
  facet_wrap(~site, ncol = 1) +
  labs(title = "Proportion of total abundance by niche")
```

```{r}
library(streamgraph)

prop_df %>%
  filter(site == "KEFJ") %>%
  streamgraph(key = "niche", value = "density_niche", date = "year", offset = "silhouette") %>%
  sg_axis_x(1, "year", "%Y") %>%
  sg_legend(show=TRUE, label="Niches: ")
```

```{r}
prop_wide %>%
  ggplot(aes(`diving fish`, `diving inverts`)) +
  geom_point(aes(color = site)) +
  geom_smooth(method = "lm") +
  labs(title = "Diving invert ~ diving fish") +
  facet_wrap(~site, ncol = 1, scales = "free")
```

```{r}
install.packages("plotly")
library(plotly)


# --- 1) Put your data frame name here ----------------------------------------
# Assuming your data is in 'prop_wide' with columns shown in your preview:
# year, site, `diving fish`, `diving inverts`, ..., temp, temp_scaled
df <- prop_wide %>%
  rename_with(janitor::make_clean_names)  # spaces -> underscores

make_surface_for_site <- function(data, site_name,
                                  x_col = "diving_inverts",
                                  y_col = "diving_fish",
                                  z_col = "temp_scaled",
                                  nx = 60, ny = 60,
                                  extrap = TRUE) {

  df_s <- data %>%
    filter(site == site_name) %>%
    select(all_of(c(x_col, y_col, z_col))) %>%
    drop_na()

  if(nrow(df_s) < 5) stop("Not enough points for site: ", site_name)

  # Interpolate irregular data onto a grid
  interp_res <- akima::interp(
    x = df_s[[x_col]],
    y = df_s[[y_col]],
    z = df_s[[z_col]],
    duplicate = "mean",
    nx = nx, ny = ny,
    extrap = extrap
  )

  # Plot surface
  plot_ly(
    x = interp_res$x,
    y = interp_res$y,
    z = interp_res$z,
    type = "surface",
    colorscale = "Viridis"
  ) %>%
    add_markers(
      data = df_s,
      x = df_s[[x_col]],
      y = df_s[[y_col]],
      z = df_s[[z_col]],
      marker = list(size = 3, color = "red"),
      inherit = FALSE
    ) %>%
    layout(
      title = paste("Surface for site:", site_name),
      scene = list(
        xaxis = list(title = x_col),
        yaxis = list(title = y_col),
        zaxis = list(title = z_col)
      )
    )
}


# --- 2) Make surfaces for each site ------------------------------------------
p_katm <- make_surface_for_site(df, "KATM")
p_kefj <- make_surface_for_site(df, "KEFJ")

# View each surface
p_katm
p_kefj

# Or show them side-by-side
subplot(p_katm, p_kefj, nrows = 1, margin = 0.05, titleX = TRUE, titleY = TRUE)

```
```{r}
#try some models
prop_wide <- janitor::clean_names(prop_wide)
fit.null <- glm(diving_fish ~ 1, data = prop_wide, family = tweedie())
fit.1 <- glm(diving_fish ~ year, data = prop_wide, family = tweedie())
fit.2 <- glm(diving_fish ~ year + site, data = prop_wide, family = tweedie())
fit.3 <- glm(diving_fish ~ year + site + diving_inverts, data = prop_wide, family = tweedie())
fit.4 <- glm(diving_fish ~ year + site + temp, data = prop_wide, family = tweedie())
fit.5 <- glm(diving_fish ~ temp, data = prop_wide, family = tweedie())
fit.6 <- glm(diving_fish ~ diving_inverts, data = prop_wide, family = tweedie())

AIC(fit.null, fit.1, fit.2, fit.3, fit.4, fit.5, fit.6) %>% arrange(desc(AIC))
```

```{r}
summary(fit.5)
```
see if there is a way to make these into good models. 
why is the above graph such a good relationship but the fish eaters vs benthivores relationship wasnt obvious from the streamplot
