---
title: "R Notebook"
output: html_notebook
---

```{r warning=FALSE}
library(tidyverse)
library(janitor)
```

```{r read in data}
obs <- read.csv("data/KBAYKATMKEFJ_2006-2023_Summer_MBM_Observations.csv")
transects <- read.csv("data/KBAYKATMKEFJ_2006-2023_Summer_MBM_Transects.csv")
```

```{r make a single df}
#make year and site columns from the trip ID
obs <- obs %>%
  mutate(
    site = substr(TripID, 1, 4),
    year = substr(TripID, 8, 11)
  ) %>%
  filter()

species <- unique(obs$Detect_SpeciesCode)
sites <- unique(obs$site)
years <- unique(obs$year)

#make transects by site
transects <- transects  %>%
  mutate(
    site = substr(TripID, 1, 4),
    year = substr(TripID, 8, 11)
  )

#identify pelagic transects
pelagic_transects <- transects %>%
  filter(TransType == "PELAGIC") %>%
  pull(TXNumber) %>%
  unique()

#now remove pelagic transects from bird obs
obs <- obs %>% filter(!TXNumber %in% pelagic_transects)
```

link obs with transect effort
```{r}
#sum species total for each survey (year and site)
obs_summary <-obs %>%
  group_by(year, site, Detect_SpeciesCode) %>%
  summarise(count = sum(Count)) %>%
  rename(species = Detect_SpeciesCode)

#find all unique combinations of year and site then repeat the bird list for each of them
sampling_events <- obs %>%
  distinct(year, site)

species_df <- species %>% 
  data.frame() %>%
  rename(species = ".")

species_list <- list()

for(i in 1:nrow(sampling_events)) {
  event <- sampling_events[i,]
  species_df$year <- event$year
  species_df$site <- event$site
  
  species_list[[i]] <- species_df
}

#convert list to dataframe
sampling_events <- bind_rows(species_list)

#this adds in zeros for the species not observed on each survey but doesnt make up surveys that didn't happen (as opposed to expand_grid)
obs_summary <- left_join(sampling_events, obs_summary) %>%
  mutate(count = replace_na(count, 0))

#this adds up all the area and distance sampled in each year
transects_summary <- transects %>% 
  group_by(year, site) %>%
  summarise(length = sum(Length_km),
            area = sum(Area_Sampled_km2))

#add the effort to the counts and calculate density
bird_densities <- left_join(obs_summary, transects_summary, by = c("site", "year")) %>%
  mutate(density = count/area) %>%
  relocate(species, .after = site)
```
find the most common birds to look at
```{r}
birds <- obs %>%
  filter(!Detect_SpeciesCode %in% c("UNCO")) %>%
  group_by(Detect_SpeciesCode) %>%
  summarise(count = sum(Count)) %>%
  arrange(desc(count)) %>%
  head(20) %>% #change this number for more or less species
  pull(Detect_SpeciesCode)
```

```{r}
bird_densities %>%
  filter(species %in% birds) %>%
  #mutate(density = scale(density)) %>%
  ggplot() +
  geom_path(aes(year, density, colour = site, group = site)) +
  geom_point(aes(year, density, colour = site, group = site)) +
  facet_wrap(~species, scales = "free_y") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r}
bird_densities %>%
  filter(species %in% birds) %>%
  mutate(density = scale(density)) %>%
  ggplot() +
  geom_path(aes(year, density, colour = species, group = species)) +
  facet_wrap(~site, scales = "free_y", nrow = 3) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r}
bird_densities %>%
  filter(species == "BLOY") %>%
  #mutate(density = scale(density)) %>%
  ggplot() +
  geom_path(aes(year, density, colour = site, group = site)) +
  facet_wrap(~species, scales = "free_y") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r}
mussel_counts <- read.csv("data/KATMKEFJWPWS_2008-2025_Mytilus_20mm_Count.csv")
mussel_size <- read.csv("data/KATMKEFJWPWS_2008-2025_Mytilus_20mm_Size.csv")
seastar_counts <- read.csv("data/KATMKEFJWPWS_2006-2025_Sea_Star_Count.csv")
```

```{r}
mussel_counts %>%
  filter(!BlockName %in% c("EPWS", "NPWS")) %>%
  mutate(Count = as.numeric(Count)) %>%
  group_by(BlockName, SiteName, YearSample) %>%
  summarise(avg = mean(Count)) %>%
  ggplot() +
  geom_point(aes(YearSample, avg, group = SiteName, colour = SiteName)) +
  facet_wrap(~BlockName, nrow = 3) +
  labs(title = "Mussel Counts")
```

```{r}
mussel_size %>%
  filter(!BlockName %in% c("EPWS", "NPWS")) %>%
  mutate(Size = as.numeric(Size)) %>%
  group_by(BlockName, SiteName, YearSample) %>%
  summarise(avg = mean(Size)) %>%
  ggplot() +
  geom_point(aes(YearSample, avg, group = SiteName, colour = SiteName)) +
  facet_wrap(~BlockName, nrow = 3) +
  labs(title = "Mussel Size")
```

```{r}
mussel_counts %>%
  filter(!BlockName %in% c("EPWS", "NPWS")) %>%
  mutate(Count = as.numeric(Count)) %>%
  group_by(BlockName, SiteName, YearSample) %>%
  summarise(avg = mean(Count)) %>%
  ggplot() +
  geom_boxplot(aes(YearSample, avg, group = YearSample)) +
  facet_wrap(~BlockName, nrow = 3, scales = "free_y") +
  labs(title = "Mussel Counts",
       y = "Count")
```

```{r}
mussel_size %>%
  filter(!BlockName %in% c("EPWS", "NPWS")) %>%
  mutate(Size = as.numeric(Size)) %>%
  group_by(BlockName, SiteName, YearSample) %>%
  summarise(avg = mean(Size)) %>%
  ggplot() +
  geom_boxplot(aes(YearSample, avg, group = YearSample)) +
  facet_wrap(~BlockName, nrow = 3, scales = "free_y") +
  labs(title = "Mussel Size",
       y = "Size (mm)")
```
noww check out the seastars
```{r}
seastar_counts %>%
  mutate(site = substr(SiteID, 1, 3)) %>%
  group_by(Year, site, Species) %>%
  summarise(avg_density = mean(Density_individual200SqM)) %>%
  ggplot() +
  geom_path(aes(Year, avg_density, color = Species)) +
  facet_wrap(~site, nrow = 3, scales = "free_y")
```

```{r}
seastar_counts %>%
  mutate(site = substr(SiteID, 1, 3)) %>%
  group_by(Year, site, Species) %>%
  summarise(avg_density = mean(Density_individual200SqM)) %>%
  ggplot() +
  geom_smooth(aes(Year, avg_density, group = site, color = site)) +
  facet_wrap(~Species, nrow = 3, scales = "free_y")
```

```{r}
bird_densities %>%
  group_by(year, site) %>%
  summarise(avg_density = mean(density)) %>%
  ggplot() +
  geom_path(aes(year, avg_density, color = site, group = site)) +
  labs(title = "all birds combined",
       subtitle = "is this stupid?")
```
combine everything into a single data frame for plotting
```{r}
seastar_wide <- seastar_counts %>%
  mutate(site = substr(SiteID, 1, 3),
         site = case_when(site == "AKP" ~ "KATM",
                          site == "PWS" ~ "WPWS",
                          site == "KEP" ~ "KEFJ")) %>%
  group_by(Year, site, Species) %>%
  summarise(avg_density = mean(Density_individual200SqM)) %>%
  pivot_wider(names_from = Species, values_from = avg_density) %>%
  rename(year = Year)

mussels_wide <- mussel_counts %>%
  rename(site = BlockName,
         year = YearSample) %>%
  filter(!site %in% c("EPWS", "NPWS")) %>%
  mutate(Count = as.numeric(Count)) %>%
  group_by(site, year) %>%
  summarise(avg_count = mean(Count, na.rm = T)) # check that this is ok and isnt getting rid of zeros

birds_wide <- bird_densities %>%
  select(!c(length, area, count)) %>%
  pivot_wider(names_from = species, values_from = density) %>%
  mutate(year = as.numeric(year))

all_taxa <- left_join(birds_wide, seastar_wide, by = c("year", "site")) %>%
  left_join(., mussels_wide, by = c("year", "site")) %>%
  rename(mussel_count = avg_count) %>%
  mutate(
    across(
      c(seq(3, ncol(.))),   
      ~ scale(.)[,1]
    )
  ) %>%
  rename_with(janitor::make_clean_names, .cols = 128:135)
```

```{r}
all_taxa %>%
  filter(site != "KBAY") %>%
  select(year, site, mussel_count, BLOY, HADU, pycnopodia_helianthoides) %>%
  pivot_longer(cols = !c(year, site), names_to = "species", values_to = "Abundance") %>%
  ggplot() +
  geom_path(aes(year, Abundance, color = species)) +
  geom_point(aes(year, Abundance, color = species)) +
  geom_hline(aes( yintercept = 0)) +
  facet_wrap(~site, nrow = 3)
```

```{r}
all_taxa %>%
  ggplot() +
  geom_point(aes(mussel_count, pisaster_ochraceus))
```

```{r}
bloy_nest <- read.csv("data/KBAYKATMKEFJWPWS_2006-2025_blackOystercatcher_Transect_Summary.csv")

bloy_nest %>%
  group_by(Year, BlockName) %>%
  summarise(avg_density = mean(as.numeric(NestDensity))) %>%
  ggplot() +
  geom_path(aes(Year, avg_density, color = BlockName)) +
  labs(title = "BLOY Density")
```
```{r}
isotopes <- read.csv("data/Isotope_data_all_regions_2014-2025.csv", fileEncoding='latin1')
isotopes <- isotopes %>% select(!c(X, X.1, X.2))

```

```{r}
isotopes %>%
  mutate(Year = as.factor(Year)) %>%
  ggplot() +
  geom_jitter(aes(d15N...., d13C...., colour = Year), alpha = 0.4)
  facet_wrap(~Year)
```

```{r}
isotopes %>%
  mutate(Year = as.factor(Year)) %>%
  ggplot() +
  geom_jitter(aes(d15N...., d13C...., colour = Species), alpha = 0.7) +
  facet_wrap(~Year)
```
```{r}
isotopes %>%
  filter(Species == "trossulus") %>%
  mutate(Year = as.factor(Year)) %>%
  ggplot() +
  geom_jitter(aes(d15N...., d13C...., colour = Year), alpha = 0.9, size = 3) +
    scale_x_continuous(limits = c(5, 10))
```
```{r}
trossulus <-isotopes %>%
  filter(Species == "trossulus") %>%
  filter(!is.na(d15N....),
         !is.na(d13C....)) %>%
mutate(Year = as.factor(Year))

hulls <- trossulus %>%
  group_by(Year) %>%
  slice(chull(d15N...., d13C....))


ggplot(trossulus) +
  geom_jitter(aes(x = d15N...., y = d13C...., colour = Year),
              alpha = 0.9, size = 3) +
  geom_polygon(data = hulls,
               aes(x = d15N...., y = d13C...., group = Year),
               alpha = 0.2, colour = "black") +
  scale_x_continuous(limits = c(5, 10)) +
  facet_wrap(~Year)
```

```{r}
trossulus %>%
  ggplot() +
  geom_jitter(aes(d15N...., d13C...., colour = Region), alpha = 0.5, size = 3) +
  facet_wrap(~Year) +
  labs(title = "mussel isotopes")
```
```{r}
isotopes %>%
  filter(Sample.ID == "POM") %>%
  ggplot() +
  geom_jitter(aes(d15N...., d13C...., colour = Region), alpha = 0.5, size = 3) +
  facet_wrap(~Year) +
  labs(title = "POM")
```


Limpet time!
```{r}
limpet_count <- read.csv("data/KATMKEFJWPWS_2010-2025_Limpet_Count.csv")
limpet_size <- read.csv("data/KATMKEFJWPWS_2006-2025_Limpet_Size.csv")
```

```{r}
limpet_count %>%
  mutate(site = substr(SiteID, 1, 3),
         site = case_when(site == "AKP" ~ "KATM",
                          site == "PWS" ~ "WPWS",
                          site == "KEP" ~ "KEFJ"),
         Density = as.numeric(Density)) %>%
  group_by(Year, site) %>%
  summarise(Density = mean(Density, na.rm = T)) %>%
  mutate(Density = scale(Density)[,1]) %>%
  ggplot() +
  geom_path(aes(Year, Density, color = site)) +
  geom_point(aes(Year, Density, color = site)) +
  labs(title = "limpet density")
```

The limpet abundances between KATM and KEFJ seem to trend oppositely 
```{r}
limpet_size %>%
  mutate(site = substr(SiteID, 1, 3),
         site = case_when(site == "AKP" ~ "KATM",
                          site == "PWS" ~ "WPWS",
                          site == "KEP" ~ "KEFJ"),
         Limpet_Size_mm = as.numeric(Limpet_Size_mm)) %>%
  group_by(Year, site) %>%
  summarise(size_avg = mean(Limpet_Size_mm, na.rm = T)) %>%
  #mutate(size_avg = scale(size_avg)[,1]) %>%
  ggplot() +
  geom_path(aes(Year, size_avg, color = site)) +
  geom_point(aes(Year, size_avg, color = site)) +
  labs(title = "limpet size")
```

but the size fluctuations seem to trend similarly between sites
```{r}
limpet_size %>%
  mutate(site = substr(SiteID, 1, 3),
         site = case_when(site == "AKP" ~ "KATM",
                          site == "PWS" ~ "WPWS",
                          site == "KEP" ~ "KEFJ"),
         Limpet_Size_mm = as.numeric(Limpet_Size_mm)) %>%
  group_by(Year, site, SiteName) %>%
  summarise(size_avg = mean(Limpet_Size_mm, na.rm = T)) %>%
  #mutate(size_avg = scale(size_avg)[,1]) %>%
  ggplot() +
  geom_path(aes(Year, size_avg, color = SiteName)) +
  geom_point(aes(Year, size_avg, color = SiteName)) +
  facet_wrap(~site, ncol = 1) +
  labs(title = "limpet size by site") +
  theme(legend.position = "none")
```

snails:
```{r}
snails <- read.csv("data/KATMKEFJWPWS_2006-2025_Nucella_Katharina_Lirabuccinum_Count.csv")
```

```{r}
snails %>%
  mutate(site = substr(SiteID, 1, 3),
         site = case_when(site == "AKP" ~ "KATM",
                          site == "PWS" ~ "WPWS",
                          site == "KEP" ~ "KEFJ")) %>%
  group_by(Year, Elevation_Position, site, SpeciesName) %>%
  summarise(count = mean(Count)) %>%
  ggplot() +
  geom_path(aes(Year, count, color = SpeciesName)) +
  geom_point(aes(Year, count, color = SpeciesName)) +
  facet_grid(site~Elevation_Position) +
  labs(title = "snail abundance")
```

check out the temperature data
```{r message=FALSE}
temperature <- list.files("data/temperature",
                       pattern = "\\.csv$",
                       full.names = TRUE) %>%
  map_dfr(read_csv)

#make month and year columns
temperature$year <- year(temperature$date)
temperature$month <- month(temperature$date)
```

```{r}
temperature %>%
  filter(exposure == "water",
         tide_height == 0.5) %>%
  ggplot() +
  geom_path(aes(date, temperature, color = site)) 
```

```{r}
temperature %>%
  filter(exposure == "water",
         month == 7,
         tide_height == 0.5) %>%
  ggplot() +
  geom_boxplot(aes(year, temperature, group = year, color = block)) +
  facet_wrap(~block, ncol =1)
```
```{r}
temperature %>%
  filter(exposure == "water",
         month == 5,
         tide_height == 0.5) %>%
  group_by(year, block) %>%
  summarise(avg_temp = mean(temperature)) %>%
  ggplot() +
  geom_path(aes(year, avg_temp, color = block)) +
  geom_point(aes(year, avg_temp, color = block)) +
  labs(title = "May Temp")
```
```{r}
temperature %>%
  filter(exposure == "water",
         month == 6,
         tide_height == 0.5) %>%
  group_by(year, block) %>%
  summarise(avg_temp = mean(temperature)) %>%
  ggplot() +
  geom_path(aes(year, avg_temp, color = block)) +
  geom_point(aes(year, avg_temp, color = block)) +
  labs(title = "June Temp")
```
```{r}
temperature %>%
  filter(exposure == "water",
         month == 7,
         tide_height == 0.5) %>%
  group_by(year, block) %>%
  summarise(avg_temp = mean(temperature)) %>%
  ggplot() +
  geom_path(aes(year, avg_temp, color = block)) +
  geom_point(aes(year, avg_temp, color = block)) +
  labs(title = "July Temp")
```

```{r}

temperature %>%
  filter(exposure == "water",
         month %in% c(4,5,6,7,8,9),
         tide_height == 0.5) %>%
  mutate(month_name = factor(month.name[month],
                             levels = month.name[4:9])) %>%
  group_by(year, block, month_name) %>%
  summarise(avg_temp = mean(temperature), .groups = "drop") %>%
  ggplot() +
  geom_path(aes(year, avg_temp, color = block)) +
  geom_point(aes(year, avg_temp, color = block)) +
  facet_wrap(~month_name)
```

```{r}
temperature %>%
  filter(exposure == "water",
         month %in% c(10,11,12,1,2,3),
         tide_height == 0.5) %>%
  mutate(month_name = factor(month.name[month],
                             levels = c("October","November","December",
                                        "January","February","March"))) %>%
  group_by(year, block, month_name) %>%
  summarise(avg_temp = mean(temperature), .groups = "drop") %>%
  ggplot() +
  geom_path(aes(year, avg_temp, color = block)) +
  geom_point(aes(year, avg_temp, color = block)) +
  facet_wrap(~month_name)
```
```{r}
winter_df <- temperature %>%
  filter(exposure == "water",
         month %in% c(10,11,12,1,2,3),
         tide_height == 0.5) %>%
  mutate(month_name = factor(month.name[month],
                             levels = c("October","November","December",
                                        "January","February","March")))

# Step 2: Compute climatology (long-term mean for each month)
climatology <- winter_df %>%
  group_by(month_name, block) %>%
  summarise(clim_mean = mean(temperature), .groups="drop")

# Step 3: Compute avg_temp by group and join climatology
winter_anoms <- winter_df %>%
  group_by(year, block, month_name) %>%
  summarise(avg_temp = mean(temperature), .groups="drop") %>%
  left_join(climatology, by="month_name") %>%
  mutate(anomaly = avg_temp - clim_mean)

# Step 4: Plot anomalies
ggplot(winter_anoms) +
  geom_col(aes(year, anomaly, fill = block.y), position = position_dodge()) +
  facet_wrap(~month_name) +
  labs(y = "Temperature anomaly (°C)")
```

```{r}
obs %>%
  filter(Detect_SpeciesCode == "SEOT") %>%
  pull(Comment) %>%
  write.csv("C:/Users/bhiggins/OneDrive - DOI/Desktop/otters.csv")

otters <- obs %>%
  filter(Detect_SpeciesCode == "SEOT") %>%
  mutate(
    x_lower = str_to_lower(Comment),

    # Define what counts as a "pup" indicator, including your new variants
    has_pup = str_detect(
      x_lower,
      paste0(
        "(",
        "pup",           # 'pup' substring (covers 'pup' and 'pups')
        "|\\bpups\\b",   # explicit plural
        "|\\bp\\b",      # bare 'p' as a whole word (e.g., '1=p', '1 = p')
        "|large\\s*p\\b",# 'large p' (with optional spaces)
        "|large\\s*up\\b",# 'large up' (with optional spaces)
        "|largep",       # 'largep' without space
        "|largeup",      # 'largeup' without space
        ")"
      )
    ),

    any_num = str_detect(x_lower, "\\d+"),

    pups = case_when(
      has_pup & any_num ~ as.numeric(str_extract(x_lower, "\\d+")), # grab leading number
      has_pup & !any_num ~ 1,                                        # 'pup' without number => 1
      TRUE ~ NA_real_
    )
  ) %>%
  select(-x_lower, -has_pup, -any_num) %>%
  mutate(pups = replace_na(pups, 0),
         adults = Count - pups)

#check
which((otters$pups + otters$adults) != otters$Count) #good to go!

```
```{r}
otter_summary <- otters %>%
  group_by(year, site) %>%
  summarise(adults = sum(adults),
            pups = sum(pups),
            ratio = pups/adults) %>%
  mutate(year = as.numeric(year))
```
```{r}
otter_summary %>%
  ggplot() +
  geom_path(aes(year, ratio, color = site)) +
  geom_point(aes(year, ratio, color = site)) +
  labs(title = "SEOT pups/adult ratio")
```

```{r}
bird_densities %>%
  mutate(year = as.numeric(year)) %>%
  filter(species == "SEOT") %>%
  ggplot() +
  geom_path(aes(year, density, color = site)) +
  geom_point(aes(year, density, color = site)) +
  labs(title = "SEOT adult counts")
```
```{r}
temp_df <- temperature %>%
  filter(exposure == "water") %>%
  group_by(year, block) %>%
  summarise(avg_temp = mean(temperature), .groups = "drop")

left_join(otter_summary, temp_df, by = c("year", "site" = "block")) %>%
  filter(site != "KBAY") %>%
  ungroup() %>%
  mutate(
    ratio2 = as.numeric(scale(ratio)),
    temp   = as.numeric(scale(avg_temp))) %>%
  ggplot() +
  geom_path(aes(year, ratio2, color = site)) +
  geom_point(aes(year, ratio2, color = site)) +
  geom_path(aes(year, temp)) +
  facet_wrap(~site, ncol = 1) +
  labs(title = "SEOT pups/adult ratio")
  
```
```{r}
all_taxa %>%
  filter(site != "KBAY") %>%
  select(year, site, mussel_count, BLOY, SEOT, pycnopodia_helianthoides) %>%
  pivot_longer(cols = !c(year, site), names_to = "species", values_to = "Abundance") %>%
  ggplot() +
  geom_path(aes(year, Abundance, color = species)) +
  geom_point(aes(year, Abundance, color = species)) +
  geom_hline(aes( yintercept = 0)) +
  facet_wrap(~site, nrow = 3)
```

```{r}
birds_most_detected <- obs %>%
  filter(!Detect_SpeciesCode %in% c("UNCO")) %>%
  group_by(Detect_SpeciesCode) %>%
  summarise(count = sum(Count),
            detections = n()) %>%
  arrange(desc(detections)) %>%
  head(20) %>% #change this number for more or less species
  pull(Detect_SpeciesCode)

#find the birds we dont want
trash <- obs %>%
  filter(!Detect_SpeciesCode %in% birds_most_detected) %>%
  pull(Detect_SpeciesCode) %>%
  unique()

#add GAK1 temp
#https://research.cfos.uaf.edu/gak1/
gak1 <- read_table(
  "https://research.cfos.uaf.edu/gak1/data/GAK1_Compilation_1970-2025/GAK1_Compilation_1970-2025.dat"
)

gak1$`Date_Time,` <- ymd_hms(gak1$`Date_Time,`)
gak1$year <- year(gak1$`Date_Time,`)

gak_Summary <- gak1 %>%
  mutate(temp = str_replace(`Temperature_[C],`, ",", "")) %>%
  mutate(temp = as.numeric(temp),
         depth = as.numeric(`Pressure_[dbar],`)) %>%
  filter(depth < 10) %>%
  group_by(year) %>%
  summarise(temp = mean(as.numeric(temp), na.rm = T)) %>%
  mutate(temp_scaled = as.numeric(scale(temp)))
  


library(ggcorrplot)
cor_mat <- all_taxa %>% 
   filter(site == "KEFJ") %>%
  select(!site) %>%
  select(-any_of(trash)) %>%
  select(!LEAU) %>%
  select(!UNJA) %>%
  left_join(., gak_Summary, by = "year") %>%
  slice(2:16) %>%
  cor()

ggcorrplot(cor_mat, type = "lower")


hc <- hclust(dist(cor_mat))

order_vec <- hc$order

ggcorrplot(cor_mat[order_vec, order_vec], type = "lower") +
  labs(title ="KEFJ")

```
```{r}
cor_katm <- all_taxa %>% 
   filter(site == "KATM") %>%
  filter(year != 2007) %>%
  select(!site) %>%
  select(-any_of(trash)) %>%
  select(!LEAU) %>%
  select(!UNJA) %>%
  select(!`mediaster_aequalis`) %>%
  left_join(., gak_Summary, by = "year") %>%
  left_join(., otter_summary[otter_summary$site == "KATM",], by = "year") %>%
  select(!site) %>%
  cor(use = "complete.obs")

hc_katm<- hclust(dist(cor_katm))

order_vec_katm <- hc_katm$order

ggcorrplot(cor_katm[order_vec_katm, order_vec_katm], type = "lower") +
  labs(title ="KATM")

```

```{r}
all_taxa %>%
  filter(site != "KBAY") %>%
  select(year, site, RFCO, SEOT) %>%
    left_join(., gak_Summary, by = "year") %>%
  left_join(., otter_summary %>%
              mutate( adults = as.numeric(scale(adults)),
                      pups = as.numeric(scale(pups))), by = c("site", "year")) %>%
  select(!temp) %>%
  pivot_longer(cols = !c(year, site), names_to = "species", values_to = "Abundance") %>%
  filter(species != "adults", species != "pups") %>%
  ggplot() +
  geom_path(aes(year, Abundance, color = species)) +
  geom_point(aes(year, Abundance, color = species)) +
  geom_hline(aes( yintercept = 0)) +
  facet_wrap(~site, nrow = 3)

```
```{r}
all_taxa %>%
  filter(site != "KBAY") %>%
  select(year, site, RFCO, SEOT) %>%
    left_join(., gak_Summary, by = "year") %>%
  left_join(., otter_summary, by = c("site", "year"))
```

```{r}
install.packages('MuMIn')
library(MuMIn)

glm(SEOT ~ PECO + site + year, data = all_taxa) %>%
```
```{r}
all_taxa %>%
   filter(site != "KBAY") %>%
   select(year, site, RFCO, SEOT) %>%
     left_join(., gak_Summary, by = "year") %>%
   left_join(., otter_summary %>%
               mutate( adults = as.numeric(scale(adults)),
                       pups = as.numeric(scale(pups))), by = c("site", "year")) %>%
   select(!temp) %>%
   pivot_longer(cols = !c(year, site), names_to = "species", values_to = "Abundance") %>%
   filter(species != "adults", species != "pups") %>%
   ggplot() +
   geom_path(aes(year, Abundance, color = species)) +
   geom_point(aes(year, Abundance, color = species)) +
   geom_hline(aes( yintercept = 0)) +
   facet_wrap(~site, nrow = 3)
```

```{r}
p.mat.katm <- all_taxa %>% 
   filter(site == "KATM") %>%
  filter(year != 2007) %>%
  select(!site) %>%
  select(-any_of(trash)) %>%
  select(!LEAU) %>%
  select(!UNJA) %>%
  select(!`mediaster_aequalis`) %>%
  left_join(., gak_Summary, by = "year") %>%
  left_join(., otter_summary[otter_summary$site == "KATM",], by = "year") %>%
  select(!c(temp_scaled, site, adults, pups)) %>%
  cor_pmat()

cor_katm <- all_taxa %>% 
   filter(site == "KATM") %>%
  filter(year != 2007) %>%
  select(!site) %>%
  select(-any_of(trash)) %>%
  select(!LEAU) %>%
  select(!UNJA) %>%
  select(!`mediaster_aequalis`) %>%
  left_join(., gak_Summary, by = "year") %>%
  left_join(., otter_summary[otter_summary$site == "KATM",], by = "year") %>%
  select(!c(temp_scaled, site, adults, pups)) %>%
  cor(use = "complete.obs")

hc_katm<- hclust(dist(cor_katm))

order_vec_katm <- hc_katm$order

ggcorrplot(cor_katm, p.mat = p.mat.katm, hc.order = F, insig = "stars") +
  labs(title ="KATM")
  ggsave("C:/Users/bhiggins/OneDrive - DOI/Desktop/KATM_data_dredging.png", height = 10, width = 10)

```

```{r}
p.mat.kefj <- all_taxa %>% 
   filter(site == "KEFJ") %>%
  filter(year != 2007) %>%
  select(!site) %>%
  select(-any_of(trash)) %>%
  select(!LEAU) %>%
  select(!UNJA) %>%
  select(!`mediaster_aequalis`) %>%
  left_join(., gak_Summary, by = "year") %>%
  left_join(., otter_summary[otter_summary$site == "KEFJ",], by = "year") %>%
  mutate(seot_pup_ratio = ratio) %>%
  select(!c(temp_scaled, site, adults, pups, ratio)) %>%
  cor_pmat()

cor_mat <- all_taxa %>% 
   filter(site == "KEFJ") %>%
  filter(year != 2007) %>%
  select(!site) %>%
  select(-any_of(trash)) %>%
  select(!LEAU) %>%
  select(!UNJA) %>%
  select(!`mediaster_aequalis`) %>%
  left_join(., gak_Summary, by = "year") %>%
  left_join(., otter_summary[otter_summary$site == "KEFJ",], by = "year") %>%
    mutate(seot_pup_ratio = ratio) %>%
  select(!c(temp_scaled, site, adults, pups, ratio)) %>%
  cor(use = "complete.obs")

#hc <- hclust(dist(cor_mat))

#order_vec <- hc$order

ggcorrplot(cor_mat, p.mat = p.mat.kefj,  insig = "stars", hc.order = F) +
  labs(title ="KEFJ")
 ggsave("C:/Users/bhiggins/OneDrive - DOI/Desktop/KEFJ_data_dredging.png", height = 10, width = 10)
```
keep making it more of a menace. add the snails!
```{r}
snails_upper <- snails %>%
  mutate(site = substr(SiteID, 1, 3),
         site = case_when(site == "AKP" ~ "KATM",
                          site == "PWS" ~ "WPWS",
                          site == "KEP" ~ "KEFJ")) %>%
  group_by(Year, Elevation_Position, site, SpeciesName) %>%
  summarise(count = mean(Count)) %>%
  pivot_wider(names_from = SpeciesName, values_from = count) %>%
  ungroup() %>%
    filter(Elevation_Position == "Upper (1.5 m MLLW)") %>%
  select(-Elevation_Position)

all_taxa <- left_join(all_taxa, snails_upper, by = c("year" = "Year", "site"))
```


thinking out loud.
all the heatwave analyses use the heatwave as a factor (pre, during, post), but looking at the temperatures, there sort of isnt a post, or at least its less straight forward. try using temp as a continious variable?


```{r}
gak_Summary %>%
  filter(year > 2006) %>%
  mutate(lta = mean(temp_scaled)) %>%
  ggplot() +
  geom_col(aes(year, temp_scaled))
```

